### 2020-07-24 edwin_original.tex
First commit

%\documentclass[reprint,amsmath,amssymb,superscriptaddress,prl,showpacs,onecolumn]{revtex4-1}
\documentclass[reprint,amsmath,amssymb,superscriptaddress,showpacs,pre]{revtex4-1}
\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
\usepackage{color}
\usepackage{epstopdf} 
\usepackage{subfig}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{amsthm}
\usepackage{hyperref}
\usepackage{algorithm}
\usepackage{algorithmic}
%\usepackage{physics}
\usepackage{lineno}
\linenumbers

%\usepackage{pstricks}
\usepackage{braket}
\DeclareMathOperator\atanh{arctanh}

%\usepackage[hypertex]{hyperref}
\usepackage[sans]{dsfont}

\usepackage{epstopdf}
\usepackage{ulem}
\usepackage{hyperref}
\usepackage{mathbbol}
\usepackage{bbm}

\let\originalleft\left
\let\originalright\right
\renewcommand{\left}{\mathopen{}\mathclose\bgroup\originalleft}
\renewcommand{\right}{\aftergroup\egroup\originalright}
%\newcommand{\bra}[1]{\ensuremath{\left< #1\right|}}
%\newcommand{\ket}[1]{\ensuremath{\left|#1\right>}}
\newcommand{\ovl}[2]{\ensuremath{\left\langle #1\middle|#2\right\rangle}}
\newcommand{\ie}{\emph{i.e.}}
\newcommand{\eg}{\emph{e.g.}}
\newcommand{\cf}{cf.}
\newcommand{\cp}{\emph{cp.}}
%\newcommand{\braket}[2]{\left< #1,#2\right>}
%%\newcommand{\<}{\langle}
%%\renewcommand{\>}{\rangle}
\providecommand{\abs}[1]{\left|#1\right|}
\providecommand{\norm}[1]{||#1||}
\newcommand{\arctanh}[1]{\mathrm{arctanh}#1}
\newcommand{\Sign}[1]{\mathrm{Sign}#1}

\def\<{\langle}
\def\>{\rangle}
\def\vx{\vec x}
\def\vh{\vec h}

\newcommand{\ud}{\mathrm{d}}
\newcommand{\etal}{\textit{ et. al. }}
\newcommand{\Ham}{\mathcal{H}}
\newcommand{\B}{\beta}
\def\matF{{\mathbb F}}
\def\Jij{J_{ij}}

\def\Imat{\mathbb{I}}
\def\ext{\qopname\relax m{ext}}

\newcommand{\mi}{\mathrm{i}}
\DeclareMathOperator{\Tr}{Tr}

%%%%

\newcommand{\ddroit}{\textrm{d}}
% \newcommand{\eexp}{\text{e}}
%\newcommand{\ie}{\emph{i.e.}$\;$}
%\newcommand{\eg}{\emph{e.g.}$\;$}
\newcommand{\Ptot}{P(A_1\ldots A_L)}
\newcommand{\xz}{\vec{x}_0}
\newcommand{\xo}{\vec{x}_1}
\newcommand{\xt}{\vec{x}_2}
\newcommand{\Lam}{\bm{\Lambda}}
\newcommand{\Sig}{\bm{\Sigma}}
\newcommand{\Sigmone}{\bm{\Sigma^{-1}}}
\def\fonetwo{\frac 1 2}
\def\Cmone{C^{-1}}
\def\Lambtwo{\Lambda^2} 
\def\tildex{\tilde{x}}
\newcommand{\HRule}{\rule{\linewidth}{0.5mm}}
\newcommand{\Xk}[1]{X^{\{#1\}}}
\newcommand{\xk}[1]{\vec{x}_{#1}}

\begin{document}
\title{Inference of global coevolutionary models from hierarchical correlated data. }

\author{Edwin Rodriguez Horta} 
%\email{erodriguezh1990@gmail.com}
\affiliation{Group of Complex Systems and Statistical Physics, Department of Theoretical Physics, University of Havana, Cuba}
\affiliation{Laboratoire de Biologie Computationnelle et Quantitative (LCQB), Institut de Biologie Paris-Seine}
\author{Pierre Barrat-Charlaix} 
%\email{ p.barrat@live.fr }
\affiliation{Biozentrum, University of Basel, 4056 Basel, Switzerland}
\author{Martin Weigt} 
\affiliation{Laboratoire de Biologie Computationnelle et Quantitative (LCQB), Institut de Biologie Paris-Seine}
%\email{martin.weigt@upmc.fr}
\author{Alejandro Lage} 
%\email{ale.lage@gmail.com}
\affiliation{Group of Complex Systems and Statistical Physics, Department of Theoretical Physics, University of Havana, Cuba}
\date{\today}
%\pacs{05.60.Gg, 03.65.Xp, 72.10.-d, 87.15.hj}

%Quantum transport (general physics)
%Tunneling (general physics)
%Theory of electronic transport (cond mat)
%Transport dynamics (biomolecules)

\begin{abstract}
The Inverse problem of Statistical Physics infers maximum-entropy  models  compatibles with a corresponding set of empirical averages extracted from a high-dimensional dataset of equilibrium configurations. Practical interest extended these methods to non-equilibrium data  like  time series, where detailed balance does not hold. However, in several applications, data samples  result from evolutionary processes and the relation between configurations is conditioned by a hierarchical structure of the population. This makes sample statistics a superposition of  signals coming from both: internal configuration correlations and relatedness correlations produced by common evolutionary  history, typically represented by a tree. How to disentangle both sources of correlation for a better parameter inference is an open question to explore. Here we propose an approach  where the  evolutive process through a phylogenetic tree is described by  a  multivariate Ornstein-Uhlenbeck dynamics characterized by  gaussian propagator and stationary distributions. Inference of these distributions from  phylogenetic non-independent samples is solved in a Bayesian framework. This procedure can be  extended to discrete variables by using a binary representation of data and approximating  binary sates by continuous variables. Our method proposes a possible way for  a better estimation of both, equilibrium and dynamic parameters, making  applications more accurate.

\end{abstract}

\maketitle

\section{\label{sec:int} Introduction}
%\label{sec:int}
Theory and applications of the inverse problem of statistical physics has had great development in recent years since the availability of large amounts of data on a microscopic scale produced by the continuous technological progress. In this context statistical physics paths: from model parameters to  observables, is reversed, toward find global models which describe the statistics of  a set of observed  configurations \cite{Inverse_problem_Berg}.

Most of the work has been focus in equilibrium reconstruction , where the aim is to learn max Entropy models, Boltzman distribution, from independent and identically distributed equilibrium configurations. Another kind of methods deal with parameters learning from either time series data or from samples of the non-equilibrium steady state: non-equilibrium reconstruction.

Less effort has been done for data biased by population structured signal, where  samples  are not independent because of presence of relatedness between configurations. These structure could be for instance of  geographical or demogrhapic  nature in a population or  phylogenetic  nature  related to a evolutionary processes shaping trait variation among species.  Here we are going to deal with data given by sequences phylogenetically related  through a tree, which describes the hierarchical correlation between configurations. An important problem is the  distinction   of patterns induced by species traits or by phylogeny.

Our source of inspiration came from the inverse problem in proteins \cite{Inverse_problem_proteins}, where  covariance signal present in an alignment of homologous proteins is described via Potts model. This covariance signal is assumed a consequence of phenotypics constraints like protein structure and function. However it is known that phylogeny relationships in homologous sequences  contaminate the statistics of  the alignment used to learn the equilibrium model. Therefore a direct inference on this alignments  leads to the existence of Potts parameters that attempt to model the full biased statistics limiting their capacity to predict phenotypic properties.

 To define and solve the inverse problem of statistical physics for data samples with hierarchical  relatedness structure is the aim of this work.

Main difficulties to address  this kind of problem  come from the necessity to use  a proper global evolutionary model which describes the relation between sequences edge connected in a tree and the discrete  characters of data variables which makes certain calculations intractable \cite{entropy_paper}.

Here we  adopt an approximation where we forget about the discrete nature of variables and model them by continuous variables instead. This transforms the Potts model into a Gaussian distribution \cite{Baldassi}, making the design of a global propagator tractable. We propose a model of evolution described by a multivariate Ornstein-Uhlenbeck process \cite{gardiner}, which is used  to generate evolutionary related data given an arbitrary phylogenetic tree. The corresponding inverse problem became in the inference of the Gaussian potential (Potts parameters) from the knowledge of tree leaves configurations and the dynamics of evolution.

To this aim, we first review in section \ref{sec:ornstein_uhlenbeck_dynamics}, main characteristics of a multivariate Ornstein-Uhlenbeck process. In section \ref{evolutionary process} we describe the evolutionary process, describing the algorithm used to generate data phylogenetically biased. The statement of the problem in a Bayesian framework is shown in section \ref{statement of the problem}. Section \ref{methods} is dedicated to describe methodology used to solve the inverse problem. Results for evolution of continuos variable sequences and  the application for configuration of discrete variables is presented in section \ref{Results}. Finally the conclusion and outlook in \ref{conc}.

\section{\label{sec:ornstein_uhlenbeck_dynamics}Short summary about Multivariate Ornstein-Uhlenbeck process}
%\label{sec:ornstein_uhlenbeck_dynamics}
Let consider a system characterized by a vector $\mathbf{x}$ with an equilibrium probability distribution given by
\begin{equation}
P(\mathbf{x}) =\frac 1 Z \exp \left\lbrace -\Ham(\mathbf{x}) \right\rbrace \;\; \mbox{with } \Ham(\mathbf{x}) =\frac{1}{2} \left\lbrace \mathbf{x}^T \bm J \mathbf{x}+ \mathbf{h} \cdot \mathbf{x} \right\rbrace \label{eq:equil}
\end{equation}
where the degrees of freedom $x_i \in \mathbb R, i\in [1\ldots N]$ are continuous variables, and the interactions among them $\Jij$ are fixed, but randomly selected 
from a given ensemble of symmetric positive definite matrices. The vector $\mathbf{h}$ describes a local field which shift a  equilibrium configuration of a zero centered   harmonic potential ($ \mathbf{x}^T \bm J \mathbf{x}$) , for simplicity we going to assume $\mathbf{h}=\mathbf{0}$ .

The Langevin equation for  non-massive $N$-dimensional particles with position $\mathbf{x} = (x_i),\;\;i\in\{1\ldots N\}$   interacting according to the Hamiltonian \ref{eq:equil} take the form:

\begin{equation}
\gamma \frac{\ddroit \mathbf{x} }{\ddroit t} = - \bm J \mathbf{x}  + \mathbf{\xi}(t)
\end{equation}
which represent  a multivariate Ornstein-Uhlenbeck   process, being $ \mathbf{\xi}(t)$ an stochastic term and $\gamma$ the viscosity.

Modeling the stochastic term $\xi_i(t)$ as a zero mean Gaussian white noise with variance $\langle \xi_i(t)\xi_j(t') \rangle = 2 K_B T\gamma\delta(t-t')$ that satisfies the fluctuation-dissipation relation, we obtain  the Ito stochastic differential equation for a multivariate O-U process \cite{singh2017multiOU}:
\begin{equation}
\ddroit x_i = -J_{ij}'x_j dt + \left( \sqrt{2 D}\right) _{ij} dW_j
\end{equation}

where:
\begin{itemize}
	\item $dW_j=\xi_j(t) dt$ represent a stochastic Wiener process. 
	\item $J_{ij}'=\frac{J_{ij}}{\gamma}$ is a stiffness stable matrix of mean regression rates.
	\item $D_{ij}=\frac{KT}{\gamma}\delta_{ij}$ is a symmetric positive definite matrix of diffusions coefficients.  
\end{itemize}

It can be shown that the corresponding Fokker-Planck equation is

\begin{equation}
\partial_t P_{1|1}=\hat{L} P_{1|1} 
\end{equation}
where  $P_{1|1}\equiv P( \mathbf{x}', t'|\mathbf{x}, t)$ is the probability density of displacement from $ \mathbf{x}$ at time $t$ to $ \mathbf{x} '$ at time $t'$, and $\hat{L} $ is the  Fokker-Planck operator given by:

\begin{equation}
\hat{L}( \mathbf{x})=-\frac{\partial}{\partial x_i}J_{ij}' x_j+\frac{\partial^2}{\partial x_i\partial x_j}D_{ij}
\end{equation}
The stationary solution for the Fokker-Planck equation $\hat{L} P_{eq}( \mathbf{x})=0$ is:
\begin{equation}
P_{eq}( \mathbf{x}) = \frac{1}{\sqrt{(2\pi)^N \vert \bm{C}\vert}}\exp\left\{ -\frac{1}{2} \mathbf{x}^T\bm{C}^{-1} \mathbf{x} \right\}
\end{equation}
which is a zero mean Gaussian distribution with $\bm C=\left\langle  \mathbf{x}  \mathbf{x}^T\right\rangle $ as the covariance matrix. If from the beginning we set $\mathbf{h}\neq\mathbf{0}$ we get a Gaussian distribution with mean value shift from zero and covariance matrix $\bm C=\left\langle \mathbf{x} \mathbf{x}^T\right\rangle- \left\langle \mathbf{x} \right\rangle \left\langle \mathbf{x} \right\rangle$.

The solution for the  Fokker-Planck equation is 
\begin{eqnarray}
\nonumber
\lefteqn{P(\mathbf{x} ' | \mathbf{x} , \Delta t) =\frac{1}{\sqrt{(2\pi)^N(1-e^{-2\Delta t})\vert\bm{\Sigma}^{-1}\vert}}}\\%\nonumber
&&\times \exp\left\{ -\frac{1}{2}(\mathbf{x} '-\bm{\mu})^T\bm{\Sigma}^{-1}(\mathbf{x} '- \bm{\mu}) \right\}
\label{eqn:multiOU}
\end{eqnarray}

where 
$$ \bm{\mu} = \Lam \mathbf{x}, \qquad \bm{\Sigma} = \bm{C} - \Lam\bm{C}\Lam, \qquad \Lam = e^{-\bm{\frac{J}{\gamma}}\Delta t}, \Delta t=t-t'$$ 
Matrices $\bm J'$, $\bm C$ and $\bm D$ are not independent but are related by the Liapunov stationary condition :
$$\bm J' \bm C+(\bm C \bm J')^T=2 \bm D$$
Using the Liapunov condition and setting $kT=1$ we get

\begin{equation}
\bm J=\bm C^{-1}
\end{equation}

Particles evolves under potential  $V(x)=\frac{1}{2}\mathbf{x}^T\bm{C}^{-1}\mathbf{x} = \frac{1}{2}\sum_{i,j} J_{ij}x_i x_j$ and the evolutionary process leads to stationary distribution similar to \ref{eq:equil} .

Another important feature  of the Ornstein-Uhlenbeck process based in it's Gauss-Markov property ensures that the times correlation function obeys the linear regression theorem

\begin{equation}
\bm G(t-t')\equiv \left\langle \mathbf{x}(t)\mathbf{x}(t')\right\rangle= e^{-\frac{\bm{ J}}{\gamma}\Delta t} \bm C=\Lam \bm{C}
\end{equation}

describing the covariance of configurations $\mathbf{x}$ and $\mathbf{x}'$ separated in time by $\Delta t$.

\section{Evolutionary process}
\label{evolutionary process}
To simulate an evolutive process for a  multivariate Ornstein-Uhlenbeck dynamics, we must use the propagator \ref{eqn:multiOU} wich provide the probability of observing sequence $\mathbf{x}'$ knowing that it has sequence $\mathbf{x}$ as an ancestor at a time $\Delta t$ in the past.

An equivalent  method   to generate  a discrete sampling of a path made by  a sequence of states at times $t_n= n \Delta t$ can be obtained from the iteration:
\begin{equation}
\mathbf{x}_{n+1}=\Lam \mathbf{x}_n+\sqrt{\Sigma}\xi_n
\label{eq:evol}
\end{equation}
where $\xi_n\sim N(0,1)$ is an $N-dimensional$ uncorrelated normal variate with zero mean and unit variance.

In this context  $\gamma$ parameter  is the inverse of mutation rate, $\Delta t/\gamma$ is the number of proposed mutations in the interval $\Delta t$ and  potential $\bm J$ determine  how probable is the mutation (if we assume variation in the potential as fitness score).

\subsection{Phylogenetic correlated samples}
Given an  ancestor  configuration $\mathbf{x}_0$,  extracted randomly from the equilibrium distribution (\ref{eq:equil}),  phylogenetic correlated sequences can be generetaded   following the algorithm \ref{alg_1} ,  for simplicity we adopted a binary and homogeneous tree where all branches have the same length.

\begin{algorithm} [H]
			\caption{Generates tree of evolved vectors $T = \{\vx_0, \vx_{1,1}, \vx_{1,2}, \ldots\}$}
				\label{alg_1}
	\begin{algorithmic}[1]
		\REQUIRE Root $\mathbf{x}_0$, branching events $K$, distance to mutants $\Delta t$.
		\ENSURE Returns $T$
		\STATE $T = \{ \mathbf{x}_0 \}$  
		\STATE $L = \{\mathbf{x}_0 \} $  \COMMENT{Leaves of the tree (the last added nodes)}
		\FOR{$(k=1; k<K+1; k+=1;)$}
		\STATE $L_{new} = \{\}$ 
		\FOR{($\mathbf{x}$ in $L$)} 
		\STATE $\mathbf{x}_{\mbox{child1}} = \mbox{Ornstein-Uhlenbeck-evolve}(\mathbf{x}, \Delta t)$
		\STATE $\mathbf{x}_{\mbox{child2}} = \mbox{Ornstein-Uhlenbeck-evolve}(\mathbf{x}, \Delta t)$
		\STATE Append $T \leftarrow= \{\mathbf{x}_{\mbox{child1}}, \mathbf{x}_{\mbox{child2}}\}$
		\STATE Append $L_{new} \leftarrow= \{\mathbf{x}_{\mbox{child1}}, \mathbf{x}_{\mbox{child2}}\}$
		\ENDFOR
		\STATE $L = L_{new}$
		\ENDFOR
		\RETURN $T = \{\mathbf{x}_0, \mathbf{x}_{1,1}, \mathbf{x}_{1,2}, \ldots\}$
	\end{algorithmic} 
\end{algorithm}

The step named Ornstein-Uhlenbeck-evolve$(\mathbf{x},\Delta t)$, carries out a Ornstein-Uhlenbeck evolution (\ref{eq:evol}) starting at configuration $\mathbf{x}$  for a time $\Delta t$.

The phylogenetic bias present at leaves of the tree depend of the time interval $\Delta t$, the mutation rate $1/\gamma$ and the potential $\bm J$.  Two configurations linked by  OU dynamics (\ref{eq:evol})   are independent  if $\bm J\Delta t /\gamma\rightarrow \infty$ being $\Lambda\rightarrow 0$. Let define the  characteristic time $\Delta t_c$ as the amount of time necessary for an exponential decay of the  quantity   $\Lambda$ by the factor $\frac{1}{e}$, to this  $\bm J\Delta t_c/\gamma \rightarrow 1$.  As $\bm J$ is a matrix  the previous condition became  in $\lambda_{max}\Delta t_c /\gamma\rightarrow1$, where  $\lambda_{max}$  is the  maximum eigenvalue of $\bm J$, describing it's higher grow direction.  We going to  arbitrary define that two configuration remain correlated  if the sampling interval between them satisfy $\Delta t \ll\Delta t_c$.

Similarly we assume  that contemporary sequences in a tree are   under a strong phylogenetic correlated regime, when the  path time between most distant sequences is lower than characteristic time $\Delta t_c$ . For a binary homogeneous tree with  $K$ branching events this condition became $\Delta t \ll \frac{\Delta t_c}{2*K}$.

 Alternatively if we consider $\bm J$ and $\Delta t$  fixed the same criteria allows us to define the  parameter $\gamma_c\rightarrow \lambda_{max}\Delta t $  as the threshold value for $\gamma$  above which two sequences are considered correlated. For a binary homoneous tree $\gamma_c\rightarrow \lambda_{max}2*K\Delta t $ determine the  characteristic value for the inverse of mutation rate. Tunning the mutation rate respect $1/\gamma_c$ we going to generate sequences with distinct levels of phylogenetic bias.

\begin{figure}[!htb]
\includegraphics[keepaspectratio=true,width=0.35\textwidth]{tree.pdf}
\caption{ Phylogenetic model for system evolution  with $K=3$ branching events \label{fig:tree}}
\end{figure}

\section{Statement of the problem}
\label{statement of the problem}
The system studied is the following: a balanced phylogenetic tree is given, with $K$ branching events and times $\Delta t_{ij}$ being assigned to each of its branches. A root configuration $\mathbf{x}$ is chosen with gaussian probability 
$$ P_{eq}(\mathbf{x}) = \frac{1}{\sqrt{(2\pi)^N \vert \bm{J}^{-1}\vert}}\exp\left\{ -\frac{1}{2}\mathbf{x}^T\bm{J}\mathbf{x} \right\}. $$ 
It then evolves using Ornstein-Uhlenbeck dynamics which is fully determined by the pairs $\left\lbrace \bm J,\gamma \right\rbrace $. At each division event of the tree, two or more  copies of the system are created and evolve independently along each branch. The final result of the process are the configurations at leaves of the tree. \\
The inference problem is then: with the knowledge of the leaves configurations $\mathbf{y}=\left[ \mathbf{x}_1,\mathbf{x}_2,...,\mathbf{x}_k\right] $ , the tree and  the type of evolution dynamics (O-U evolution), is it possible to reconstruct the potential $\bm J=\bm{C}^{-1}$ acting on the system and the dynamics parameter $\gamma$?

The Bayesian approach  to this is:
\begin{itemize}
	\item Compute the probability of observing leaves configurations given the potential $\bm{J}$,and mutation rate $1/\gamma$ given by the likelihood: $P(\mathbf{y}\vert\bm{J},\gamma)$.
	\item Compute posterior of the parameters  using Bayes formula: $$P(\bm{J},\gamma\vert \mathbf{y}) = \frac{P(\mathbf{y}\vert\bm{J},\gamma)P(\bm{J},\gamma)}{Z}$$
	\item Maximize  posterior of the model by  maximum likelihood approach: $ \max_{\bm J,\gamma} P(\mathbf{y}\vert\bm{J},\gamma)$  which is exact  when this function is strictly concave and prior distribution is uniform.
\end{itemize}

\section{Methods}
\label{methods}
Since the process is always Gaussian the distribution of the leaves has to be Gaussian itself. Furthermore, we know the covariance between any two elements. Within one leaf, the covariance is the equilibrium covariance $\bm {J}^{-1}$, and among leaves it has to be $\Lambda^{\Delta t_{ij}}{\bm J^{-1}}  $
%\[ {\bm g}_{ij}(\Delta t_{ij},\bm{J}) = \bm \Lambda^{\Delta t_{ij}}{\bm J^{-1}}   \]
with $\bm \Lambda = \exp(-\bm J/\gamma)$ and $\Delta t_{ij}$ is the  path time   between nodes $i$ and $j$ along the branches of the tree.

The distribution of the leaves is given by 
\begin{eqnarray}
\nonumber
\lefteqn{P(\mathbf{y}=\left[ \mathbf{x}_1,\ldots,\mathbf{x}_R\right] ) = \frac 1 {\sqrt{ 2 \pi^{N R} \det \mathbb{G}}  }}\\%\nonumber
&&\times \exp\left(-\frac 1 2 [\mathbf{x}_1,\ldots,\mathbf{x}_R] \mathbb{G}^{-1}[\mathbf{x}_1,\ldots,\mathbf{x}_R]^t \right)
\label{eqn:leaves_dist}
\end{eqnarray}
where $\mathbb{G}$ is a block matrix whose structure is induced by the  phylogenetic tree  and   its elements are given by:

\begin{equation}
\bm G_{i,j} =
\begin{cases}
\bm J^{-1} & \text{$i=j$}.\\
\bm \Lambda^{\Delta t_{ij}}{\bm J^{-1}} & \text{otherwise}  
\end{cases}
\end{equation}

The log likelihood is
\begin{equation}\label{eq:log_likelihood}
L(\mathbf{y}|\bm J,\gamma) = \frac 1 2 \log \det  \mathbb{G}^{-1}  -\frac 1 2 \mathbf{y}\mathbb{G}^{-1} \mathbf{y}^t
\end{equation}

Trivial maximization over a full fledge $\mathbb{G}$ would give back the experimental correlation between leaves, but it would lack the structure of $\mathbb{G}$. Maximization over $J_{i,j}$ could be attempted through the chain rule
\[  \frac{\partial L}{\partial J_{i,j}} = \sum_{p,q}\frac{\partial L}{\partial G^{-1}_{p,q}} \frac{\partial G^{-1}_{p,q}}{\partial J_{ij}} \]
Although $\frac{\partial L}{\partial G^{-1}_{p,q}}$ is simple to compute, the part $ \frac{\partial G^{-1}_{p,q}}{\partial J_{ij}}$ is not.

To numerically optimize the log likelihood is necessary   to compute  the determinant of  $\mathbb{G}$   for each change in the potential matrix $\bm{J}$.  It is a huge problem for systems of realistic sizes with complexity $O(N^3*R^3)$ where $R$ is the number of leafs in the tree and $N$ the size of configuration vector  $\mathbf{x}_i$. However we could simplify it  using the symmetric properties of  this matrix, as we will show below.
\subsection{Simpler case: Binary and homogeneous tree.}
Let assume that the tree is binary and completely homogeneous, all branches have the same length $\Delta t$. For instance the sequences covariance matrix for a binary tree with $K=2$ branching events and four leaves is
\begin{equation} 
\mathbb{G} =  \left(
\begin{array}{cccc}
{ \bm J^{-1}} & {\bm J^{-1}} \Lambda ^{2\Delta t} & {\bm J^{-1}} \Lambda ^{4\Delta t} & {\bm J^{-1}} \Lambda ^{4\Delta t} \\
{\bm J^{-1}} \Lambda ^{2\Delta t} & {\bm J^{-1}} & {\bm J^{-1}} \Lambda ^{4\Delta t} & {\bm J^{-1}} \Lambda ^{4\Delta t} \\
{\bm J^{-1}} \Lambda ^{4\Delta t} & {\bm J^{-1}} \Lambda ^{4\Delta t} & {\bm J^{-1}} & {\bm J^{-1}} \Lambda ^{2\Delta t} \\
{\bm J^{-1}} \Lambda ^{4\Delta t} & {\bm J^{-1}} \Lambda ^{4\Delta t} & {\bm J^{-1}} \Lambda ^{2\Delta t} & {\bm J^{-1}} \\
\end{array}\label{eq:CL}
\right)
\end{equation}

For hypergeometric matrices as \ref{eq:CL} there is $K+1$ different eigenvalues  given by:

\begin{equation}
\label{eq:lambda_hyper}
\lambda_k =
\begin{cases}
J^{-1}(1+\sum_{l=1}^{k-1} 2^{l-1}\Lambda^{2l\Delta t}-2^{k-1}\Lambda^{2k\Delta t} )\;\;\; k\in[1,K] \\
J^{-1}(1+\sum_{l=1}^{K} 2^{l-1}\Lambda^{2l\Delta t}) \;\;\;\;\;\;\;k=K+1
\end{cases}
\end{equation}

where $\lambda_{K+1}\ge\lambda_{K}\cdots\ge\lambda_{1}$. For $k<K+1$ degenerancy of eigenvalues $\lambda_k$ is given by : $d_k=2^{K-k}$.

The associated eigenvectors are independent of potential $\bm J$ and reflects the events in the phylogenetic tree  where each eigenvector  $k$ captures the duplication events in the $(K+1 -k) 	st$ generation. Summarizing :

\begin{equation} 
\nonumber
	\bm u_k =
	\begin{cases}
	{(\underbrace{\overbrace{1,\ldots,1}^{2^{k-1}},\overbrace{-1,\ldots,-1}^{2^{k-1}}}_Q,0,\ldots,0)\ \bigcup \Gamma(u_k) } \;\;\; k\in[1,K] \\
	(1,1,1,\ldots,1,1,1)\;\;\;\;\;\;\;k=K+1
	\label{eq:eigenvec_simpler_case}
\end{cases}	
\end{equation}

where $\Gamma(u_k)$ represent $d_k$ combinations resulting of shift the block of length $Q$ without overlapping, generating all eigenvectors with the same corresponding eigenvalue $\lambda_k$.  The eigenvectors are orthogonal to eachother, and can be  normalized and arranged horizontally into a matrix $U$. With this in mind  the log-likelihood can be written as

\begin{eqnarray}
\nonumber
\lefteqn{L=\frac 1 2 \log\det \mathbb{G} ^{-1}-\frac 1 2 y \mathbb{G} ^{-1}y^t}\\
&&=\frac 1 2 \log\prod_{k}\lambda_k^{-1}  -\frac 1 2 \mathbf{y}  \bm U' \mbox{Diag}\left({\frac 1 \lambda_k}\right) \bm U \mathbf{y}^t \\\nonumber
&&=-\frac 1 2\sum_{k}\log\lambda_k-\sum_k \frac{\langle \mathbf{u}_k | \mathbf{y}\rangle^2}{\lambda_k} 
\end{eqnarray}

\subsection{Likelihood maximization for leaves as  one-dimensional  vectors: N=1}

For one dimension ($N=1$) the log-likelihood can be maximized over the only two parameter  $p_{i=1,2}=\gamma , J = \sigma^{-2}$, being $\sigma$ standard deviation , following the 
gradient
\begin{equation}
\nabla L=\sum_i\frac{\partial L}{\partial p_i}\hat{p_i} 
\end{equation}
with 
\begin{eqnarray}
\nonumber
\label{eq:gradient1D}
\lefteqn{\frac{\partial L}{\partial p_i} = -\sum_{k}^R \frac 1 {\lambda_k} \frac{\partial \lambda_k}{\partial p_i}  +\sum_{k}^R \frac{\langle \mathbf{u}_k | \mathbf{y}\rangle^2}{\lambda_k^2} \frac{\partial \lambda_k}{\partial p_i}} \\ &&= \sum_{k}^R  \left(\frac{\langle \mathbf{u}_k | \mathbf{y}\rangle^2}{\lambda_k} -1 \right) \frac 1 {\lambda_k} \frac{\partial \lambda_k}{\partial p_i}	 
\end{eqnarray}

Using  equation \ref{eq:gradient1D} we implement a gradient ascent method to maximize the log-likelihood of data generated by algorithm \ref{alg_1} with $\gamma_0=1.0$, $\Delta t=0.5 t_c$ and $\sigma_{eq} \sim N(0,1)$ . In the figure below  we show the convexity of the log-likelihood function and the ability of our algorithm to find the value of $ \sigma^2$ which maximizes it ($\sigma^2_{max}$),  being a better approximation for the true  standard deviation ($\sigma^2_{eq}$) than the empirical one ($\sigma^2_{emp}$) .

\begin{figure}[!htb]
	\includegraphics[keepaspectratio=true,width=0.40\textwidth]{log-likelihood_vs_sigma_1D.png}
\caption{ Log-likelihood function for data sampled trough a tree with $R=2^{13} $ and $dt=0.5tc$.}
\end{figure}

\subsection{Generalized inference method: N-dimensional case}
For the most general case the coupling matrix $\bm J$ is a $N\times N$  matrix and as consequence $\mathbb{G}$ a $RN\times RN$ matrix where $N$ is the sequence length and $R$ the leafs number. 
The eigensystem of $\bm J$ allows to write:

$$\bm  J \ket{\textbf{s}_i} =\rho_i\ket{\textbf{s}_i }$$
and
$$\bm J=\sum_{i}\rho_i \ket{\textbf{s}_i}  \bra{\textbf{s}_i}  $$

been $\ket{\textbf{s}_i}$ and $\rho_i$ eigenvectors and eigenvalues of $\bm J$. As each entry $m,n$ of $\mathbb{G}$ is a function of $\bm J$ we have:

$$\mathbb{G}=\bm G_{m n}(\bm J)$$

with $m,n \in (1:R)$ and projecting on $\ket{\textbf{s}_i}$   $$\bm G_{m n}(\bm J)\ket{\textbf{s}_i} =\bm G_{m n}(\rho_i)\ket{\textbf{s}_i} $$  note now that to replace $\bm J$ matriz by it's eigenvalues $\rho_i$   produce $\bm G_{m n}(\rho_i)$ , a hypergeometric matrix  with $\dim(\bm G_{m n}(\rho_i))=R\times R $ and eigensystem ($\lambda_k(\rho_i)$,$\textbf{u}^k(\rho_i)$).

For a homogeneous and binary phylogenetic tree such as the one presented in figure \ref{fig:tree}, $\lambda_k (\rho_i)$, $\textbf{u}^k(\rho_i) $ are given by the expressions provided by \ref{eq:lambda_hyper} and \ref{eq:eigenvec_simpler_case}, however in general cases $ \lambda_k (\rho_i) $, $\textbf{u}^k(\rho_i)$ must be calculated for every $ i \in (1: N) $

Lets consider the vector built as the Kronecker product  $$\ket{\textbf{u}^{ki}} \otimes\ket{\textbf{s}_i} =(u^{ki}_1 \ket{\textbf{s}_i} \dotsc u^{ki}_R \ket{\textbf{s}_i} )^T=u^{ki}_{\alpha}s^{i}_{\beta}$$
this vectors and the eigenvalues $\lambda_{k}(\rho_i)=\lambda_{k,i}$ conform the eigensystem of the full matrix  $\mathbb{G}$ as is shown below:

\begin{eqnarray*}\mathbb{G}\ket{\textbf{u}^{ki}} \otimes\ket{\textbf{s}_i} &=&\sum_{n}\bm G_{mn}(\bm J)(\ket{\textbf{u}^{ki}} \otimes\ket{\textbf{s}_i})_n\\ &=&\sum_{n}\bm G_{mn}(\bm J)\ket{\textbf{u}^{ki}}_n\otimes\ket{\textbf{s}_i}\\&=&\sum_{n}\bm G_{mn}(\bm J)\ket{\textbf{s}_i}\otimes \ket{\textbf{u}^{ki}}_n\\&=&\sum_{n}\bm G_{mn}(\rho_i)\ket{ \textbf{s}_i} \otimes \ket{\textbf{u}^{ki}}_n\\&=&\sum_{n}\bm G_{mn}(\rho_i)\ket{\textbf{u}^{ki}}_n\otimes\ket{\textbf{s}_i}\\&=&\lambda_k(\rho_i) \ket{\textbf{u}^{ki}}\otimes\ket{\textbf{s}_i}=\lambda_{k,i} u^{ki}_{\alpha}s^{i}_{\beta} 
\end{eqnarray*}

Therefore $$\mathbb{G}\ket{\textbf{u}^{ki}} \otimes\ket{\textbf{s}_i}=\lambda_k(\rho_i) \ket{\textbf{u}^{ki}} \otimes\ket{\textbf{s}_i} $$

Then the log likelihood for the general case could be written as:
\begin{equation}
L= -\frac 1 2\sum_{k,i}\log\lambda_{k,i}-\frac 1 2 \sum_{k,i} \frac{\left( \sum_{\alpha,\beta} u^{ki}_{\alpha} s^{i}_{\beta} y_{\alpha\beta} \right) ^2}{\lambda_{k,i}}
\label{eq: log_lik ND}
\end{equation}
where $k,\alpha \in (1:R)$ and $i,\beta \in (1:N)$ then re-arranging the second term we get,

\begin{equation}
L(\mathbf{y}|\rho_i,\mathbf{s}_i,\gamma)= -\frac 1 2\sum_{k,i}\log\lambda_{k,i}-\frac{1}{2}  \sum_{i} \mathbf{s}_i^{T} \mathbf{A}^{i} \mathbf{s}_i
\label{eq: log_lik ND_new_not}
\end{equation}

where

\begin{equation}
A^{i}_{\beta\delta}=\sum_{k,\alpha,\gamma} \frac{u^{ki}_{\alpha}  u^{ki}_{\gamma}y_{\alpha\beta} y_{\gamma\delta}}{\lambda_{ki}}
\end{equation}

\subsubsection{Regularization}
An $l2$-regularization term, with strength $r_J=10^{-2}$ , is added to expression \ref{eq: log_lik ND_new_not} in order to enforce strict concavity of the log-likelihood function and to remove infinite values of couplings motivated by  overfitting. This term correspond to gaussian prior information describing sparsity of the coupling matrix and leads to the objective function:

\begin{equation}
L(\mathbf{y}|\rho_i,\mathbf{s}_i,\gamma)= -\frac 1 2\sum_{k,i}\log\lambda_{k,i}-\frac{1}{2}  \sum_{i} \mathbf{s}_i^{T} \mathbf{A}^{i} \mathbf{s}_i- r_J*\sum_{i} \rho^2_i
\label{eq: log_lik_ND_regul}
\end{equation}

\subsection{Optimization scheme}

Equation \ref{eq: log_lik_ND_regul}  can be maximized by exploring the space of the coupling matrices by varying its $ N $ eigenvectors $\mathbf{s}_i$  and $ N $ eigenvalues $\rho_i$ in the positive direction of the log likelihood gradient .

As $ \mathbf{s}_i$ is a basis set in an n-dimensional linear space,  our optimization problem is subject to the orthogonality constraint on the set $ \mathbf{s}_i,i=1,2,...,N$. This implies that the transformation of a base formed by the vectors $\mathbf{s}_i$ is established by an orthogonal matrix.

An arbitrary orthogonal matrix $\bm S$ can be parameterized  representing its elements in terms of $N(N-1)/2$ independent variables $ \left\lbrace \bm \phi\right\rbrace $ and then each base vector is parameterized according to the $ith$ column of the  matrix. Optimization of log-likelihood,  $L(\boldmath{\bm \phi})\equiv L(\bm S(\boldmath{\bm \phi})) $, on these variables will impose on our problem the  orthogonality constraint without  to introduce Lagrange multipliers in our variational problem. Gradients of the scalar function on this variables  must be computed using the chain-rule transformation:

\begin{equation}
\frac{\partial L}{\partial \phi_\mu}=\sum_{p,q} \frac{\partial L}{\partial S_{p,q}}\frac{\partial S_{p,q}}{\partial \phi_\mu}
\end{equation}

There are several methods to perform this procedure \cite{Shepard_param_OM}, we implement here a parameterization algorithm of an orthogonal matrix in terms of generalized Eulerian angles \cite{Raffenetti1970GEA}, which establishes a recursive procedure that allows the computation of matrix elements as a function of $N(N-1)/2$  independent angular variables $\theta_ {pq}, p =1,2, ..., N; q =1,2, ..., N; p <q $, and also the orthogonal matrix derivative with respect to these angular variables. This allow to compute the gradient as:

\begin{equation}
\nabla L=\sum_i\frac{\partial L}{\partial \rho_i}\hat{\rho_i} +\sum_{p,q}\frac{\partial L}{\partial \theta_{pq}}  \hat{\theta}_{pq} +\frac{\partial L}{\partial \gamma}\hat{\gamma} 
\end{equation}

where 
\begin{equation}
\frac{\partial L}{\partial \rho_i} = -\frac{1}{2}\left\lbrace  \sum_k \frac{1}{\lambda_{ki}}\frac{\partial \lambda_{ki}}{\partial\rho_i} +\mathbf{s}_i^{T} \frac{\partial \mathbf{A}^{i} }{\partial\rho_i}\mathbf{s}_i \right\rbrace -2 r_{J} \rho_i
\end{equation}

\begin{equation}
\frac{\partial L}{\partial \gamma} = -\frac{1}{2}\left\lbrace  \sum_{k,i} \frac{1}{\lambda_{ki}}\frac{\partial \lambda_{ki}}{\partial\gamma} +\mathbf{s}_i^{T} \frac{\partial \mathbf{A}^{i} }{\partial\gamma}\mathbf{s}_i \right\rbrace 
\end{equation}

and

\begin{equation}
\frac{\partial L}{\partial \theta_{p,q} }=-\sum_{i} \frac{\partial \mathbf{s}_i^{T} }{\partial \theta_{p,q} } \mathbf{A}^{i}  \mathbf{s}_i
\end{equation}

For  gradient evaluation,derivatives of the eigenvalues $\lambda_ {k, i} $ and eigenvectors $u^{k, i}_{\alpha} $ of the matrix $\bm G(\rho_i) $ respect $\rho_i$ parameter must be computed:

\begin{eqnarray}
\nonumber
\lefteqn{\frac{\partial \mathbf{A}^{i} }{\partial\rho_i}=2 \sum_{k} \frac{1}{\lambda_{k,i}}\sum_{\alpha} (u_{\alpha}^{k,i})' y_{\alpha\beta}\sum_{\gamma} u^{k,i}_{\gamma} y_{\gamma\delta}}\\&&-\frac{1}{\lambda_{k,i}^2} \lambda'_{k,i}\sum_{\alpha}u^{k,i}_{\alpha}y_{\alpha\beta}\sum_{\gamma} u^{k,i}_{\gamma} y_{\gamma\delta}
\end{eqnarray}
where

\begin{equation}\lambda'_{k,i}=\frac{\partial \lambda_k(\rho_i) }{\partial\rho_i}\end{equation}

\begin{equation}(u_{\alpha}^{k,i}) ' =\frac{\partial u^{k}_{\alpha}(\rho_i) }{\partial\rho_i}\end{equation}

If the eigenvalues $\lambda_k(\rho_i)$ and eigenvectors $\bm{u}^{k}(\rho_i)$ of  $\bm G(\rho_i)$ are  already available and all eigenvalues are distinct (as in the case of a non-homogenous phylogenetic tree) a simple analytical expression can be derived for the calculation of its corresponding derivatives with respect to the $\rho_i$ parameter \cite{matrix_cook_book} :

\begin{equation}
\frac{\partial \lambda_k(\rho_i) }{\partial\rho_i}=\bm{u}^{k}(\rho_i) ^T\frac{\partial \bm G(\rho_i) }{\partial\rho_i} \bm {u}^{k}(\rho_i) 
\end{equation}
\begin{eqnarray}
\nonumber
\frac{\partial u^{k}_{\alpha}(\rho_i)}{\partial\rho_i}=\sum_{p\neq k}\frac{\left( u^{p}_{\alpha}(\rho_i)^T\frac{\partial \bm G(\rho_i) }{\partial\rho_i} u^{k}_{\alpha}(\rho_i)\right) }{\lambda_{k}(\rho_i)-\lambda_{p}(\rho_i)} u^{p}_{\alpha}(\rho_i)
\end{eqnarray}

Using previous expressions it is possible to maximize the log-likelihood \ref{eq: log_lik_ND_regul}  over parameters: $\gamma$, $\rho_i, i = 1,2, .. N $ and $\theta_ {pq}, p = 1,2, ..., N; q = 1,2,..., N; p <q $. This nonlinear optimization problem was addressed by implementing a quasi-Newton Method algorithms \cite{NLopt}.
\subsection{Initializing parameters}
From empirical covariance matrix $c^*=\frac{1}{N}\sum_{n=1}^{N}\mathbf{x}_n \mathbf{x}_n^T$ we could extract information  to  initialize the parameters in the optimization process.

The eigenvalues of empirical couplings $j^*=1/c^*$  will define $\rho^0_i$ ands its eigenvectors  an orthogonal matrix $\bm S^0$ which satisfy the equations systems:
\begin{equation}
S_{ij}(\theta^0)=S^0_{ij}
\label{recurrent_eq}
\end{equation}
to find the orthogonal parameters which reproduce the initial orthogonal matrix, equation \ref{recurrent_eq} must be inverted or equivalently the solution set $\bm \theta^0$ could be numerically obtained minimizing the function:

\begin{equation}
f(\bm \theta^0)=\sum_{i,j}\left[ S^0_{ij}-S_{ij}(\bm \theta^0)\right] ^2
\label{obtain_theta0}
\end{equation}

\subsubsection{Empirical parameter $\gamma_0$}
The probability distribution for two leaves $\mathbf{x}_1$ and  $\mathbf{x}_2$  separated by time $t_{12}$ , $P(\mathbf{x}_1,\mathbf{x}_2|t_{12})$ is given by the equation (12) for $R=2$. With this distribution we can calculate analytically the average of the scalar product $\mathbf{x}_1\cdot\mathbf{ x}_2$:

\begin{eqnarray}
\nonumber
\lefteqn{\left\langle \mathbf{x}_1\cdot\mathbf{ x}_2\right\rangle = \int \mathbf{dx}_1 \mathbf{dx}_2 \sum_{i} x_1(i) x_2(i)  P(\mathbf{ x}_1,\mathbf{ x}_2|t_{12})}\\\nonumber&&=\sum_{i} \left\langle  x_1(i) x_2(i)\right\rangle_{t_{12}}
=\sum_{i} \left( \bm \Lambda^{t_{12}} \bm J^{-1}\right)_{ii}=\Tr \left(\bm \Lambda^{t_{12}} \bm J^{-1} \right)  
\end{eqnarray}

If the eigensystem of $\bm J$ is known : $\left\lbrace\rho_i, \mathbf{s}_i\right\rbrace $, we get:

\begin{equation}
\left\langle \mathbf{ x}_1\cdot\mathbf{ x}_2\right\rangle =\Tr \left(\bm \Lambda^{t_{12} } \bm J^{-1}\right) =\sum_{i} \frac{ e^{-\frac{\rho_i t_{12}}{\gamma}}}{\rho_i}
\end{equation}

then as we know from the data the empirical values of the pairs $\left\lbrace t,\mathbf{ x}_1\cdot\mathbf{x}_2 \right\rbrace $ we could fit the parameter $\gamma$ minimizing the  function:

\begin{equation}
F(\gamma)=\sum_{a,b}\left[\mathbf{ x}_b\mathbf{ x}_b -\sum_{i} \frac{e^{-\frac{\rho_i t_{a,b}}{\gamma}}}{\rho_i}\right] ^2
\label{gamma_inverse_eq}
\end{equation}

using $\rho_i^0$ in minimization of \ref{gamma_inverse_eq} we obtain the empirical value $\gamma_0$.

An algorithm based on gradient descent method is used  to minimize functions \ref{obtain_theta0} and \ref{gamma_inverse_eq}.
\section{ Results}
\label{Results}
In order to test the methodology developed , we generate  data according to algorithm \ref{alg_1} using a non-homogeneous tree with branch length randomly selected,  root configuration was drawn from a Gaussian distribution $P^0(\mathbf{ x})\sim N(0,\bm J^{-1})$. The couplings matrix was selected as a positive semidefinite matrix drawn from a normal distribution with $\mu_J=0.8$ and $\sigma_{J}=0.2$. To make the matrix sparse we dilute the couplings with density $0.3$.

We choose three different values of parameter $\gamma$ in order to simulate  different regimes: with weak  $\gamma_w=0.1\gamma_c$ ,intermediate $\gamma_i=0.3\gamma_c$ and  strong  $\gamma_s=0.5\gamma_c$  phylogenetic signal. For each different  regime, $30$ repetitions  for the sampling-inference process were made obtaining covariance and coupling matrices using our inference procedure and the empirical estimators.

In figures \ref{results_gamma10_n10},  \ref{results_corr_map_J_10_n10}, \ref{results_corr_map_C_10_n10} we compare the two estimation procedure in term of distance and correlation with the original parameters. As can be seen in   our method is able of inferring covariance and  coupling matrices closer to the equilibrium ones for all regimes, and  differences are more clear if there is more phylogeny contamination in  alignments.

On the other hand if we use the original coupling matrix to create a  contact map between sites, assuming as contact non zero entries, then we could explore the ability of our method to predict topology of the coevolutionary network.  Figure \ref{PPV_n10} show positive predictive values averaged over 30 samples, been the   log-likelihood maximization  a better approach to the contact structure of the original couplings matrix.

 About computations made is useful highlight that as  small dimensions matrices were chosen ( $N=10$) these are not sparse enough, so the prediction of contacts may not be sufficiently enlightening in the comparison.

\section{Conclusions }
\label{conc}

\textbf{Acknowledgments:}

\begin{figure*}[!htb]
	\begin{minipage}[b]{0.325\textwidth}
		\begin{center}
			\includegraphics[keepaspectratio=true,width=1\textwidth]{gamma_N10_M1024_optJ_30samp_gamma_c_error_label.pdf}
		\end{center}
	\end{minipage}
	%-------------------------------------
	%-------------------------------------
	\begin{minipage}[b]{0.325\textwidth}
		\begin{center}
			\includegraphics[keepaspectratio=true,width=1\textwidth]{gamma_N10_M1024_optJ_30samp_3gamma_c_error_label.pdf}
		\end{center}
	\end{minipage}
	%-------------------------------------
	%-------------------------------------
	\begin{minipage}[b]{0.325\textwidth}
		\begin{center}
			\includegraphics[keepaspectratio=true,width=1\textwidth]{gamma_N10_M1024_optJ_30samp_5gamma_c_error_label.pdf}
		\end{center}
	\end{minipage}

	\vspace{-1mm}
	\caption{{ Inference of parameters $\gamma$ for $N=10$ and tree diferent regimes \textbf{left}: $\gamma=\gamma_w$, \textbf{center}: $\gamma=\gamma_i$, \textbf{right} $\gamma=\gamma_s$  }} .
	\label{results_gamma10_n10}
\end{figure*}

\begin{figure*}[!htb]
	\begin{minipage}[b]{0.325\textwidth}
		\begin{center}
			\includegraphics[keepaspectratio=true,width=1\textwidth]{frob_dist_vs_pearson_coup_N10_M1024_optJ_30samp_gamma_c.pdf}
		\end{center}
	\end{minipage}
	%-------------------------------------
	%-------------------------------------
	\begin{minipage}[b]{0.325\textwidth}
		\begin{center}
			\includegraphics[keepaspectratio=true,width=1\textwidth]{frob_dist_vs_pearson_coup_N10_M1024_optJ_30samp_3gamma_c.pdf}
		\end{center}
	\end{minipage}
	%-------------------------------------
	%-------------------------------------
	\begin{minipage}[b]{0.325\textwidth}
		\begin{center}
			\includegraphics[keepaspectratio=true,width=1\textwidth]{frob_dist_vs_pearson_coup_N10_M1024_optJ_30samp_5gamma_c.pdf}
		\end{center}
	\end{minipage}

	\vspace{-1mm}
	\caption{{ Fobenious distance vs slope of linear regression between empirical/infered couplings matrix and true couplings matrix   for $N=10$ and tree diferent regimes \textbf{left}: $\gamma=\gamma_w$, \textbf{center}: $\gamma=\gamma_i$, \textbf{right} $\gamma=\gamma_s$  }} .
	\label{results_corr_map_J_10_n10}
\end{figure*}

\begin{figure*}[!htb]
	\begin{minipage}[b]{0.325\textwidth}
		\begin{center}
			\includegraphics[keepaspectratio=true,width=1\textwidth]{lin_reg_vs_pearson_corr_N10_M1024_optJ_30samp_gamma_c.pdf}
		\end{center}
	\end{minipage}
	%-------------------------------------
	%-------------------------------------
	\begin{minipage}[b]{0.325\textwidth}
		\begin{center}
			\includegraphics[keepaspectratio=true,width=1\textwidth]{lin_reg_vs_pearson_corr_N10_M1024_optJ_30samp_2gamma_c.pdf}
		\end{center}
	\end{minipage}
	%-------------------------------------
	%-------------------------------------
	\begin{minipage}[b]{0.325\textwidth}
		\begin{center}
			\includegraphics[keepaspectratio=true,width=1\textwidth]{lin_reg_vs_pearson_corr_N10_M1024_optJ_30samp_5gamma_c.pdf}
		\end{center}
	\end{minipage}

	\vspace{-1mm}
	\caption{{   Pearson correlation vs slope of linear regression between empirical/infered covariance matrix and true couplings matrix   for $N=10$ and tree diferent regimes \textbf{left}: $\gamma=\gamma_w$, \textbf{center}: $\gamma=\gamma_i$, \textbf{right} $\gamma=\gamma_s$  }} .
	\label{results_corr_map_C_10_n10}
\end{figure*}

\begin{figure*}[!htb]
	\begin{minipage}[b]{0.325\textwidth}
		\begin{center}
			\includegraphics[keepaspectratio=true,width=1\textwidth]{PPV_av30_N10_M1024_optJ_gamma_c_no_diag.pdf}
		\end{center}
	\end{minipage}
	%-------------------------------------
	%-------------------------------------
	\begin{minipage}[b]{0.325\textwidth}
		\begin{center}
			\includegraphics[keepaspectratio=true,width=1\textwidth]{PPV_av30_N10_M1024_optJ_3gamma_c_no_diag.pdf}
		\end{center}
	\end{minipage}
	%-------------------------------------
	%-------------------------------------
	\begin{minipage}[b]{0.325\textwidth}
		\begin{center}
			\includegraphics[keepaspectratio=true,width=1\textwidth]{PPV_av30_N10_M1024_optJ_5gamma_c_no_diag.pdf}
		\end{center}
	\end{minipage}

	\vspace{-1mm}
	\caption{{  Positive predictive values averaged over 30 samples,  for $N=10$ and tree diferent regimes \textbf{left}: $\gamma=\gamma_w$, \textbf{center}: $\gamma=\gamma_i$, \textbf{right} $\gamma=\gamma_s$  }} .
	\label{PPV_n10}
\end{figure*}

%\begin{figure*}[!htb]
%	\begin{minipage}[b]{0.325\textwidth}
%		\begin{center}
%			\includegraphics[keepaspectratio=true,width=1\textwidth]{Images/new_im/cor_scatter_n6_M1024_gamma10_37samp.png}
%		\end{center}
%	\end{minipage}
%	%-------------------------------------
%	%-------------------------------------
%	\begin{minipage}[b]{0.325\textwidth}
%		\begin{center}
%			\includegraphics[keepaspectratio=true,width=1\textwidth]{Images/new_im/frob_dist_vs_slope_C_n6_gamma10_37samples.png}
%		\end{center}
%	\end{minipage}
%	%-------------------------------------
%	%-------------------------------------
%	\begin{minipage}[b]{0.325\textwidth}
%		\begin{center}
%			\includegraphics[keepaspectratio=true,width=1\textwidth]{Images/new_im/cor_corr_map_n6_M1024_gamma10_37samp.png}
%		\end{center}
%	\end{minipage}
%	\\
%	\vspace{1mm}
%	\begin{minipage}[b]{0.325\textwidth}
%		\begin{center}
%			\includegraphics[keepaspectratio,width=1\textwidth]{Images/new_im/frob_dist_vs_slope_J_n6_gamma10_37samples.png}
%		\end{center}
%	\end{minipage}
%	%-------------------------------------
%	%-------------------------------------
%	\begin{minipage}[b]{0.325\textwidth}
%		\begin{center}
%			\includegraphics[keepaspectratio=true,width=1\textwidth]{Images/new_im/coup_corr_map_n6_M1024_gamma10_37samp.png}
%		\end{center}
%	\end{minipage}
%	%-------------------------------------
%	%-------------------------------------
%	\begin{minipage}[b]{0.325\textwidth}
%		\begin{center}
%			\includegraphics[keepaspectratio=true,width=1\textwidth]{Images/new_im/coup_ppv_n6_M1024_gamma10_37samp.png}
%		\end{center}
%	\end{minipage}
%	\\
%	\vspace{-1mm}
%	\caption{{ covariance and coupling matrices comparison for the regime with weak phylogenetic bias:$N=6$ and $\gamma=10$   }}
%	\label{results_gamma10_n6}
%\end{figure*}
%
%
%
%\begin{figure*}[!htb]
%	\begin{minipage}[b]{0.325\textwidth}
%		\begin{center}
%			\includegraphics[keepaspectratio=true,width=1\textwidth]{Images/new_im/cor_scatter_n6_M1024_gamma100_37samp.png}
%		\end{center}
%	\end{minipage}
%	%-------------------------------------
%	%-------------------------------------
%	\begin{minipage}[b]{0.325\textwidth}
%		\begin{center}
%			\includegraphics[keepaspectratio=true,width=1\textwidth]{Images/new_im/frob_dist_vs_slope_C_n6_gamma100_37samples.png}
%		\end{center}
%	\end{minipage}
%	%-------------------------------------
%	%-------------------------------------
%	\begin{minipage}[b]{0.325\textwidth}
%		\begin{center}
%			\includegraphics[keepaspectratio=true,width=1\textwidth]{Images/new_im/cor_corr_map_n6_M1024_gamma100_37samp.png}
%		\end{center}
%	\end{minipage}
%	\\
%	\vspace{1mm}
%	\begin{minipage}[b]{0.325\textwidth}
%		\begin{center}
%			\includegraphics[keepaspectratio=true,width=1\textwidth]{Images/new_im/frob_dist_vs_slope_J_n6_gamma100_37samples.png}
%		\end{center}
%	\end{minipage}
%	%-------------------------------------
%	%-------------------------------------
%	\begin{minipage}[b]{0.325\textwidth}
%		\begin{center}
%			\includegraphics[keepaspectratio=true,width=1\textwidth]{Images/new_im/coup_scatter_n6_M1024_gamma100_37samp.png}
%		\end{center}
%	\end{minipage}
%	%-------------------------------------
%	%-------------------------------------
%	\begin{minipage}[b]{0.325\textwidth}
%		\begin{center}
%			\includegraphics[keepaspectratio=true,width=1\textwidth]{Images/new_im/coup_ppv_n6_M1024_gamma100_37samp.png}
%		\end{center}
%	\end{minipage}
%	\\
%	\vspace{-1mm}
%	\caption{{ covariance and coupling matrices comparison for the regime with strong phylogenetic bias: $N=6$ and $\gamma=100$   }}
%	\label{results_gamma100_n6}
%\end{figure*}
%
%\begin{figure*}[!htb]
%	\begin{minipage}[b]{0.325\textwidth}
%		\begin{center}
%			\includegraphics[keepaspectratio=true,width=1\textwidth]{Images/new_im/gamma_N10_M1024_optJandC_36sim.pdf}
%		\end{center}
%	\end{minipage}
%	%-------------------------------------
%	%-------------------------------------
%	\begin{minipage}[b]{0.325\textwidth}
%		\begin{center}
%			\includegraphics[keepaspectratio=true,width=1\textwidth]{Images/new_im/frob_dist_vs_pearson_corr_N10_M1024_optJandC_36sim.pdf}
%		\end{center}
%	\end{minipage}
%	%-------------------------------------
%	%-------------------------------------
%	\begin{minipage}[b]{0.325\textwidth}
%		\begin{center}
%			\includegraphics[keepaspectratio=true,width=1\textwidth]{Images/new_im/lin_reg_vs_pearson_corr_N10_M1024_optJandC_36sim.pdf}
%		\end{center}
%	\end{minipage}
%	\\
%	\vspace{1mm}
%	\begin{minipage}[b]{0.325\textwidth}
%		\begin{center}
%			\includegraphics[keepaspectratio=true,width=1\textwidth]{Images/new_im/frob_dist_vs_pearson_coup_N10_M1024_optJandC_36sim.pdf}
%		\end{center}
%	\end{minipage}
%	%-------------------------------------
%	%-------------------------------------
%	\begin{minipage}[b]{0.325\textwidth}
%		\begin{center}
%			\includegraphics[keepaspectratio=true,width=1\textwidth]{Images/new_im/scatter_all_coup_mat_N10_M1024_optJ_andC_36sim.pdf}
%		\end{center}
%	\end{minipage}
%	%-------------------------------------
%	%-------------------------------------
%	\begin{minipage}[b]{0.325\textwidth}
%		\begin{center}
%			\includegraphics[keepaspectratio=true,width=1\textwidth]{Images/new_im/PPV_av_N10_M1024_optJandC_36sim.pdf}
%		\end{center}
%	\end{minipage}
%	\\
%	\vspace{-1mm}
%	\caption{{ covariance and coupling matrices comparison for :$N=10$ and $\gamma=10$   }}
%	\label{results_gamma10_n10}
%\end{figure*}

\newpage

\bibliography{bib_phylo}

\bibliographystyle{unsrt}

\end{document}

### 2020-07-24 main.tex
First commit

%\documentclass[reprint,amsmath,amssymb,superscriptaddress,prl,showpacs,onecolumn]{revtex4-1}
\documentclass[preprint,amsmath,amssymb,superscriptaddress,showpacs,pre]{revtex4-1}

\usepackage{graphicx}% Include figure files
\usepackage{dcolumn}% Align table columns on decimal point
\usepackage{bm}% bold math
\usepackage{color}
\usepackage{epstopdf} 
\usepackage{amsmath,amsfonts,amssymb,amsthm}
% \usepackage{hyperref}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{braket}
\usepackage[sans]{dsfont}
\usepackage{mathbbol}
\usepackage{bbm}

%% Added by Pierre
\usepackage{lineno}
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography

\linenumbers

\DeclareMathOperator\atanh{arctanh}
\let\originalleft\left
\let\originalright\right
\renewcommand{\left}{\mathopen{}\mathclose\bgroup\originalleft}
\renewcommand{\right}{\aftergroup\egroup\originalright}
%\newcommand{\bra}[1]{\ensuremath{\left< #1\right|}}
%\newcommand{\ket}[1]{\ensuremath{\left|#1\right>}}
\newcommand{\ovl}[2]{\ensuremath{\left\langle #1\middle|#2\right\rangle}}
\newcommand{\ie}{\emph{i.e.}}
\newcommand{\eg}{\emph{e.g.}}
\newcommand{\cf}{cf.}
\newcommand{\cp}{\emph{cp.}}
%\newcommand{\braket}[2]{\left< #1,#2\right>}
%%\newcommand{\<}{\langle}
%%\renewcommand{\>}{\rangle}
\providecommand{\abs}[1]{\ensuremath{\left|#1\right|}}
\providecommand{\norm}[1]{||#1||}
\newcommand{\arctanh}[1]{\mathrm{arctanh}#1}
\newcommand{\Sign}[1]{\mathrm{Sign}#1}

\def\<{\langle}
\def\>{\rangle}
\def\vx{\vec x}
\def\vh{\vec h}

\newcommand{\ud}{\mathrm{d}}
\newcommand{\etal}{\textit{ et. al. }}
\newcommand{\Ham}{\mathcal{H}}
\newcommand{\B}{\beta}
\def\matF{{\mathbb F}}
\def\Jij{J_{ij}}

\def\Imat{\mathbb{I}}
\def\ext{\qopname\relax m{ext}}

\newcommand{\mi}{\mathrm{i}}
\DeclareMathOperator{\Tr}{Tr}

%%%%

\newcommand{\ddroit}{\textrm{d}}
\newcommand{\Ptot}{P(A_1\ldots A_L)}
\newcommand{\xz}{\vec{x}_0}
\newcommand{\xo}{\vec{x}_1}
\newcommand{\xt}{\vec{x}_2}
\newcommand{\Lam}{\bm{\Lambda}}
\newcommand{\Sig}{\bm{\Sigma}}
\newcommand{\Sigmone}{\bm{\Sigma^{-1}}}
\def\fonetwo{\frac 1 2}
\def\Cmone{C^{-1}}
\def\Lambtwo{\Lambda^2} 
\def\tildex{\tilde{x}}
\newcommand{\HRule}{\rule{\linewidth}{0.5mm}}
\newcommand{\Xk}[1]{X^{\{#1\}}}
\newcommand{\xk}[1]{\vec{x}_{#1}}

% Pierre
\newcommand{\pierre}[1]{{\color{red}Pierre: #1}}
\newcommand{\curlynormalpar}[1]{\exp\left\{-\frac{1}{2}\left( #1 \right)\right\}}
\newcommand{\curlynormal}[1]{\exp\left\{-\frac{1}{2} #1 \right\}}
\newcommand{\iSig}{\bm{\Sigma^{-1}}}
\newcommand{\iC}{\bm{C}^{-1}}
\newcommand{\av}[1]{\left\langle #1 \right\rangle}
\newcommand{\vsa}{\vec{s}_a}
\newcommand{\vuka}{\vec{u}_{ka}}

\begin{document}

\title{Inference of global coevolutionary models from hierarchical correlated data.}
\date{}

\author{Edwin Rodriguez Horta} 
\affiliation{Group of Complex Systems and Statistical Physics, Department of Theoretical Physics, University of Havana, Cuba}
\affiliation{Laboratoire de Biologie Computationnelle et Quantitative (LCQB), Sorbonne Université, France}
%
\author{Pierre Barrat-Charlaix} 
\affiliation{Biozentrum, Universität Basel, Switzerland}
%
\author{Alejandro Lage} 
\affiliation{Group of Complex Systems and Statistical Physics, Department of Theoretical Physics, University of Havana, Cuba}
%
\author{Martin Weigt} 
\email{Correspondence to: Martin Weigt, \bf{martin.weigt@upmc.fr}}
\affiliation{Laboratoire de Biologie Computationnelle et Quantitative, Sorbonne Université, France}

\begin{abstract}
	The Inverse problem of Statistical Physics infers maximum-entropy  models  compatibles with a corresponding set of empirical averages extracted from a high-dimensional dataset of equilibrium configurations. Practical interest extended these methods to non-equilibrium data  like  time series, where detailed balance does not hold. However, in several applications, data samples  result from evolutionary processes and the relation between configurations is conditioned by a hierarchical structure of the population. This makes sample statistics a superposition of  signals coming from both: internal configuration correlations and relatedness correlations produced by common evolutionary  history, typically represented by a tree. How to disentangle both sources of correlation for a better parameter inference is an open question to explore. Here we propose an approach  where the  evolutive process through a phylogenetic tree is described by  a  multivariate Ornstein-Uhlenbeck dynamics characterized by  gaussian propagator and stationary distributions. Inference of these distributions from  phylogenetic non-independent samples is solved in a Bayesian framework. This procedure can be  extended to discrete variables by using a binary representation of data and approximating  binary sates by continuous variables. Our method proposes a possible way for  a better estimation of both, equilibrium and dynamic parameters, making  applications more accurate. 
\end{abstract}

\maketitle

\section{Introduction}
\label{sec:int}
Theory and applications of the inverse problem of statistical physics has had great development in recent years since the availability of large amounts of data on a microscopic scale produced by the continuous technological progress. In this context statistical physics paths: from model parameters to  observables, is reversed, toward find global models which describe the statistics of  a set of observed  configurations \cite{Inverse_problem_Berg}.

Most of the work has been focus in equilibrium reconstruction , where the aim is to learn max Entropy models, Boltzman distribution, from independent and identically distributed equilibrium configurations. Another kind of methods deal with parameters learning from either time series data or from samples of the non-equilibrium steady state: non-equilibrium reconstruction.

Less effort has been done for data biased by population structured signal, where  samples  are not independent because of presence of relatedness between configurations. These structure could be for instance of  geographical or demogrhapic  nature in a population or  phylogenetic  nature  related to a evolutionary processes shaping trait variation among species.  Here we are going to deal with data given by sequences phylogenetically related  through a tree, which describes the hierarchical correlation between configurations. An important problem is the  distinction   of patterns induced by species traits or by phylogeny.

Our source of inspiration came from the inverse problem in proteins \cite{Inverse_problem_proteins}, where  covariance signal present in an alignment of homologous proteins is described via Potts model. This covariance signal is assumed a consequence of phenotypics constraints like protein structure and function. However it is known that phylogeny relationships in homologous sequences  contaminate the statistics of  the alignment used to learn the equilibrium model. Therefore a direct inference on this alignments  leads to the existence of Potts parameters that attempt to model the full biased statistics limiting their capacity to predict phenotypic properties.

 To define and solve the inverse problem of statistical physics for data samples with hierarchical  relatedness structure is the aim of this work.

Main difficulties to address  this kind of problem  come from the necessity to use  a proper global evolutionary model which describes the relation between sequences edge connected in a tree and the discrete  characters of data variables which makes certain calculations intractable \cite{entropy_paper}.

Here we  adopt an approximation where we forget about the discrete nature of variables and model them by continuous variables instead. This transforms the Potts model into a Gaussian distribution \cite{Baldassi}, making the design of a global propagator tractable. We propose a model of evolution described by a multivariate Ornstein-Uhlenbeck process \cite{gardiner}, which is used  to generate evolutionary related data given an arbitrary phylogenetic tree. The corresponding inverse problem became in the inference of the Gaussian potential (Potts parameters) from the knowledge of tree leaves configurations and the dynamics of evolution.

To this aim, we first review in section \ref{sec:ornstein_uhlenbeck_dynamics}, main characteristics of a multivariate Ornstein-Uhlenbeck process. In section \ref{evolutionary process} we describe the evolutionary process, describing the algorithm used to generate data phylogenetically biased. The statement of the problem in a Bayesian framework is shown in section \ref{statement of the problem}. Section \ref{methods} is dedicated to describe methodology used to solve the inverse problem. Results for evolution of continuos variable sequences and  the application for configuration of discrete variables is presented in section \ref{Results}. Finally the conclusion and outlook in \ref{conc}.

\section{The multivariate Ornstein-Uhlenbeck process}
\label{sec:ornstein_uhlenbeck_dynamics}

We consider a system characterized by $L$ continuous degrees of freedom and whose state is fully described by a vector $\vx$ of length $L$. 
These degrees of freedom can be continuous phenotypic traits of some living organism, or the sequence of a gene or a protein if a continuous approximation is made. 
At equilibrium, $\vx$ is normally distributed:
\begin{equation}
	P_{eq}(\vx) = \frac{1}{Z(\bm{J})}\curlynormal{\vx^T\bm{J}\vx},
	\label{eq:eq_distribution}
\end{equation}
where $\bm{J}$ is the symmetric positive definite \emph{coupling matrix} and $Z(\bm{J})$ is a normalization constant. 
We are interested in inferring this coupling matrix from a given amount of observed states $\vx$ of the system.
If these observations were independent from each other, $\bm{J}$ would simply be equal to the inverse of the \emph{covariance matrix} of the data, written $\bm{C}$.

However, we are concerned here with the case where these observations are not independent. 
On the contrary, they are the result of a dynamical process which took place during a finite amount of time, and is thus not necessarily equilibrated. 
This dynamical process is described below.

We suppose that the considered system evolves according to the following Langevin equation 
\begin{equation}
	\gamma^{-1}\frac{\text{d}\vx}{\text{d}t} = - \bm{J}\vx + \vec{\xi}(t).
	\label{eq:langevin}
\end{equation}
Here, $\vec{\xi}(t)$ is a vector of uncorrelated white noises, and $\gamma^{-1}$ is the characteristic timescale governing the dynamics. 
In short, equation~\ref{eq:langevin} states that the system described by $\vx$ undergoes Brownian motion in an energy landscape characterized by the coupling matrix $\bm{J}$.

We are not interested in $\vx$ directly, but rather by its probability distribution $P(\vx,t\vert\,\vx_0)$, that is the probability to find the system in state $\vx$ knowing it was in state $\vx_0$ and evolved for time $t$. 
The Fokker-Planck equation corresponding to equation~\ref{eq:langevin} is straightforward to write: 
\begin{equation}
	\gamma \partial_t P = \left(-\sum_{a,b=1}^L\frac{\partial}{\partial x_a} J_{ab}x_b + \sum_{a=1}^L\frac{\partial^2}{\partial x_a^2}\right) P, 
	\label{eq:FP}
\end{equation}
where the parenthesized expression on the right hand side is understood as an operator on $P$. 
The solution to~\ref{eq:FP} is a multivariate normal distribution~\cite{singh2017multiOU}:
\begin{equation}
	P(\vx, \Delta t|\, \vx_0) = \left((2\pi)^N\det\iSig\right)^{-1/2}
	\curlynormal{(\vx-\vec{\mu})^T\iSig(\vx- \vec{\mu})},
	\label{eq:OUpropagator}
\end{equation}
where we introduce matrices $\Sig$ and $\Lam$ as well as vector $\vec{\mu}$: 
\begin{equation}
	\Lam = e^{-\gamma\bm{J}}, \qquad \vec{\mu} = \Lam^{\Delta t} \vx_0, \qquad \Sig = \bm{J}^{-1}(\mathbb{1} - \Lam^{2\Delta t}).
	\label{eq:def_lambda}
\end{equation}
Equations~\ref{eq:OUpropagator} and~\ref{eq:def_lambda} define a multivariate \emph{Ornstein-Uhlenbeck} process (OU).

It is important to note that since matrix $\Lam$  is the exponential of $-\bm{J}$, it is symmetric, has strictly negative eigenvalues and commutes with $\bm{J}$. 
We also underline the fact that $\Sig$ and $\vec{mu}$ depend on $\Delta t$, although this dependence is absent in our notation.

By taking $\gamma\Delta t \gg 1$ and using the fact that $\bm{J}$ has strictly positive eigenvalues , one immediately recovers equation~\ref{eq:eq_distribution}. 
This allows us to compute the joint distribution of two configurations $\vx_1$ and $\vx_2$ separated by a time $\Delta t$ by multiplying equations~\ref{eq:eq_distribution} and~\ref{eq:OUpropagator}: 
\begin{equation}
	P(\vx_1, \vx_2\vert\,\Delta t) \propto \curlynormalpar{\vx_1^T\iSig\vx_1 + \vx_2^T\iSig\vx_2 - 2\vx_1^T\Lam^{\Delta t}\iSig\vx_2}. 
	\label{eq:OUjoint}	
\end{equation}
This last equation is important as it allows one to compute the joint covariance of configurations $\vx_1$ and $\vx_2$. 
The probability distribution in~\ref{eq:OUjoint} is a normal distribution whose inverse covariance matrix is defined by blocks: $\Sig$ on the diagonal and $-\Lam^{\Delta t}\Sig$ off-diagonal. 
By inverting this block matrix, given that $\Lam$ and $\Sig$ commute and are invertible, one obtains the following covariance: 
\begin{equation}
	\langle \vx_1\vx_2^T \rangle_{\Delta t} = \Lam^{\Delta t}\bm{J}^{-1} = \Lam^{\Delta t}\bm{C}. 
	\label{eq:pairwisecov}
\end{equation}
As said above, the aim of this work is to infer matrix $\bm{J}$ from a set of observations $\vx$ that are generated this OU process. 
More precisely, any two observations $\vx_1$ and $\vx_2$ will be separated by a time $\Delta t$ during which the OU process took place. 
Equation~\ref{eq:pairwisecov} then allows us to distinguish two regimes. 
Let us call $\{\rho_a\}$ the eigenvalues of $\bm{J}$. 
Using the definition of $\Lam$, we immediatly have that the eigenvalues of matrix $\Lam^{\Delta t}\bm{C}$, \ie  the left hand side of equation~\ref{eq:pairwisecov}, are of the form $\rho_a^{-1}e^{-\gamma\rho_a\Delta t}$.

Since the $\rho_a$ are positive, the eigenvalues of $\Lam^{\Delta t}\bm{C}$ vanish exponentially for large enough times. 
The timescale of this exponential decrease is set by the characteristic time $\tau_c^{-1} = \gamma\rho_{min}$, where $\rho_{min}$ is the smallest eigenvalue of $\bm{J}$. 
Thus, for $\Delta t / \tau_c \gg 1$, $\vx_1$ and $\vx_2$ are uncorrelated. 
If this is verified for all pairs of observations, the regime is that of \emph{weakly correlated} data, and the inference of $\bm{J}$ can simply be performed by inverting the empirical correlation matrix. 
Inversely, for $\Delta t / \tau_c \ll 1$, $\vx_1$ and $\vx_2$ are highly correlated, defining a \emph{strongly correlated} regime. 
It should be noted that when $\Delta t = 0$, the joint correlation matrix of $\vx_1$ and $\vx_2$ becomes non invertible, and equation~\ref{eq:pairwisecov} becomes irrelevant.

\section{Statement of the problem}
\label{statement of the problem}

\begin{figure}[!htb]
	\includegraphics[width=0.6\textwidth]{Figures/tree.pdf}
	\caption{ \pierre{We should get a better figure here.}}
	\label{fig:sample_tree}
\end{figure}

The central problem discussed here is the inference of the probability distribution describing samples that are hierarchically correlated by a tree. 
Formally, we assume that the data consists of $N$ real valued vectors of length $L$, noted $\{\vx_i\}$ with $i\in\{1\ldots N\}$. 
Taken individually, the $\vx_i$s are normally distributed according to equation~\ref{eq:eq_distribution}. with mean $0$.
By construction, the equilibrium correlation between components of vectors $\vx$ is given by the inverse of the couplings, that is $\langle x^a x^b\rangle-\langle x^a\rangle\langle x^b\rangle = \bm C_{ab} = (\bm J^{-1})_{ab}$. 
This implies that inferring the coupling matrix defining the probability distribution amounts to finding the \emph{equilibrium} covariance matrix $\mathbf{C}$.

However, this covariance cannot be directly measured as we consider observations that are not independently distributed.
Instead, the set of measured configurations $\{\vx_i\}$ is the result of an Ornstein-Uhlenbeck (OU) process taking place on a tree $\mathcal{T}$, such as the one shown in figure~\ref{fig:sample_tree}. 
The configuration $\vx_i$ of each node $c$ in the tree is the result of the evolution given by the equation~\ref{eq:langevin}, with initial condition $\vx_{a(i)}$, $a(i)$ being the ancestral node of $i$, and running time $\Delta t_{ia(i)}$, that is the \emph{branch length} between $a(i)$ and $i$. 
This process is initialized with a root note $\vx_r$ that is distributed according to equation~\ref{eq:eq_distribution}.
Observed data points are leaves of this tree, which is assumed to be perfectly known.
The nature of the chosen evolution process implies that the distribution of a leaf configuration $\vx_i$ conditioned on the knowledge of another leaf configuration $\vx_j$ is given by equation~\ref{eq:OUpropagator}, where $\Delta t=\Delta t_{ij}$ represents the total branch length separating leaf $i$ and leaf $j$.

The OU process is fully characterized by the quadratic potential $\bm J = \bm C^{-1}$ and a rate $\gamma$. 
Hence, the configurations of the leaves, that is the data, are determined only by $\bm C$, $\gamma$, and the tree $\mathcal{T}$. 
The aim of this work is to design a method to infer the most likely values of $\bm C$ and $\gamma$ from the knowledge of the data $\mathcal{D}=\{\vx_i\}$ and the shape of $\mathcal{T}$. 
There are two notable extreme cases to this problem. 
The first one is the $\Delta t_r/\tau_c\longrightarrow 0$ case, where $\Delta t_r$ is the typical branch length distance from leaves to root. 
Leaf configurations are then almost identical to the root, \emph{i.e.} $\vx_i\simeq\vx_r$, and inference of $\bm C$ becomes impossible. 
The second one is the $\Delta t_r/\tau_c\longrightarrow\infty$ case, where equilibration of the OU process takes place almost instantly when compared to branch lengths of $\mathcal{T}$.
Leaf configurations are then independently drawn from the equilibrium distribution $P_{eq}$, and $\bm C$ can be readily estimated from the empirical covariances. 
Here, we deal with the intermediary cases where the tree makes it impossible to simply estimate $\bm C$ from empirical covariances, but where $\tau_c$ is not large enough to remove all signatures of the energy potential.

We adopt a Bayesian approach to this inference problem by writing the probability of a given set of parameters as a function of the data using the following relation
\begin{equation}
	P(\bm C, \gamma\vert\mathcal{D},\mathcal{T}) \propto P(\mathcal{D}\vert\bm C, \gamma,\mathcal{T})\cdot P(\bm C, \gamma),
	\label{eq:bayesian_likelihood}
\end{equation}
where the proportionality constant does not depend on the parameters $\bm C$ or $\gamma$, and where $P(\bm C, \gamma)$ can be any arbitrarily chosen prior distribution. The difficulty in equation~\ref{eq:bayesian_likelihood} lies in the estimation of the term $P(\mathcal{D}\vert\bm C, \gamma, \mathcal{T})$, that is the likelihood of the data given a set of parameters $\bm C$ and $\gamma$ and the tree $\mathcal{T}$.  We detail the computation of this likelihood in the following section.

\subsection{Calculation of the likelihood} % (fold)
\label{sub:Calculation_of_the_likelihood}

The joint distribution of two configurations $\vx_1$ and $\vx_2$ separated by time $\Delta t$ is given by equation~\ref{eq:OUjoint} and corresponds to a joint normal distribution. 
This means that the vector $\vec{X}=[\vx_1, \vx_2]$, \emph{i.e.} the concatenation of vectors $\vx_1$ and $\vx_2$, follows a normal distribution with null means.
Of importance here is that this property of the OU process can be extended to the joint distribution of any subset nodes in a tree. 
In other words, if we now define $\vec{X}=[\vx_1,\ldots,\vx_N]$ to be the concatenation of the configurations of all leaves of $\mathcal{T}$, we can write the distribution of $\vec{X}$ as
\begin{equation}
	P(\vec{X}\vert\,\bm{C},\gamma,\mathcal{T}) = \left((2\pi)^{LN}\det\mathbb{G}\right)^{-\frac{1}{2}}\curlynormal{\vec{X}^T\mathbb{G}^{-1}\vec{X}},
	\label{eq:multiOU_jointall}
\end{equation}
where $\mathbb{G}$ is the \emph{joint correlation matrix} and depends on the tree as well as on $\bm{C}$ and $\gamma$.
This property is shown in more detail in section \ref{sec:joint_probability_of_leaves_is_a_gaussian} of the Supplementary Material.\\
Equation~\ref{eq:multiOU_jointall} allows us to compute the log-likelihood of the data $\vec{X}$ as a function of $\vec{X}$ itself and of the joint correlation matrix. 
Indeed, taking its logarithm immediatly gives
\begin{equation}
	\mathcal{L}_{\mathcal{D}}(\mathbb{G}) = -\frac 1 2 \log \det  \mathbb{G}  -\frac 1 2 \vec{X}^T\mathbb{G}^{-1} \vec{X} + \text{cte}. 
	\label{eq:likelihood_2}
\end{equation}

We will now use the fact that the joint distribution of the leaves is a Gaussian to compute $\mathbb{G}$ as a function of the tree and the parameters of the OU process.
The joint correlation matrix is of dimensions $(L\cdot N)\times(L\cdot N)$ where $L$ is the length of individual Gaussian vectors $\vx_i$. 
It describes the correlation of each leaf of the tree to every other leaf, in the sense that the $L\times L$ block $\mathbb{G}_{ij}$ with $1\leq i,j\leq N$ corresponds to the covariance of configurations $\vx_i$ and $\vx_j$. 
In other words, for indices $a$ and $b$ such that $1\leq a,b\leq L$, we have
\begin{equation}
	\mathbb{G}_{ij}(a,b) = \av{x_i^a x_j^b}. 
	\label{eq:jointcov_1}
\end{equation}
Since $\vx_i$ and $\vx_j$ are leaves of the tree, they are separated by an evolution of the OU proces for a time $\Delta t_{ij}$.
This means that without knowledge of the internal nodes, $\vx_i$ and $\vx_j$ are distributed according to equation~\ref{eq:OUjoint}. 
In particular, this implies that their covariance matrix is given by equation \ref{eq:pairwisecov}.
From this we can derive the value of the block $(i,j)$ of the joint correlation matrix, obtaining the following equation for $\mathbb{G}$: 
\begin{equation}
	\mathbb{G}_{ij} =
	\begin{cases}
		\bm C & \text{if $i=j$}.\\
		\Lam^{\Delta t_{ij}}{\mathbf{C}} & \text{otherwise}  
	\end{cases}.
	\label{eq:jointcov_2}
\end{equation}
We remind here that $\Lam = e^{-\gamma\iC}$ depends only on $\gamma$ and $\mathbf{C}$, and commutes with $\mathbf{C}$.  \\

We can now use the structure of $\mathbb{G}$ described in equation~\ref{eq:jointcov_2} to re-write equation~\ref{eq:likelihood_2} in a way that we can compute. 
More precisely, we will attempt to write the likelihood as a function of eigenmodes of $\iC$. \\
Let us first note $\left\{\rho_a, \vsa\right\}$ the eigenvalues and eigenvectors of $\iC$, where the index $a$ runs from $1$ to $L$ and vectors $\vsa$ are of dimention $L$. 
By definition, we have $\rho_a>0$ for all $a$.  
The properties of the exponential of matrices imply that vectors $\vsa$ are also eigenvectors of the matrix $\Lam$. 
Thus, we can easily write the eigenmodes of the block $\mathbb{G}_{ij}$ as $\left\{z(\rho_a, \Delta t_{ij}), \vsa\right\}$ where we introduced
\begin{equation}
	z(\rho_a, \Delta t_{ij}) = \rho_a^{-1}e^{-\gamma\rho_a\Delta t_{ij}}.
	\label{eq:z}
\end{equation}
By convention, $\Delta t_{ii} = 0$ and $z(\rho_a, \Delta t_{ii}) = \rho_a^{-1}$. \\
We also introduce matrix $\mathbf{G}^a$ of dimensions $N\times N$ and defined for indices $1\leq i,j\leq N$ as 
\begin{equation}
	\mathbf{G}^a_{ij} = z(\rho_a, \Delta t_{ij}). 
	\label{eq:subG_def}
\end{equation}
In other words, for a given index $1\leq a\leq L$, $\mathbf{G}^a$ is the matrix built by replacing blocks of $\mathbb{G}$ by their $a$th eigenvalue.
Matrices $\mathbf{G}^a$ are symmetric have their own eigenmodes, that we note $\left\{\lambda_{ka}, \vuka\right\}$ with $k$ running from $1$ to $N$.

To obtain the eigenmodes of $\mathbb{G}$ as a function of the $\vsa$ and $\vuka$, we construct the \emph{Kronecker} product of vectors $\vsa$ and $\vuka$, defining vectors $\vec{S}_{ka}$ of dimension $L\times N$: (\pierre{should we say tensor product? I had never heard of Kronecker product before... I'm fine with it, but people might be more familiar with tensor?})
\begin{equation}
	\begin{split}
		\vec{S}_{ka} &= \vuka \otimes \vsa \\
		&= [u_{ka}^1\cdot\vsa, \ldots, u_{ka}^N\cdot\vsa].
		% &= [u_{ka}(1)s_a(1), \ldots, u_{ka}(1)s_a(L); \ldots; u_{ka}(N)s_a(1), \ldots, u_{ka}(N)s_a(L)]
	\end{split}
	\label{eq:Sdef}
\end{equation}
The $i$th block vector of $\vec{S}_{ka}$ will be written as $\vec{S}_{ka}^i = u_{ka}^i\cdot\vsa$. 
We can now show that $\vec{S}_{ka}$ are eigenvectors of matrix $\mathbb{G}$ by considering the $i$th block vector of the product $\mathbb{G}\cdot\vec{S}_{ka}$:
\begin{equation}
	\begin{split}
		\left(\mathbb{G}\cdot\vec{S}_{ka}\right)^i &= \sum_{j=1}^N \mathbb{G}_{ij} u_{ka}^j\cdot\vsa \\
		&= \sum_{j=1}^N z(\rho_a, \Delta t_{ij}) u_{ka}^j\cdot \vsa \\
		&= (\mathbf{G}^a\cdot\vuka)^i \cdot \vsa\\
		&= \lambda_{ka}(u_{ka}^i\cdot\vsa) \\
		&= \lambda_{ka}\vec{S}_{ka}^i,
	\end{split}
\end{equation}
where we have first used the fact that $\vsa$ is an eigenvector of $\mathbb{G}_{ij}$, then the definition of $\mathbf{G}^a$, and finally the fact that $\vuka$ is an eigenvector of $\mathbf{G}^a$.
This shows that the eigenmodes of $\mathbb{G}$ are $\left\{ \lambda_{ka}, \vec{S}_{ka} \right\}$ with $1\leq k\leq N$ and $1\leq a \leq L$. 
Since $\mathbb{G}$ is the covariance matrix of a Gaussian distribution, we also obtain that the $\lambda_{ka}$ are strictly positive.

Equation~\ref{eq:likelihood_2} can now be rewritten using the eigen-decomposition of $\mathbb{G}$: 
\begin{equation}
	\begin{split}
		\mathcal{L}_{\mathcal{D}}(\mathbb{G}) &= -\frac 1 2 \sum_{k=1}^N\sum_{a=1}^L \log\lambda_{ka} - \frac 1 2 \sum_{k=1}^N\sum_{a=1}^L \lambda_{ka}^{-1}(\vec{X}\cdot \vec{S}_{ka})^2\\
		&= -\frac 1 2 \sum_{k,a} \left(\log\lambda_{ka} + \lambda_{ka}^{-1}	\left(\sum_{i=1}^N u_{ka}^i\vx_i\cdot\vsa\right)^2 \right).
	\end{split}
	\label{eq:likelihood_3}
\end{equation}
Equation~\ref{eq:likelihood_3} expresses the likelihood as a function of $\vuka$, $\lambda_{ka}$ and $\vsa$. 
However, the definition of matrix $\mathbf{G}^a$ given by equation~\ref{eq:subG_def} makes clear that its eigenmodes $\{\lambda_{ka}, \vuka \}$ depend only of the eigenvalues $\rho_a$ of $\iC$, on $\gamma$, as well as of the structure of the tree through the quantities $\Delta t_{ij}$, although this dependence cannot be analytically expressed in a simple manner. 
This means that the likelihood in equation~\ref{eq:likelihood_3} is in fact a function of $\{\rho_a, \vsa\}$, \emph{i.e.} the eigenmodes of $\iC$ and of the time scale parameter $\gamma$.

% subsection derivation_of_the_likelihood (end)

\subsection{Maximization of the likelihood} % (fold)
\label{sub:maximization_of_the_likelihood}

As stated at the beginning of this section, our main task is to find the equilibrium covariance matrix $\mathbf{C}$ that maximizes the likelihood of the data. 
We also need to find the optimal time scale $\gamma$. 
In equation~\ref{eq:likelihood_3}, the likelihood is expressed as a function of $\gamma$ and $\{\rho_a, \vsa\}$, \emph{i.e.} the eigenvalues and eigenvectors of $\iC$, either directly or through the quantities $\{\lambda_{ka}, \vuka \}$. 
We know attempt to maximize the likelihood with respect to the eigenmodes $\{\rho_a, \vsa\}$ and $\gamma$. \\
In order to perform this optimization, we will need to conduct optimization with the eigenvectors $\{\vsa\}$ as a variable. 
Since $\iC$ is a symmetric matrix, its eigenvectors form an orthogonal basis of the vector-space of dimension $L$. 
Thus, they are fully characterized by $L(L-1)/2$ scalar \emph{Eulerian angles} $\{\theta_{\alpha\beta}\}$ with $1\leq \alpha < \beta \leq L$~\cite{Raffenetti1970GEA}(\pierre{\cite{Shepard_param_OM} was cited as well, do we keep it?}). 
We will use these Eulerian angles to parametrize the $\{\vsa\}$. The $\{\theta_{\alpha\beta}\}$ along with the $L$ eigenvalues $\rho_a$ allow to fully parametrize the $L(L+1)/2$ distinct values of $\iC$. \\

We now set to computing the gradient of the likelihood $\mathcal{L}_{\mathcal{D}}(\mathbb{G})$ with respect to parameters $\{\rho_a, \theta_{\alpha\beta}\}$ and $\gamma$. 
To make explicit the dependence of eigenvalues and eigenvectors of matrices $\mathbf{G}^a$ with respect to these parameters, we introduce the notation $\vec{u}_k(\rho_a,\gamma) = \vuka$ and $\lambda_k(\rho_a,\gamma) = \lambda_{ka}$. 
Note that from the definition of $\mathbf{G}^a$ in equation~\ref{eq:subG_def}, its eigenvalues and vectors depend only on the eigenvalues of $\iC$ and not on its eigenvectors. 
In the same way, we will now write $\mathbf{G}(\rho_a,\gamma)$ instead of $\mathbf{G}^a$. \\
The gradient of the likelihood is obtained by differentiating equation~\ref{eq:likelihood_3} with respect to parameters of interest. 
This gives us  three equations for the gradient: 
% dL / d rho
\begin{equation}
	\begin{split}
		\frac{\partial \mathcal{L}}{\partial\rho_a} &= -\frac 1 2 \sum_{k=1}^N \left\{ \frac{\partial\lambda_k}{\partial\rho_a}\lambda_k^{-1} - \frac{\partial\lambda_k}{\partial\rho_a} \lambda_k^{-2}\left(\sum_{i=1}^N u_{k}^i\vx_i\cdot\vsa\right)^2 \right. \\
		&+ \left. 2\lambda_k^{-1} \left(\sum_{i=1}^N u_k^i\vx_i\cdot\vsa\right) \left(\sum_{i=1}^N \frac{\partial u_k^i}{\partial\rho_a} \vx_i\cdot\vsa\right) \right\}, 
	\end{split}
	\label{eq:gradlikelihood_1}
\end{equation}
% dL / d theta
\begin{equation}
	\frac{\partial \mathcal{L}}{\partial\theta_{\alpha\beta}} = \sum_{k=1}^N \lambda_k^{-1} \left(\sum_{i=1}^N u_k^i\vx_i\cdot\vsa\right) \left(\sum_{i=1}^N u_k^i \vx_i\cdot\frac{\partial \vsa}{\partial\theta_{\alpha\beta}}\right),
	\label{eq:gradlikelihood_2}
\end{equation}
and
% dL / d gamma
\begin{equation}
	\begin{split}
		\frac{\partial \mathcal{L}}{\partial\gamma} &= -\frac 1 2 \sum_{k=1}^N \left\{ \frac{\partial\lambda_k}{\partial\gamma}\lambda_k^{-1} - \frac{\partial\lambda_k}{\partial\gamma} \lambda_k^{-2}\left(\sum_{i=1}^N u_{k}^i\vx_i\cdot\vsa\right)^2 \right. \\
		&+ \left. 2\lambda_k^{-1} \left(\sum_{i=1}^N u_k^i\vx_i\cdot\vsa\right) \left(\sum_{i=1}^N \frac{\partial u_k^i}{\partial\gamma} \vx_i\cdot\vsa\right) \right\}, 
	\end{split}
	\label{eq:gradlikelihood_3}
\end{equation}
The derivatives of $\vec{u}_k(\rho,\gamma)$ and $\lambda_k(\rho,\gamma)$ with respect to $\rho$ can then be computed using the following equations~\cite{matrix_cook_book}: 
\begin{equation}
	\frac{\partial\lambda_i(\rho,\gamma)}{\partial\rho} = \vec{u}_k(\rho,\gamma)^T\frac{\partial\mathbf{G}(\rho,\gamma)}{\partial\rho}\vec{u}_k(\rho,\gamma)
	\label{eq:grad_eigval}
\end{equation}
and
\begin{equation}
	\frac{\partial \vec{u}_k(\rho,\gamma)}{\partial\rho} = \sum_{l\neq k} \left( \vec{u}_k(\rho,\gamma)^T\frac{\partial\mathbf{G}(\rho,\gamma)}{\partial\rho}\vec{u}_l(\rho,\gamma) \right) \left(\lambda_k(\rho,\gamma) - \lambda_l(\rho,\gamma)\right)^{-1} \vec{u}_l(\rho,\gamma).
	\label{eq:grad_eigvec}
\end{equation}
Equivalent equations can be written for their derivatives with respect to $\gamma$. 
\pierre{So what if we some eigenvalues cross?}.\\
The computation of the gradient of $\mathcal{L}$ for a given set of parameters $\{\rho_a, \theta_{\alpha\beta}\}$ then goes as follows. 
For each eigenvalue $\rho_a$, we compute and diagonalize matrix $\mathbf{G}(\rho_a)$ to obtain its eigenmodes $\vec{u}_k(\rho_a)$ and $\lambda_k(\rho_a)$. 
Using equations~\ref{eq:grad_eigval} and ~\ref{eq:grad_eigvec} and their equivalent form for $\gamma$, we also numerically compute their derivatives with respect to $\rho_a$ and $\gamma$. 
This gives us all the quantities to estimate the gradient of $\mathcal{L}$ with respect to $\rho_a$ using equation~\ref{eq:gradlikelihood_1}. \\

 This nonlinear optimization problem was addressed by implementing a quasi-Newton Method algorithms \cite{NLopt}. \pierre{More details about this would be great. }

% subsection maximization_of_the_likelihood (end)

\section{ Results}
\label{Results}
We base the test of methodology developed on a sampling-inference procedure defined by two steps: 
\begin{enumerate}
\item  Generation of simulated  data using   a non-homogeneous  tree in algorithm \ref{alg_1}  with initial parameters $\gamma_{init}$ and $\bm J_{init}$. The tree structure  used is  binary with $K=10$ levels and   branch length randomly selected in $(0,1)$ interval.
\item Inference procedure by maximizing Equation~\ref{eq:likelihood_3} finding the maximum likelihood estimators $\gamma_{max}$ and $\bm J_{max}$  . 
\end{enumerate}

As was explained in section \ref{evolutionary process} , phylogenetic bias present at leaves of the tree depend of the set of time branch lengths  $\Delta t_{ij}$, the mutation rate $1/\gamma$ and the potential $\bm J$. This mean that in our simulations given the tree structure and $\bm J=\bm J_{init}$  we could define a threshold value for $\gamma$ above which  sequences in the  alignment are considered  non-independent: $\gamma_c=\lambda_{max}  \Delta t_{max}$, where  $\lambda_{max}$ is the maximum eigenvalue of $\bm J_{init}$ and $\Delta t_{max}$ is the  higher path time between two leaves in the tree.   Tunning the mutation rate respect $1/\gamma_c$ allow to generate sequences with distinct levels of phylogenetic bias. We choose three different values of parameter $\gamma$ in order to simulate  different regimes: with weak  $\gamma_{init}=0.1\gamma_c$ ,intermediate $\gamma_{init}=0.3\gamma_c$ and  strong  $\gamma_{init}=0.5\gamma_c$  phylogenetic signal.

In order to limit the influence of a particular selection of the initial conditions in our simulations,  30 repetitions with different set of initial parameters were made.  Sequences with length $L=10$  was drawn in the root from a Gaussian distribution $P^0(\mathbf{ x})\sim N(0,\bm J_{init}^{-1})$. The different couplings matrices for each data generation was selected as a positive semidefinite matrix drawn from a normal distribution with mean $\mu_{J_{init}}=0.8$ and standard deviation $\sigma_{J_{init}}=0.2$. To make the matrix sparse we dilute the couplings with density $0.3$.

In figures \ref{results_gamma10_n10},  \ref{results_corr_map_J_10_n10}, \ref{results_corr_map_C_10_n10} we compare the maximum likelihood estimators with the empirical ones which were obtained  from the inversion of the empirical covariance matrix ($\bm J_0=\bm C^{-1}_0$) and from minimization of the expression \ref{gamma_inverse_eq} on the parameter $\gamma_0$.  As can be seen    our method is able of inferring covariance and  coupling matrices closer to the initial  ones for all regimes, and  differences are more clear if there is more phylogenetic contamination in  alignments.

On the other hand we wonder if our inference algorithm make a difference in the task of  reproduce the topology of the co-evolutionary network defined by Hamiltonian \ref{eq:equil}, identifying the strong pairs. As our original coupling matrix is sparse  we could explore the ability of maximum likelihood estimators to identify the strongest non-zero entries of the original coupling matrix $\bm J_{init}$.  Figure \ref{PPV_n10} show , for different levels of phylogeny contamination,  positive predictive values averaged over 30 samples. Comparison with the corresponding result from empirical estimators  shows that taking phylogeny into account, as is the case of our inference algorithm, also helps in this direction.

\begin{figure*}[!htb]
\includegraphics[keepaspectratio=true,width=1.0\textwidth]{Figures/gamma_panel_L10_M1024.png}
	\vspace{-1mm}
\caption{{ Results of inference of the  maximum likelihood ($\gamma_{max}$) and empirical ($\gamma_{0}$)  estimators of parameter $\gamma$. Each point correspond to a different sampling-inference process as result of simulations with different initial parameters: $\gamma_{init},\bm J_{init}$. The length of the sequence vectors was $L=10$ and the number of sequences  $M=1024$ as consequence of a binary non-homogeneous tree with $K=10$ levels. Three different regimes were explored incrementing from left to right the  phylogenetic bias present at leaves sequences where \textbf{A}: weak phylogenetic signal and   $\gamma_{init}=0.1*\gamma_c$, \textbf{B}: intermediate phylogenetic signal  and $\gamma_{init}=0.3*\gamma_c$, \textbf{C}: strong phylogenetic signal and $\gamma_{init}=0.5*\gamma_c$. }} \label{results_gamma10_n10}
\end{figure*}

\begin{figure*}[!htb]
	\includegraphics[keepaspectratio=true,width=1.0\textwidth]{Figures/couplings_panel_frob_dist_vs_lin_reg_slope_L10_M1024_same_range.png}
\vspace{-1mm}
\caption{{ Fobenious distance and  slope of linear regression between empirical ($\bm J_{0}$) / inferred couplings matrix ($\bm J_{max}$) and initial couplings matrix $\bm J_{init}$.  Each point correspond to a different sampling-inference process as result of simulations with different initial parameters: $\gamma_{init},\bm J_{init}$. The length of the sequence vectors was $L=10$ and the number of sequences  $M=1024$ as consequence of a binary non-homogeneous tree with $K=10$ levels. Three different regimes were explored incrementing from left to right the  phylogenetic bias present at leaves sequences where \textbf{A}: weak phylogenetic signal and   $\gamma_{init}=0.1*\gamma_c$, \textbf{B}: intermediate phylogenetic signal  and $\gamma_{init}=0.3*\gamma_c$, \textbf{C}: strong phylogenetic signal and $\gamma_{init}=0.5*\gamma_c$.}} 
\label{results_corr_map_J_10_n10}
\end{figure*}

\begin{figure*}[!htb]
	\includegraphics[keepaspectratio=true,width=1.0\textwidth]{Figures/covariance_panel_pearson_corr_vs_lin_reg_slope_L10_M1024_same_range.png}
	\vspace{-1mm}
\caption{{  Pearson correlation and  slope of linear regression between empirical ($\bm C_{0}$) /inferred ($\bm C_{max}$) covariance  matrix and to  initial covariance matrix $\bm C_{init}=\bm J^{-1}_{init}$.  Each point correspond to a different sampling-inference process as result of simulations with different initial parameters: $\gamma_{init},\bm J_{init}$. The length of the sequence vectors was $L=10$ and the number of sequences  $M=1024$ as consequence of a binary non-homogeneous tree with $K=10$ levels. Three different regimes were explored incrementing from left to right the  phylogenetic bias present at leaves sequences where \textbf{A}: weak phylogenetic signal and   $\gamma_{init}=0.1*\gamma_c$, \textbf{B}: intermediate phylogenetic signal  and $\gamma_{init}=0.3*\gamma_c$, \textbf{C}: strong phylogenetic signal and $\gamma_{init}=0.5*\gamma_c$.    \textbf{center}: $\gamma=\gamma_i$, \textbf{right} $\gamma=\gamma_s$.  }} 
\label{results_corr_map_C_10_n10}
\end{figure*}

\begin{figure*}[!htb]
	\includegraphics[keepaspectratio=true,width=1.0\textwidth]{Figures/PPV_panel_L10_M1024.png}
	\vspace{-1mm}
\caption{{  Positive predictive values (PPV) averaged over 30 samples,  for each of the three different regimes: \textbf{A}: weak phylogenetic signal and   $\gamma_{init}=0.1*\gamma_c$, \textbf{B}: intermediate phylogenetic signal  and $\gamma_{init}=0.3*\gamma_c$, \textbf{C}: strong phylogenetic signal and $\gamma_{init}=0.5*\gamma_c$.  Positive predictive values are found as a proportion of agreement in the strongest couplings of the matrix $\bm J_{init}$ . The length of the sequences was $L=10$ so that couplings matrix has $10^2$ entries of which on average only 16 off diagonals elements  are nonzero  given the sparse character of the initial matrices. }} .
\label{PPV_n10}
\end{figure*}

\section{Discussion}
\label{sec:discussion}

\textbf{Acknowledgments:}

\clearpage
\bibliography{bib_phylo}

\bibliographystyle{unsrt}

\newpage
\appendix
\setcounter{figure}{0}
\renewcommand{\figurename}{Figure S}
\setcounter{table}{0}
\renewcommand{\tablename}{Table S}

\section{Supplementary material} % (fold)
\label{sec:supplementary_material}

\subsection{Generating artificial data} % (fold)
\label{sub:generating_artificial_data}

	We are interested in the case where the described Ornstein-Uhlenbeck process takes place on a tree. 
	For example, if configurations $\vec{x}$ represent quantitative traits of some organisms, the tree can represent the genealogy or phylogeny of these organisms. 
	Therefore, we have to be able to simulate the OU process on a tree. 
	In practice, given a rooted tree such as the one shown in figure~\ref{fig:sample_tree} of the main text, we want to sample a configuration $\vec{x}$ for every node in such a way that equation~\ref{eq:OUjoint} holds for every pair of nodes, the time $\Delta t$ then being the branch length connecting them. \\
	We use a simple methodology to achieve this. 
	First, note that given an arbitrary configuration $\vec{x}_0$ and a time $\Delta t$, we can generate a new configuration $\vec{x}$ distributed according to the propagator in equation~\ref{eq:OUpropagator} by the transformation
	\begin{equation}
		\vx = \Lam^{\Delta t}\vx_0 + \Sig^{1/2}\vec{\eta}
		\label{eq:OUsample}
	\end{equation}
	where $\Lam$ and $\Sig$ are defined in equation~\ref{eq:def_lambda}, and $\vec{\eta}$ is a vector of uncorrelated variables with distributions $\mathcal{N}(0,1)$. 
	Moreover, if $\vec{x_0}$ is distributed according to equation~\ref{eq:eq_distribution}, then $\vx$ and $\vx_0$ are distributed according to equation~\ref{eq:OUjoint}. 
	Note that equation \ref{eq:OUsample} is quite different from the Langevin equation~\ref{eq:langevin}, even though they have similar forms. 
	Whereas the Langevin equation describes the motion of $\vx$ in the potential $\bm{J}$, eq.~\ref{eq:OUsample} directly samples from the OU process. \\
	Given an already sampled internal node in the tree, equation~\ref{eq:OUsample} allows to sample a configuration for each of its children. 
	To sample the whole tree, we first sample the root note $\vx_0$ from the equilibrium distribution~\ref{eq:eq_distribution}. 
	By recursive applications of~\ref{eq:OUsample}, we then simply work our way down the tree until all leaves are sampled.

 % subsection generating_artificial_data (end)

\subsection{Initializing parameters} % (fold)
\label{sub:initializing_parameters}

	\subsubsection{Eigenvalues and eigenvectors of $\mathbf{C}^{-1}$} % (fold)
	\label{ssub:eigenvalues_and_eigenvectors}

	The initial value that we take for the covariance matrix is the empirical one
	\begin{equation*}
		\mathbf{C}^{emp} = \frac{1}{N}\sum_{i=1}^{N}\vx_i \cdot\vx_i^{\,T}.
	\end{equation*}
	Its eigenmodes $\{\rho^0_a, \vsa^{\,0}\}$ determine the starting point of the optimization.

	[Insert here how we go from $\vsa^{\,0}$ to the corresponding set of Eulerian angles or to the corresponding skew symmetric matrix.]

	% From empirical covariance matrix $c^*=\frac{1}{N}\sum_{n=1}^{N}\mathbf{x}_n \mathbf{x}_n^T$ we could extract information  to  initialize the parameters in the optimization process.

	% The eigenvalues of empirical couplings $j^*=1/c^*$  will define $\rho^0_i$ ands its eigenvectors  an orthogonal matrix $\bm S^0$ which satisfy the equations systems:
	% \begin{equation}
	% S_{ij}(\theta^0)=S^0_{ij}
	% \label{recurrent_eq}
	% \end{equation}
	% to find the orthogonal parameters which reproduce the initial orthogonal matrix, equation \ref{recurrent_eq} must be inverted or equivalently the solution set $\bm \theta^0$ could be numerically obtained minimizing the function:

	% \begin{equation}
	% f(\bm \theta^0)=\sum_{i,j}\left[ S^0_{ij}-S_{ij}(\bm \theta^0)\right] ^2
	% \label{obtain_theta0}
	% \end{equation}

	% subsubsection eigenvalues_and_eigenvectors (end)

	\subsubsection{Time scale parameter $\gamma$}

	The optimization also requires that we initialize the time scale $\gamma$. 
	For this, we try to find the optimal $\gamma$ given the data $\mathbf{X}$, the tree, and the OU process defined by the empirical covariance matrix.

	The probability distribution $P$ for the configurations of two leaves $\vx_i$ and  $\vx_j$  separated by time $\Delta t_{ij}$ is given by equation~\ref{eq:OUjoint} of the main text. 
	With this distribution we can calculate analytically the average of the scalar product $\vx_i^{\,T}\vx_j$:
	\begin{align*}
		\langle \vx_i^{\,T}\vx_j \rangle_P &= \int \ddroit \vx_i \ddroit \vx_j P(\vx_i,\vx_j|\Delta t_{ij}) \sum_{a=1}^L   x_i^{a}x_j^{a}  \\
		&=\sum_{i=a}^L \langle  x_i^{a} x_j^a \rangle_{P}.\\
	\end{align*}
	The covariance $\langle  x_i^{a} x_j^a \rangle_{P}$ between observations separated by a time $\Delta t_{ij}$ is given by equation~\ref{eq:pairwisecov} of the main text.
	Using this, we now have
	\begin{align}
		\nonumber\langle \vx_i^{\,T}\vx_j \rangle_P &= \sum_{a=1}^L \left( \Lam^{\Delta t_{ij}} \mathbf{C} \right)_{aa}\\
		\nonumber&= \Tr\left( \Lam^{\Delta t_{ij}} \mathbf{C} \right)\\
		&= \sum_{a=1}^L \rho_a^{-1}e^{-\gamma\rho_a\Delta t_{ij}}.
		\label{eq:scalar_product}
	\end{align}
	Having initialized the covariance matrix $\mathbf{C}$ at its empirical value, we know the values of all members of the r.h.s. of equation~\ref{eq:scalar_product} except that of $\gamma$. 
	On the other hand, equivalent versions of equation~\ref{eq:scalar_product} can be written for all pairs of configurations $i$ and $j$. 
	To find an initial value of $\gamma$ which is consistent with the data and the empirical covariance matrix, we search for one that best explains the observed scalar products between configurations. 
	We thus define $\gamma^0$ to be the argument minimizing the functional $F(\gamma)$: 
	\begin{equation}
	 	F(\gamma) = \sum_{1\leq i<j\leq N}\left[ \vx_i^{\,T}\vx_j - \sum_{a=1}^L \rho_a^{-1}e^{-\gamma\rho_a\Delta t_{ij}} \right].
	\end{equation} 
	As $F$ depends on one scalar parameter, it is straightforward to minimize it, allowing us to initialize $\gamma$ to a reasonable value.

% subsection initializing_parameters (end)

\subsection{Joint probability of leaves is a gaussian} % (fold)
\label{sec:joint_probability_of_leaves_is_a_gaussian}

	A quick description of the idea. 
	Let's assume that we have a (sub-)tree $\mathcal{T}$ constituted of nodes $\mathbf{X}=(\vx_1,\ldots,\vx_N)$. $\mathbf{X}$ comprises all the leaves \emph{and} internal nodes. 
	We assume that $P(X)$ is a normal distribution, that is we can write it as 
	\begin{equation}
		P(\mathbf{X}) \propto \curlynormal{\mathbf{X}^{T}\mathbb{C}^{\mathcal{T}}\mathbf{X}},
		\label{eq:tree_gaussian}
	\end{equation}
	with $\mathbb{C}^{\mathcal{T}}$ being the corresponding correlation matrix. 
	Note that the superscript $\mathcal{T}$ implies that the correlation matrix depends on the shape of the tree .
	We will now ``graft'' another node $\vec{y}$ on $\mathcal{T}$ by joining it to some node $\vx_r\in \mathbf{X}$. \pierre{Need a small figure to sketch the idea. Will draw it later.}
	It is assumed that the evolution separating the configurations of $\vx_r$ and of $\vec{y}$ has taken place according to the Ornstein-Uhlenbeck (OU) process.
	If that is the case, the distribution of $\vec{y}$ conditional on $\vx_r$, $P(\vec{y}\vert\,\vx_r)$ is described by equation~\ref{eq:OUpropagator} of the main text. 
	We will now prove that the joint distribution $P(\vec{y},X)$ is a normal distribution. 
	If we achieve this, it follows that if the configurations of all nodes in a tree were generated by the OU process, then their joint distribution is a Gaussian. 
	Since node $\vec{y}$ attaches only to $\mathcal{T}$ only through node $\vx_r$, we can write
	\begin{equation*}
		P(\vec{y},X) = P(\vec{y}\vert\,X)P(X) = P(\vec{y}\vert \,\vx_r)P(X).
	\end{equation*}
	Combining equations~\ref{eq:OUpropagator} of the main text and \ref{eq:tree_gaussian}, we obtain
	\begin{equation}
		P(\vec{y},X) = \curlynormalpar{\vec{y}^{T}\iSig \vec{y} - 2\vx_r^{T}\Lam\iSig\vec{y} + \vx_r\Lam^2\iSig\vx_r + \mathbf{X}\mathbb{C}^{\mathcal{T}}\mathbf{X}}.
	\end{equation}
	If we now define the joint vector $\mathbf{X}'=(\vx_1,\ldots,\vx_N,\vec{y})$, with $\vec{y}$ having index $N+1$, it is clear that $\mathbf{X}'$ is normally distributed with a new correlation matrix $\mathbb{C}'$ having the following properties \pierre{the same if $i,j$ are different from $r$ or $N+1$, and new values if they correspond to $r$ or $N+1$}... \\
	Thus, for a tree on which configuration of nodes have been obtained by the OU process, the joint distribution of configurations of all nodes is a normal distribution. 
	This implies that the distribution of any subset of nodes is also a Gaussian, and thus that the joint distribution of \emph{leaves} is a Gaussian.

% section joint_probability_of_leaves_is_a_gaussian (end)

% section supplementary_material (end)

\end{document}

### 2021-01-25 main.tex
Intro draft

@@
 \usepackage{bbm}
+\usepackage{subfigure}
 
@@
 % Pierre
-\newcommand{\pierre}[1]{{\color{red}Pierre: #1}}
 \newcommand{\curlynormalpar}[1]{\exp\left\{-\frac{1}{2}\left( #1 \right)\right\}}
@@
 
+\newcommand{\sref}[1]{S\ref{#1}}
+\newcommand{\pierre}[1]{{\color{red}Pierre: #1}}
+\newcommand{\martin}[1]{{\color{blue}Martin: #1}}
+
 \begin{document}
 
-\title{Inference of global coevolutionary models from hierarchical correlated data.}
+\title{Global multivariate model learning from hierarchically correlated data}
 \date{}
@@
 \author{Edwin Rodriguez Horta} 
-\affiliation{Group of Complex Systems and Statistical Physics, Department of Theoretical Physics, University of Havana, Cuba}
-\affiliation{Laboratoire de Biologie Computationnelle et Quantitative (LCQB), Sorbonne Université, France}
+\affiliation{Sorbonne Universit\'{e}, CNRS, Institut de Biologie
+  Paris-Seine, Laboratoire de Biologie Computationnelle et
+  Quantitative -- LCQB, Paris, France}
+\affiliation{Group of Complex Systems and Statistical Physics, Department of Theoretical Physics, University of Havana, Havana, Cuba}
 %
 \author{Pierre Barrat-Charlaix} 
-\affiliation{Biozentrum, Universität Basel, Switzerland}
+\affiliation{Biozentrum, Universität Basel, Basel, Switzerland}
 %
 \author{Alejandro Lage} 
-\affiliation{Group of Complex Systems and Statistical Physics, Department of Theoretical Physics, University of Havana, Cuba}
+\affiliation{Group of Complex Systems and Statistical Physics, Department of Theoretical Physics, University of Havana, Havana, Cuba}
 %
@@
 \email{Correspondence to: Martin Weigt, \bf{martin.weigt@upmc.fr}}
-\affiliation{Laboratoire de Biologie Computationnelle et Quantitative, Sorbonne Université, France}
+\affiliation{Sorbonne Universit\'{e}, CNRS, Institut de Biologie
+  Paris-Seine, Laboratoire de Biologie Computationnelle et
+  Quantitative -- LCQB, Paris, France}
 
@@
 \label{sec:int}
+
+In the last decades, biology has seen impressive progress in experimental techniques which has resulted in a large increase of available data. 
+This is especially visible in the case of biological sequences,  with databases now harboring a vast amount of high-quality DNA or protein sequences. [UNIPROT/GENBANK]
+This has in turn fuelled a development of quantitative methods to model forces guiding the evolution of sequences or other biological traits [REFS?]. 
+A common idea in this context is that it is possible to use characteristics of homologous genes or organisms to construct quantitative models of the selection acting on them. 
+A successful example is the representation of protein sequences by probabilistic models in the so-called DCA method [REF].
+
+Models built in this way usually assume that biological characteristics of related sequences or organisms are evolving under the same constraints but are independent from each other in the statistical sense.
+This allows simple formulation for model learning. 
+However, phylogenetic relations between evolving entities imply that these cannot be considered independent from each other, and doing so can result in important biases when learning models. 
+For instance, it has been shown that phylogenetic relations between protein sequences induce non-trivial correlations that are not related to protein function [COLWELL paper].
+
+In order to correct for these biases, it is necessary to disentangle the statistical effects due to selection forces and those due to phylogeny. 
+However, this becomes a complex problem when dealing with complex models where traits under study are not independent from each other. 
+Accounting for these biases then requires one to understand how correlated traits evolve along a tree. 
+History of modeling this with BM or OU processes [Felsenstein, Hansen]. Also, field of PCM
+
+Increasing data --> new quantitative techniques in biology. 
+One big recent success: pairwise sequence models, blabla
+However, models learned from data with phylogenetic correlations can suffer from biases.
+To account for these biases in the case of complex models, it's necessary to understand how correlated traits evolve along a tree (or: quantities that are modeled and typically not independent evolve along a tree)
+
+There is a history of looking at multi parameter complex models and their relation with phylogeny. 
+in this context, it is often assumed that quantitative traits that evolve under selection can be modeled using an OU process. 
+This is usually done in the field of PCMs, with most common models being Brownian motion or OU processes.
+
+Here, we want to understand how to find parameters values of an OU gaussian process from data correlated through a tree. 
+This could be applied to the evolution of any correlated biological traits, but also potentially to protein sequences.
+
+The MS is divided as follows: >>>
+
+\paragraph{Old} % (fold)
+\label{par:old}
+
 Theory and applications of the inverse problem of statistical physics has had great development in recent years since the availability of large amounts of data on a microscopic scale produced by the continuous technological progress.
@@
 
+% paragraph old (end)
+
 \section{The multivariate Ornstein-Uhlenbeck process}
@@
 
-We consider a system characterized by $L$ continuous degrees of freedom and whose state is fully described by a vector $\vx$ of length $L$. 
+We consider a system characterized by $L$ continuous degrees of freedom and whose state is fully described by an $L$-dimensional vector $\vx\in\mathbb{R}^L$.
 These degrees of freedom can be continuous phenotypic traits of some living organism, or the sequence of a gene or a protein if a continuous approximation is made.
-At equilibrium, $\vx$ is normally distributed:
+At equilibrium, $\vx$ is assumed to be normally distributed,
 \begin{equation}
-	P_{eq}(\vx) = \frac{1}{Z(\bm{J})}\curlynormal{\vx^T\bm{J}\vx},
+	P_{eq}(\vx) = \frac{1}{Z(\bm{J})}\curlynormal{\vx^T\bm{J}\vx}\ ,
 	\label{eq:eq_distribution}
 \end{equation}
-where $\bm{J}$ is the symmetric positive definite \emph{coupling matrix} and $Z(\bm{J})$ is a normalization constant. 
-We are interested in inferring this coupling matrix from a given amount of observed states $\vx$ of the system.
-If these observations were independent from each other, $\bm{J}$ would simply be equal to the inverse of the \emph{covariance matrix} of the data, written $\bm{C}$.
+where $\bm{J}$ is the symmetric, positive definite \emph{coupling matrix} and $Z(\bm{J}) = \sqrt{(2\pi)^L/\det \bm{J}}$ is the normalization constant; the means of all components of $\vx$ are set to zero without loss of generality.
+We are interested in inferring the coupling matrix from a given amount of observed states $\vx$ of the system.
+If these observations were independent from each other, due to the simple Gaussian form of Eq.~(\ref{eq:eq_distribution}), $\bm{J}$ would simply be equal to the inverse of the empirical \emph{covariance matrix} of the data, written $\bm{C}=\bm{J}^{-1}$.
 
-However, we are concerned here with the case where these observations are not independent. 
-On the contrary, they are the result of a dynamical process which took place during a finite amount of time, and is thus not necessarily equilibrated. 
+However, here we are concerned with the case where observations are not independent. 
+On the contrary, they result from a dynamical process taking place during a finite amount of time. 
+Different datapoints are therefore correlated even if individually assumed to be at equilibrium. 
 This dynamical process is described below.
@@
 \end{equation}
-Here, $\vec{\xi}(t)$ is a vector of uncorrelated white noises, and $\gamma^{-1}$ is the characteristic timescale governing the dynamics. 
-In short, equation~\ref{eq:langevin} states that the system described by $\vx$ undergoes Brownian motion in an energy landscape characterized by the coupling matrix $\bm{J}$.
+Here, $\vec{\xi}(t)$ is a vector of uncorrelated white noise, and $\gamma^{-1}$ is the characteristic timescale governing the dynamics. 
+In short, Eq.~(\ref{eq:langevin}) states that the system described by $\vx$ undergoes Brownian motion in a quadratic energy landscape characterized by the coupling matrix $\bm{J}$.
 
-We are not interested in $\vx$ directly, but rather by its probability distribution $P(\vx,t\vert\,\vx_0)$, that is the probability to find the system in state $\vx$ knowing it was in state $\vx_0$ and evolved for time $t$.
-The Fokker-Planck equation corresponding to equation~\ref{eq:langevin} is straightforward to write: 
+We are not interested in $\vx$ directly, but rather in its probability distribution $P(\vx\vert\,\vx_0,\Delta t)$, \ie~in the probability to find the system in state $\vx$ knowing it was in state $\vx_0$ some time $\Delta t$ before.
+The Fokker-Planck equation corresponding to Eq.~(\ref{eq:langevin}) is straightforward to write, 
 \begin{equation}
-	\gamma \partial_t P = \left(-\sum_{a,b=1}^L\frac{\partial}{\partial x_a} J_{ab}x_b + \sum_{a=1}^L\frac{\partial^2}{\partial x_a^2}\right) P, 
+	\gamma^{-1} \partial_t P = \left(-\sum_{a,b=1}^L\frac{\partial}{\partial x_a} J_{ab}x_b + \sum_{a=1}^L\frac{\partial^2}{\partial x_a^2}\right) P, 
 	\label{eq:FP}
 \end{equation}
-where the parenthesized expression on the right hand side is understood as an operator on $P$. 
-The solution to~\ref{eq:FP} is a multivariate normal distribution~\cite{singh2017multiOU}:
+where the parenthesized expression on the right hand side is understood as an operator acting on $P$. 
+The solution to Eq.~(\ref{eq:FP}) is a multivariate normal distribution~\cite{singh2017multiOU}:
 \begin{equation}
-	P(\vx, \Delta t|\, \vx_0) = \left((2\pi)^N\det\iSig\right)^{-1/2}
+	P(\vx|\, \vx_0, \Delta t) = \left[(2\pi)^N\det\Sig\right]^{-1/2}
 	\curlynormal{(\vx-\vec{\mu})^T\iSig(\vx- \vec{\mu})},
@@
 \end{equation}
-where we introduce matrices $\Sig$ and $\Lam$ as well as vector $\vec{\mu}$: 
+where we introduce the matrices $\Sig$ and $\Lam$ as well as the vector $\vec{\mu}$ as 
 \begin{equation}
-	\Lam = e^{-\gamma\bm{J}}, \qquad \vec{\mu} = \Lam^{\Delta t} \vx_0, \qquad \Sig = \bm{J}^{-1}(\mathbb{1} - \Lam^{2\Delta t}).
+	\Lam = e^{-\gamma\bm{J}}, \qquad \vec{\mu} = \Lam^{\Delta t} \vx_0, \qquad \Sig = \bm{J}^{-1}(\mathbb{1} - \Lam^{2 \Delta t}).
 	\label{eq:def_lambda}
 \end{equation}
-Equations~\ref{eq:OUpropagator} and~\ref{eq:def_lambda} define a multivariate \emph{Ornstein-Uhlenbeck} process (OU).
+Eqs.~(\ref{eq:OUpropagator}) and~(\ref{eq:def_lambda}) define a multivariate \emph{Ornstein-Uhlenbeck} (OU) process.
 
-It is important to note that since matrix $\Lam$  is the exponential of $-\bm{J}$, it is symmetric, has strictly negative eigenvalues and commutes with $\bm{J}$. 
-We also underline the fact that $\Sig$ and $\vec{mu}$ depend on $\Delta t$, although this dependence is absent in our notation.
+Note that since matrix $\Lam$ is an exponential of $\bm{J}$, it is symmetric, has strictly positive eigenvalues and commutes with $\bm{J}$. 
+We also underline that $\Sig$ and $\vec{\mu}$ depend on $\Delta t$, although this dependence is not explicitly written in our notation to make it less heavy.
 
-By taking $\gamma\Delta t \gg 1$ and using the fact that $\bm{J}$ has strictly positive eigenvalues , one immediately recovers equation~\ref{eq:eq_distribution}. 
-This allows us to compute the joint distribution of two configurations $\vx_1$ and $\vx_2$ separated by a time $\Delta t$ by multiplying equations~\ref{eq:eq_distribution} and~\ref{eq:OUpropagator}: 
-\begin{equation}
-	P(\vx_1, \vx_2\vert\,\Delta t) \propto \curlynormalpar{\vx_1^T\iSig\vx_1 + \vx_2^T\iSig\vx_2 - 2\vx_1^T\Lam^{\Delta t}\iSig\vx_2}. 
+By taking $\gamma\Delta t \gg 1$ and using the fact that $\bm{J}$ has strictly positive eigenvalues, one immediately recovers Eq.~(\ref{eq:eq_distribution}). 
+We can thus compute the joint distribution of two configurations $\vx_1$ and $\vx_2$ separated by time difference $\Delta t$ by multiplying Eqs.~(\ref{eq:eq_distribution}) and~(\ref{eq:OUpropagator}),
+\begin{eqnarray}
+    P(\vx_1, \vx_2\vert\,\Delta t) &= & P(\vx_1 |\, \vx_2,\Delta t) \times P_{eq}(\vx_2)
+    \nonumber\\
+    &\propto& \curlynormalpar{\vx_1^T\iSig\vx_1 + \vx_2^T\iSig\vx_2 - 2\vx_1^T\Lam^{\Delta t}\iSig\vx_2}. 
 	\label{eq:OUjoint}	
-\end{equation}
-This last equation is important as it allows one to compute the joint covariance of configurations $\vx_1$ and $\vx_2$. 
-The probability distribution in~\ref{eq:OUjoint} is a normal distribution whose inverse covariance matrix is defined by blocks: $\Sig$ on the diagonal and $-\Lam^{\Delta t}\Sig$ off-diagonal. 
+\end{eqnarray}
+This equation illustrates the \emph{time reversibility} of the OU process. 
+Indeed, the distribution does not depend on which configuration $\vx_1$ or $\vx_2$ came first.
+
+Equation \eqref{eq:OUjoint} allows for computing the joint covariance of the correlated equilibrium configurations $\vx_1$ and $\vx_2$. 
+The probability distribution in Eq.~(\ref{eq:OUjoint}) is normal with an inverse covariance matrix defined by blocks: $\Sig$ on the diagonal and $-\Lam^{\Delta t}\Sig$ off-diagonal. 
 By inverting this block matrix, given that $\Lam$ and $\Sig$ commute and are invertible, one obtains the following covariance: 
@@
 \end{equation}
-As said above, the aim of this work is to infer matrix $\bm{J}$ from a set of observations $\vx$ that are generated this OU process. 
-More precisely, any two observations $\vx_1$ and $\vx_2$ will be separated by a time $\Delta t$ during which the OU process took place. 
-Equation~\ref{eq:pairwisecov} then allows us to distinguish two regimes. 
-Let us call $\{\rho_a\}$ the eigenvalues of $\bm{J}$. 
-Using the definition of $\Lam$, we immediatly have that the eigenvalues of matrix $\Lam^{\Delta t}\bm{C}$, \ie  the left hand side of equation~\ref{eq:pairwisecov}, are of the form $\rho_a^{-1}e^{-\gamma\rho_a\Delta t}$.
 
-Since the $\rho_a$ are positive, the eigenvalues of $\Lam^{\Delta t}\bm{C}$ vanish exponentially for large enough times. 
-The timescale of this exponential decrease is set by the characteristic time $\tau_c^{-1} = \gamma\rho_{min}$, where $\rho_{min}$ is the smallest eigenvalue of $\bm{J}$. 
+Eq.~(\ref{eq:pairwisecov}) allows us to readily distinguish two regimes. 
+Let us call $\rho_a$ the eigenvalues of $\bm{J}$. 
+The eigenvalues of $\Lam^{\Delta t}\bm{C}$ are then equal to $\rho_a^{-1}e^{-\gamma\rho_a\Delta t}$. 
+Since all $\rho_a$ are positive, the eigenvalues of $\Lam^{\Delta t}\bm{C}$ vanish exponentially over time.
+The slowest timescale of exponential decay is set by $\tau_c^{-1} = \gamma\rho_{min}$, with $\rho_{min}$ being the smallest eigenvalue of $\bm{J}$.
 Thus, for $\Delta t / \tau_c \gg 1$, $\vx_1$ and $\vx_2$ are uncorrelated. 
-If this is verified for all pairs of observations, the regime is that of \emph{weakly correlated} data, and the inference of $\bm{J}$ can simply be performed by inverting the empirical correlation matrix.
+If this is verified for all pairs of observations $\vx_i$ and $\vx_j$, the regime is that of \emph{uncorrelated} data -- the inference of $\bm{J}$ can simply be performed by inverting the empirical covariance matrix extracted from the data.
 Inversely, for $\Delta t / \tau_c \ll 1$, $\vx_1$ and $\vx_2$ are highly correlated, defining a \emph{strongly correlated} regime. 
-It should be noted that when $\Delta t = 0$, the joint correlation matrix of $\vx_1$ and $\vx_2$ becomes non invertible, and equation~\ref{eq:pairwisecov} becomes irrelevant.
+It should be noted that for $\Delta t = 0$, the joint correlation matrix of $\vx_1$ and $\vx_2$ becomes non invertible, and Eq.~(\ref{eq:pairwisecov}) becomes irrelevant.
+Actually, $\vx_1$ and $\vx_2$ coincide at that point, \ie~we have $P(\vx_1,\vx_2|\,\Delta t=0) = P_{eq}(\vx_1) \times \delta(\vx_1 - \vx_2)$ using the $L$-dimensional Dirac distribution.
 
-\section{Statement of the problem}
-\label{statement of the problem}
+\section{Methods}
+\label{sec:methods}
+
+\subsection{Statement of the problem}
+\label{sub:statement_of_the_problem}
 
 \begin{figure}[!htb]
-	\includegraphics[width=0.6\textwidth]{Figures/tree.pdf}
-	\caption{ \pierre{We should get a better figure here.}}
+	\includegraphics[width=0.8\textwidth]{Figures/tree.pdf}
+	\caption{Schematic representation of a tree $\mathcal{T}$ underlying the data generating process.
+The process starts at the root node $r$ with a configuration $\vx_r$ sampled from $P_{eq}(\vx_r)$.
+The dynamics consist in independent realizations of the OU process on all branches from ancestral nodes $a(n)$ to child nodes $n$ over times corresponding to the branch length $\Delta t_{n,a(n)}$, initialized in the ancestral configuration $\vx_{a(n)}$.
+The observable data only consist of configurations of the leaf nodes (grey circles in the figure), while configurations of ancestral nodes remain unknown.
+There are no restrictions on the topology of tree $\mathcal{T}$ and the length of the branches.}
 	\label{fig:sample_tree}
@@
 
-The central problem discussed here is the inference of the probability distribution describing samples that are hierarchically correlated by a tree. 
-Formally, we assume that the data consists of $N$ real valued vectors of length $L$, noted $\{\vx_i\}$ with $i\in\{1\ldots N\}$. 
-Taken individually, the $\vx_i$s are normally distributed according to equation~\ref{eq:eq_distribution}. with mean $0$.
-By construction, the equilibrium correlation between components of vectors $\vx$ is given by the inverse of the couplings, that is $\langle x^a x^b\rangle-\langle x^a\rangle\langle x^b\rangle = \bm C_{ab} = (\bm J^{-1})_{ab}$.
+The problem discussed here is the inference of the probability distribution describing samples that are hierarchically correlated by a tree, cf.~Fig.~\ref{fig:sample_tree}. 
+Formally, we assume that the data consists of $N$ real-valued vectors of length $L$, denoted $\{\vx_i\}\in\mathbb{R}^L$ with $i=1,...,N$. 
+Taken individually, the $\vx_i$ are distributed according to Eq.~\ref{eq:eq_distribution}, \ie~according to a multivariate Gaussian of zero mean and covariance $\bm C$.
+By construction, the equilibrium covariance between any pair of elements of a given vector $\vx=(x^1,...,x^L)$ is given by the inverse of the coupling matrix: $\langle x^a x^b\rangle-\langle x^a\rangle\langle x^b\rangle = \bm C_{ab} = (\bm J^{-1})_{ab}$ for all $a,b=1,...,L$.
 This implies that inferring the coupling matrix defining the probability distribution amounts to finding the \emph{equilibrium} covariance matrix $\mathbf{C}$.
@@
 However, this covariance cannot be directly measured as we consider observations that are not independently distributed.
-Instead, the set of measured configurations $\{\vx_i\}$ is the result of an Ornstein-Uhlenbeck (OU) process taking place on a tree $\mathcal{T}$, such as the one shown in figure~\ref{fig:sample_tree}.
-The configuration $\vx_i$ of each node $c$ in the tree is the result of the evolution given by the equation~\ref{eq:langevin}, with initial condition $\vx_{a(i)}$, $a(i)$ being the ancestral node of $i$, and running time $\Delta t_{ia(i)}$, that is the \emph{branch length} between $a(i)$ and $i$.
-This process is initialized with a root note $\vx_r$ that is distributed according to equation~\ref{eq:eq_distribution}.
-Observed data points are leaves of this tree, which is assumed to be perfectly known.
-The nature of the chosen evolution process implies that the distribution of a leaf configuration $\vx_i$ conditioned on the knowledge of another leaf configuration $\vx_j$ is given by equation~\ref{eq:OUpropagator}, where $\Delta t=\Delta t_{ij}$ represents the total branch length separating leaf $i$ and leaf $j$.
+Instead, the set of measured configurations $\{\vx_i\}_{i=1,...,N}$ is the result of an Ornstein-Uhlenbeck (OU) process taking place on a tree $\mathcal{T}$, as is illustrated in Fig.~\ref{fig:sample_tree}:
+\begin{itemize}
+    \item The process starts at the root node $r$ with a state vector $\vx_r$ drawn from the equilibrium distribution $P_{eq}$.
+    \item On each branch $(n,a(n))$ of length $\Delta t_{n,a(n)}$ connecting node $n$ with its ancestral node $a(n)$, the dynamics follow Eq.~(\ref{eq:langevin}), starting from initial condition $\vx_{a(n)}$, and running for time $\Delta t_{n,a(n)}$.
+In other words, given the state $\vx_{a(n)}$ of the ancestral node, $\vx_n$ is sampled from $P(\vx_n |\, \vx_{a(n)}, \Delta t_{n,a(n)})$, see Eq.~(\ref{eq:OUpropagator})
+    \item As a consequence, OU processes on branches stemming from common ancestral node evolve independently, but from an identical initial condition.
+    \item Observed data vectors correspond to the states of the leaves of the tree at the end of this process. The states of the internal nodes are not part of the observed data and remain unknown. 
+\end{itemize}
+This process is thought to represent the evolution of biologic traits along a phylogenetic tree, with the leave nodes corresponding to traits observed in today's species.
+Note that due to the reversible nature of our OU process, the joint probability of any pair of leaf configurations $\vx_i$ and $\vx_j$, with $i,j\in\{1,...,N\}$, is given by  $P(\vx_i,\vx_2|\,\Delta t_{ij})$ (Eq.~(\ref{eq:OUjoint}), with $\Delta t_{ij}$ denoting the total branch length of the path connecting $i$ and $j$ in the tree.
 
-The OU process is fully characterized by the quadratic potential $\bm J = \bm C^{-1}$ and a rate $\gamma$. 
-Hence, the configurations of the leaves, that is the data, are determined only by $\bm C$, $\gamma$, and the tree $\mathcal{T}$. 
-The aim of this work is to design a method to infer the most likely values of $\bm C$ and $\gamma$ from the knowledge of the data $\mathcal{D}=\{\vx_i\}$ and the shape of $\mathcal{T}$. 
-There are two notable extreme cases to this problem. 
-The first one is the $\Delta t_r/\tau_c\longrightarrow 0$ case, where $\Delta t_r$ is the typical branch length distance from leaves to root. 
-Leaf configurations are then almost identical to the root, \emph{i.e.} $\vx_i\simeq\vx_r$, and inference of $\bm C$ becomes impossible. 
-The second one is the $\Delta t_r/\tau_c\longrightarrow\infty$ case, where equilibration of the OU process takes place almost instantly when compared to branch lengths of $\mathcal{T}$.
-Leaf configurations are then independently drawn from the equilibrium distribution $P_{eq}$, and $\bm C$ can be readily estimated from the empirical covariances. 
-Here, we deal with the intermediary cases where the tree makes it impossible to simply estimate $\bm C$ from empirical covariances, but where $\tau_c$ is not large enough to remove all signatures of the energy potential.
+The OU process is characterized by the quadratic potential $\bm J = \bm C^{-1}$ and the rate $\gamma$. 
+Hence, the joint statistics of the leaf configurations $\{\vx_i\}_{i=1,...,N}$ (\ie~the data) is fully determined by $\bm C$, $\gamma$, and the tree $\mathcal{T}$. 
+The aim of this work is to derive a method for inferring the most likely values of $\bm C$ and $\gamma$ given the knowledge of the data $\mathcal{D}=\{\vx_i\}_{i=1,...,N}$ and the underlying tree $\mathcal{T}$.
+We consider here that both the topology and the branch lengths of $\mathcal{T}$ are known.
 
-We adopt a Bayesian approach to this inference problem by writing the probability of a given set of parameters as a function of the data using the following relation
+This problem shows two notable extreme cases: 
+The first one is the case where the typical branch length of the tree is short compared to the timescales of the OU process.
+As a consequence, leaf configurations are close to identical to the root, \ie~$\vx_i\simeq\vx_r$, and the inference of $\bm C$ becomes impossible. 
+The second one is the opposite case where the typical branch length of the tree is long compared to the longest timescale of the OU process $\tau_c$.
+In this case, the configuration of a child node is close to independent from that of its ancestor, and leaf configurations can be considered as independent samples from the equilibrium distribution $P_{eq}$.
+$\bm C$ can then be readily estimated by computing the empirical covariance matrix. 
+We are interested here in the intermediate regime where substantial tree-mediated correlations between data
+make it impossible to simply estimate $\bm C$ with the empirical covariance, but the depth of the tree introduces enough variability in the data for one to hope of reconstructing the energy potential $\bm{J}$.
+
+We adopt a Bayesian inference approach by writing the probability of a given set of parameters $\{\bm C, \gamma\}$ given the data $\{\mathcal{D},\mathcal{T}\}$ using Bayes' equation
 \begin{equation}
@@
 \end{equation}
-where the proportionality constant does not depend on the parameters $\bm C$ or $\gamma$, and where $P(\bm C, \gamma)$ can be any arbitrarily chosen prior distribution.
-The difficulty in equation~\ref{eq:bayesian_likelihood} lies in the estimation of the term $P(\mathcal{D}\vert\bm C, \gamma, \mathcal{T})$, that is the likelihood of the data given a set of parameters $\bm C$ and $\gamma$ and the tree $\mathcal{T}$.
-We detail the computation of this likelihood in the following section.
+with the proportionality constant not depending on the parameters $\{\bm C, \gamma\}$.
+Here, $P(\bm C, \gamma)$ can be any arbitrarily chosen prior distribution.
+The difficulty in Eq.~(\ref{eq:bayesian_likelihood}) lies in the estimation of the likelihood $P(\mathcal{D}\vert\bm C, \gamma, \mathcal{T})$,
+\ie~of the joint probability of the datapoints $\mathcal{D}=\{\vx_i\}_{i=1,...,N}$ for an OU process given by its parameters $\{\bm C, \gamma\}$ and the tree $\mathcal{T}$.
+We detail the computation of this probability in the following section.
 
@@
 
-The joint distribution of two configurations $\vx_1$ and $\vx_2$ separated by time $\Delta t$ is given by equation~\ref{eq:OUjoint} and corresponds to a joint normal distribution. 
-This means that the vector $\vec{X}=[\vx_1, \vx_2]$, \emph{i.e.} the concatenation of vectors $\vx_1$ and $\vx_2$, follows a normal distribution with null means.
-Of importance here is that this property of the OU process can be extended to the joint distribution of any subset nodes in a tree. 
-In other words, if we now define $\vec{X}=[\vx_1,\ldots,\vx_N]$ to be the concatenation of the configurations of all leaves of $\mathcal{T}$, we can write the distribution of $\vec{X}$ as
+The joint distribution of two configurations $\vx_1$ and $\vx_2$ separated by time $\Delta t$ is given by Eq.~(\ref{eq:OUjoint}) and corresponds to a joint normal distribution. 
+This means that the vector $\vec{X}=[\vx_1, \vx_2]$, \emph{i.e.} the concatenation of vectors $\vx_1$ and $\vx_2$, follows a normal distribution with zero mean and variance described above in Eqs.~(\ref{eq:def_lambda}).
+Of importance here is that this property of the OU process can be extended to the joint distribution of any subset of nodes in a tree. 
+In other words, if we now define $\vec{X}=[\vx_1,\ldots,\vx_N]$ to be the concatenation of all configurations in our dataset $\mathcal D$, we can write the distribution of $\vec{X}$ as
 \begin{equation}
@@
 \end{equation}
-where $\mathbb{G}$ is the \emph{joint correlation matrix} and depends on the tree as well as on $\bm{C}$ and $\gamma$.
-This property is shown in more detail in section \ref{sec:joint_probability_of_leaves_is_a_gaussian} of the Supplementary Material.\\
-Equation~\ref{eq:multiOU_jointall} allows us to compute the log-likelihood of the data $\vec{X}$ as a function of $\vec{X}$ itself and of the joint correlation matrix. 
-Indeed, taking its logarithm immediatly gives
+where $\mathbb{G}$ is the \emph{joint covariance matrix} and depends on the tree as well as on $\bm{C}$ and $\gamma$.
+
+The joint covariance matrix is a matrix of dimension $(L\cdot N)\times(L\cdot N)$, built by $N\times N$ blocks of size $L\times L$ with entries
 \begin{equation}
-	\mathcal{L}_{\mathcal{D}}(\mathbb{G}) = -\frac 1 2 \log \det  \mathbb{G}  -\frac 1 2 \vec{X}^T\mathbb{G}^{-1} \vec{X} + \text{cte}. 
-	\label{eq:likelihood_2}
-\end{equation}
-
-We will now use the fact that the joint distribution of the leaves is a Gaussian to compute $\mathbb{G}$ as a function of the tree and the parameters of the OU process.
-The joint correlation matrix is of dimensions $(L\cdot N)\times(L\cdot N)$ where $L$ is the length of individual Gaussian vectors $\vx_i$. 
-It describes the correlation of each leaf of the tree to every other leaf, in the sense that the $L\times L$ block $\mathbb{G}_{ij}$ with $1\leq i,j\leq N$ corresponds to the covariance of configurations $\vx_i$ and $\vx_j$.
-In other words, for indices $a$ and $b$ such that $1\leq a,b\leq L$, we have
-\begin{equation}
-	\mathbb{G}_{ij}(a,b) = \av{x_i^a x_j^b}. 
+	\mathbb{G}_{ij}(a,b) = \av{x_i^a x_j^b} - \av{x_i^a}\av{x_j^b},\ \ \ \ i,j\in\{1,...,N\}; a,b\in\{1,...,L\},
 	\label{eq:jointcov_1}
 \end{equation}
-Since $\vx_i$ and $\vx_j$ are leaves of the tree, they are separated by an evolution of the OU proces for a time $\Delta t_{ij}$.
-This means that without knowledge of the internal nodes, $\vx_i$ and $\vx_j$ are distributed according to equation~\ref{eq:OUjoint}. 
-In particular, this implies that their covariance matrix is given by equation \ref{eq:pairwisecov}.
-From this we can derive the value of the block $(i,j)$ of the joint correlation matrix, obtaining the following equation for $\mathbb{G}$: 
+where the (zero) marginals $\av{x_i^a}$ and $\av{x_j^b}$ are explicitly written for clarity.
+Each block $\mathbb{G}_{ij}$ is describing the connected correlations between two data vectors $\vx_i$ and $\vx_j$, which are separated by time $\Delta t_{ij}$, resulting as the sum of all branch lengths of the path connecting $i$ and $j$ on tree $\mathcal{T}$.
+Because the OU process is time reversible, we can directly apply Eq.~(\ref{eq:pairwisecov}) and give all blocks of $\mathbb{G}$ in closed form,
 \begin{equation}
@@
 	\begin{cases}
-		\bm C & \text{if $i=j$}.\\
-		\Lam^{\Delta t_{ij}}{\mathbf{C}} & \text{otherwise}  
-	\end{cases}.
+		\bm C & \text{if $i=j$}\\
+		\Lam^{\Delta t_{ij}}{\bm C} & \text{otherwise}  ,
+	\end{cases}
 	\label{eq:jointcov_2}
 \end{equation}
-We remind here that $\Lam = e^{-\gamma\iC}$ depends only on $\gamma$ and $\mathbf{C}$, and commutes with $\mathbf{C}$.  \\
+using the (currently unknown) covariance matrix $\bm C$ of a single equilibrium vector $\vx$.
+We remind here that $\Lam = e^{-\gamma\iC}$ depends only on $\gamma$ and $\bm C$, and commutes with $\bm C$.
+As a direct consequence, all blocks $\mathbb{G}_{ij}$ commute with each other and with $\bm C$.
 
-We can now use the structure of $\mathbb{G}$ described in equation~\ref{eq:jointcov_2} to re-write equation~\ref{eq:likelihood_2} in a way that we can compute. 
-More precisely, we will attempt to write the likelihood as a function of eigenmodes of $\iC$. \\
-Let us first note $\left\{\rho_a, \vsa\right\}$ the eigenvalues and eigenvectors of $\iC$, where the index $a$ runs from $1$ to $L$ and vectors $\vsa$ are of dimention $L$. 
+Eq.~(\ref{eq:multiOU_jointall}) allows us to compute the log-likelihood of the data $\vec{X}$ as a function of $\vec{X}$ itself and of the joint covariance matrix.
+Indeed, taking its logarithm immediately gives
+\begin{equation}
+	\mathcal{L}_{\mathcal{D}}(\mathbb{G}) = -\frac 1 2 \log \det  \mathbb{G}  -\frac 1 2 \vec{X}^T\mathbb{G}^{-1} \vec{X} + \text{const}\ ,
+	\label{eq:likelihood_2}
+\end{equation}
+but this expression is impractical for any numerical evaluation due to the large dimension of $\mathbb{G}$.
+However, the particular block structure of $\mathbb{G}$ described in Eq.~(\ref{eq:jointcov_2}) allows us to simplify the expression.
+To do so, we first introduce the eigenvalues and eigenvectors
+$\left\{\rho_a, \vsa\right\}$ of $\iC$, where the index $a$ runs from $1$ to $L$ and vectors $\vsa$ are of dimension $L$.
 By definition, we have $\rho_a>0$ for all $a$.
-The properties of the exponential of matrices imply that vectors $\vsa$ are also eigenvectors of the matrix $\Lam$. 
-Thus, we can easily write the eigenmodes of the block $\mathbb{G}_{ij}$ as $\left\{z(\rho_a, \Delta t_{ij}), \vsa\right\}$ where we introduced
+Using now Eq.~(\ref{eq:jointcov_2}), we immediately see that the vectors $\vsa$ are also eigenvectors of the individual blocks $\mathbb{G}_{ij}$ with eigenvalues $z(\rho_a, \Delta t_{ij})$ where we introduced
 \begin{equation}
-	z(\rho_a, \Delta t_{ij}) = \rho_a^{-1}e^{-\gamma\rho_a\Delta t_{ij}}.
+	z(\rho_a, \Delta t_{ij}) = \rho_a^{-1}e^{-\gamma\rho_a\Delta t_{ij}} \ .
 	\label{eq:z}
 \end{equation}
-By convention, $\Delta t_{ii} = 0$ and $z(\rho_a, \Delta t_{ii}) = \rho_a^{-1}$. \\
-We also introduce matrix $\mathbf{G}^a$ of dimensions $N\times N$ and defined for indices $1\leq i,j\leq N$ as 
+By convention, $\Delta t_{ii} = 0$ and the diagonal blocks are thus included via $z(\rho_a, \Delta t_{ii}) = \rho_a^{-1}$.
+
+As the next step, we introduce $N\times N$-dimensional matrices ${\bm G}^a, a=1,...,L,$ with elements 
 \begin{equation}
-	\mathbf{G}^a_{ij} = z(\rho_a, \Delta t_{ij}). 
+	{\bm G}^a_{ij} = z(\rho_a, \Delta t_{ij})\ ,\ \ \ \ 1\leq i,j\leq N\ .
 	\label{eq:subG_def}
 \end{equation}
-In other words, for a given index $1\leq a\leq L$, $\mathbf{G}^a$ is the matrix built by replacing blocks of $\mathbb{G}$ by their $a$th eigenvalue.
-Matrices $\mathbf{G}^a$ are symmetric have their own eigenmodes, that we note $\left\{\lambda_{ka}, \vuka\right\}$ with $k$ running from $1$ to $N$.
+In other words, for a given index $1\leq a\leq L$, ${\bm G}^a$ is the matrix built by replacing all blocks of $\mathbb{G}$ by their respective $a$th eigenvalue.
+Matrices ${\bm G}^a$ are symmetric and have their own eigenmodes, that we denote by $\left\{\lambda_{ka}, \vuka\right\}_{k=1,...,N}$.
 
-To obtain the eigenmodes of $\mathbb{G}$ as a function of the $\vsa$ and $\vuka$, we construct the \emph{Kronecker} product of vectors $\vsa$ and $\vuka$, defining vectors $\vec{S}_{ka}$ of dimension $L\times N$: (\pierre{should we say tensor product?
-I had never heard of Kronecker product before...
-I'm fine with it, but people might be more familiar with tensor?})
+To obtain the eigenmodes of the joint covariance matrix $\mathbb{G}$ as a function of the $\vsa$ and $\vuka$, we construct the direct product of vectors $\vsa$ and $\vuka$, defining vectors $\vec{S}_{ka}$ of dimension $L\times N$:
 \begin{equation}
@@
 \end{equation}
-The $i$th block vector of $\vec{S}_{ka}$ will be written as $\vec{S}_{ka}^i = u_{ka}^i\cdot\vsa$. 
+The $i$th block vector of $\vec{S}_{ka}$ will thus be written as $\vec{S}_{ka}^i = u_{ka}^i\cdot\vsa$. 
 We can now show that $\vec{S}_{ka}$ are eigenvectors of matrix $\mathbb{G}$ by considering the $i$th block vector of the product $\mathbb{G}\cdot\vec{S}_{ka}$:
@@
 		&= \lambda_{ka}(u_{ka}^i\cdot\vsa) \\
-		&= \lambda_{ka}\vec{S}_{ka}^i,
+		&= \lambda_{ka}\vec{S}_{ka}^i \ .
 	\end{split}
 \end{equation}
-where we have first used the fact that $\vsa$ is an eigenvector of $\mathbb{G}_{ij}$, then the definition of $\mathbf{G}^a$, and finally the fact that $\vuka$ is an eigenvector of $\mathbf{G}^a$.
-This shows that the eigenmodes of $\mathbb{G}$ are $\left\{ \lambda_{ka}, \vec{S}_{ka} \right\}$ with $1\leq k\leq N$ and $1\leq a \leq L$. 
-Since $\mathbb{G}$ is the covariance matrix of a Gaussian distribution, we also obtain that the $\lambda_{ka}$ are strictly positive.
+We have first used the fact that $\vsa$ is an eigenvector of $\mathbb{G}_{ij}$, then the definition of ${\bm G}^a$, and finally the fact that $\vuka$ is an eigenvector of ${\bm G}^a$. 
+This demonstrates that the eigenmodes of $\mathbb{G}$ are $\left\{ \lambda_{ka}, \vec{S}_{ka} \right\}$ with $1\leq k\leq N$ and $1\leq a \leq L$. 
+Since $\mathbb{G}$ is the covariance matrix of a Gaussian distribution, we conclude the $\lambda_{ka}$ to be strictly positive.
 
-Equation~\ref{eq:likelihood_2} can now be rewritten using the eigen-decomposition of $\mathbb{G}$: 
+Note that this decomposition of the eigenvectors leads to a drastic decrease in computational complexity for diagonalizing $\mathbb{G}$ (at given $\bm C$, $\gamma$ and $\cal T$), and in consequence also for calculating the likelihood according to Eq.~(\ref{eq:likelihood_2}), which depends on the inverse covariance matrix $\mathbb{G}^{-1}$.
+Matrix $\mathbb{G}$ has linear dimension $LN$, so the numerical diagonalization or inversion takes time ${\cal O}( (L N)^3)$.
+This is hardly achievable for systems of realistic length $L$ of the state vector, and sufficient number $N$ of data points for model learning.
+Following the above description, we need to first diagonlize $\bm C^{-1}$ (or equivalently $\bm C$), which requires time of ${\cal O}(L^3)$, followed by inversion of the $L$ matrices ${\bm G}^a$, each one having linear dimension $N$.
+The total time complexity therefore results in ${\cal O}(L^3) + {\cal O}(L\cdot N^3)$, and the calculation can be easily achieved even on a standard PC.
+This observation is essential for inference, since we need to redo this calculation for many realizations of $\bm C$ and $\gamma$, in order to find the ones maximizing the likelihood given the data $\cal D$ and the tree $\cal T$.
+\martin{As is shown in App.
+XXX, this calculation simplifies even more when considering a fully balanced and homogeneous tree.
+In this case, the matrices $\bm G^a$ commute and can be diagonalized simultaneously and analytically for any value of $\rho^a$.
+
+For the case of arbitrary trees, }Eq.~(\ref{eq:likelihood_2}) can now be rewritten using the eigendecomposition of $\mathbb{G}$: 
 \begin{equation}
@@
 \end{equation}
-Equation~\ref{eq:likelihood_3} expresses the likelihood as a function of $\vuka$, $\lambda_{ka}$ and $\vsa$. 
-However, the definition of matrix $\mathbf{G}^a$ given by equation~\ref{eq:subG_def} makes clear that its eigenmodes $\{\lambda_{ka}, \vuka \}$ depend only of the eigenvalues $\rho_a$ of $\iC$, on $\gamma$, as well as of the structure of the tree through the quantities $\Delta t_{ij}$, although this dependence cannot be analytically expressed in a simple manner.
+Eq.~(\ref{eq:likelihood_3}) expresses the likelihood as a function of $\vuka$, $\lambda_{ka}$ (resulting from the tree $\cal T$ and given $\rho^a$) and $\vsa$ (resulting from $\bm C$). 
+However, the definition of $\bm G^a$ Eq.~(\ref{eq:subG_def}) makes clear that its eigenmodes $\{\lambda_{ka}, \vuka \}$ depend only of the eigenvalues $\rho_a$ of $\iC$, on $\gamma$, as well as of the structure of the tree through the quantities $\Delta t_{ij}$, although this dependence cannot be analytically expressed in a simple manner.
 This means that the likelihood in equation~\ref{eq:likelihood_3} is in fact a function of $\{\rho_a, \vsa\}$, \emph{i.e.} the eigenmodes of $\iC$ and of the time scale parameter $\gamma$.
@@
 
-\subsection{Maximization of the likelihood} % (fold)
+\subsection{Maximizing the likelihood} % (fold)
 \label{sub:maximization_of_the_likelihood}
 
-As stated at the beginning of this section, our main task is to find the equilibrium covariance matrix $\mathbf{C}$ that maximizes the likelihood of the data. 
+As stated at the beginning of this section, our main task is to find the equilibrium covariance matrix $\bm C$ that maximizes the likelihood of the data. 
 We also need to find the optimal time scale $\gamma$. 
-In equation~\ref{eq:likelihood_3}, the likelihood is expressed as a function of $\gamma$ and $\{\rho_a, \vsa\}$, \emph{i.e.} the eigenvalues and eigenvectors of $\iC$, either directly or through the quantities $\{\lambda_{ka}, \vuka \}$.
-We know attempt to maximize the likelihood with respect to the eigenmodes $\{\rho_a, \vsa\}$ and $\gamma$. \\
-In order to perform this optimization, we will need to conduct optimization with the eigenvectors $\{\vsa\}$ as a variable. 
-Since $\iC$ is a symmetric matrix, its eigenvectors form an orthogonal basis of the vector-space of dimension $L$. 
-Thus, they are fully characterized by $L(L-1)/2$ scalar \emph{Eulerian angles} $\{\theta_{\alpha\beta}\}$ with $1\leq \alpha < \beta \leq L$~\cite{Raffenetti1970GEA}(\pierre{\cite{Shepard_param_OM} was cited as well, do we keep it?}).
-We will use these Eulerian angles to parametrize the $\{\vsa\}$.
-The $\{\theta_{\alpha\beta}\}$ along with the $L$ eigenvalues $\rho_a$ allow to fully parametrize the $L(L+1)/2$ distinct values of $\iC$.
-\\
+In Eq.~(\ref{eq:likelihood_3}), the likelihood is expressed as a function of $\gamma$ and $\{\rho_a, \vsa\}$, \ie~the eigenvalues and eigenvectors of $\iC$, either directly or through the quantities $\{\lambda_{ka}, \vuka \}$.
+We know attempt to maximize the likelihood with respect to the eigenmodes $\{\rho_a, \vsa\}$ and to the time scale $\gamma$.
 
-We now set to computing the gradient of the likelihood $\mathcal{L}_{\mathcal{D}}(\mathbb{G})$ with respect to parameters $\{\rho_a, \theta_{\alpha\beta}\}$ and $\gamma$. 
-To make explicit the dependence of eigenvalues and eigenvectors of matrices $\mathbf{G}^a$ with respect to these parameters, we introduce the notation $\vec{u}_k(\rho_a,\gamma) = \vuka$ and $\lambda_k(\rho_a,\gamma) = \lambda_{ka}$.
-Note that from the definition of $\mathbf{G}^a$ in equation~\ref{eq:subG_def}, its eigenvalues and vectors depend only on the eigenvalues of $\iC$ and not on its eigenvectors. 
-In the same way, we will now write $\mathbf{G}(\rho_a,\gamma)$ instead of $\mathbf{G}^a$. \\
-The gradient of the likelihood is obtained by differentiating equation~\ref{eq:likelihood_3} with respect to parameters of interest. 
-This gives us  three equations for the gradient: 
+In order to perform this optimization, we need to compute the gradient of the likelihood with repsect to the eigenvectors $\{\vsa\}$. 
+Since $\iC$ is a symmetric matrix, its eigenvectors form an orthogonal basis of the vector-space of dimension $L$ and their components cannot be changed independently.  
+One possible parametrization for the $\{\vsa\}$ consists in using $L(L-1)/2$ scalar \emph{Eulerian angles} $\{\theta_{\alpha\beta}\}$ with $1\leq \alpha < \beta \leq L$~\cite{Raffenetti1970GEA, Shepard_param_OM}.
+With the $L$ eigenvalues $\rho_a$, this results in $L(L+1)/2$ independent values that fully parametrize the $L(L+1)/2$ values of $\iC$. 
+A second possibility, that we have found faster in practice, is to express the matrix of the $\{\vsa\}$ as the exponential of a skew-symetric matrix, see section \ref{sub:parametrizations_of_eigenvectors} of the SM.
+However, this parametrization does not allow a simple analytical expression of the gradient of the likelihood, and we use it along with automatic differentiation.
+\pierre{Edwin, can you add a reference to the package you used in the .bib and here?}
+For this reason, we use the Eulerian angles below to express the gradient of the likelihood.
+
+As a first step, we need to compute the gradient of the likelihood $\mathcal{L}_{\mathcal{D}}(\mathbb{G})$ with respect to all parameters $\{\rho_a, \theta_{\alpha\beta}\}$ and $\gamma$. 
+To make explicit the dependences of eigenvalues and eigenvectors of the matrices $\bm G^a$ on these parameters, we introduce the notation $\vec{u}_k(\rho_a,\gamma) = \vuka$ and $\lambda_k(\rho_a,\gamma) = \lambda_{ka}$.
+Note that from the definition of $\bm G^a$ in Eq.~(\ref{eq:subG_def}), its eigenvalues and vectors depend only on the eigenvalues of $\iC$ and not on its eigenvectors. 
+In the same way, we will now write $\bm G(\rho_a,\gamma)$ instead of $\bm G^a$.
+
+The gradient of the likelihood is obtained by differentiating Eq.~(\ref{eq:likelihood_3}) with respect to the parameters of interest. 
+This gives us  three equations: 
 % dL / d rho
@@
 \begin{equation}
-	\frac{\partial\lambda_i(\rho,\gamma)}{\partial\rho} = \vec{u}_k(\rho,\gamma)^T\frac{\partial\mathbf{G}(\rho,\gamma)}{\partial\rho}\vec{u}_k(\rho,\gamma)
+	\frac{\partial\lambda_i(\rho,\gamma)}{\partial\rho} = \vec{u}_k(\rho,\gamma)^T\frac{\partial\bm{G}(\rho,\gamma)}{\partial\rho}\vec{u}_k(\rho,\gamma)
 	\label{eq:grad_eigval}
@@
 \begin{equation}
-	\frac{\partial \vec{u}_k(\rho,\gamma)}{\partial\rho} = \sum_{l\neq k} \left( \vec{u}_k(\rho,\gamma)^T\frac{\partial\mathbf{G}(\rho,\gamma)}{\partial\rho}\vec{u}_l(\rho,\gamma) \right) \left(\lambda_k(\rho,\gamma) - \lambda_l(\rho,\gamma)\right)^{-1} \vec{u}_l(\rho,\gamma).
+	\frac{\partial \vec{u}_k(\rho,\gamma)}{\partial\rho} = \sum_{l\neq k} \left( \vec{u}_k(\rho,\gamma)^T\frac{\partial\bm{G}(\rho,\gamma)}{\partial\rho}\vec{u}_l(\rho,\gamma) \right) \left(\lambda_k(\rho,\gamma) - \lambda_l(\rho,\gamma)\right)^{-1} \vec{u}_l(\rho,\gamma).
 	\label{eq:grad_eigvec}
@@
 Equivalent equations can be written for their derivatives with respect to $\gamma$.
-\pierre{So what if we some eigenvalues cross?}.\\
+
 The computation of the gradient of $\mathcal{L}$ for a given set of parameters $\{\rho_a, \theta_{\alpha\beta}\}$ then goes as follows. 
-For each eigenvalue $\rho_a$, we compute and diagonalize matrix $\mathbf{G}(\rho_a)$ to obtain its eigenmodes $\vec{u}_k(\rho_a)$ and $\lambda_k(\rho_a)$. 
+For each eigenvalue $\rho_a$, we compute and diagonalize matrix $\bm{G}(\rho_a)$ to obtain its eigenmodes $\vec{u}_k(\rho_a)$ and $\lambda_k(\rho_a)$. 
 Using equations~\ref{eq:grad_eigval} and ~\ref{eq:grad_eigvec} and their equivalent form for $\gamma$, we also numerically compute their derivatives with respect to $\rho_a$ and $\gamma$. 
-This gives us all the quantities to estimate the gradient of $\mathcal{L}$ with respect to $\rho_a$ using equation~\ref{eq:gradlikelihood_1}. \\
+This gives us all the quantities to estimate the gradient of $\mathcal{L}$ with respect to $\rho_a$ using equation~\ref{eq:gradlikelihood_1}.
 
- This nonlinear optimization problem was addressed by implementing a quasi-Newton Method algorithms \cite{NLopt}. \pierre{More details about this would be great. }
+ The optimization is performed by a quasi-Newton method \cite{NLopt}. Details are presented in section \ref{sub:optimization_scheme} of the SM.
 
@@
 
-\section{ Results}
-\label{Results}
-We base the test of methodology developed on a sampling-inference procedure defined by two steps: 
-\begin{enumerate}
-\item  Generation of simulated  data using   a non-homogeneous  tree in algorithm \ref{alg_1}  with initial parameters $\gamma_{init}$ and $\bm J_{init}$.
-The tree structure  used is  binary with $K=10$ levels and   branch length randomly selected in $(0,1)$ interval.
-\item Inference procedure by maximizing Equation~\ref{eq:likelihood_3} finding the maximum likelihood estimators $\gamma_{max}$ and $\bm J_{max}$  . 
-\end{enumerate}
+\section{Results}
+\label{sec:Results}
+% \subsection{Choosing an appropriate time scale interval}
 
-As was explained in section \ref{evolutionary process} , phylogenetic bias present at leaves of the tree depend of the set of time branch lengths  $\Delta t_{ij}$, the mutation rate $1/\gamma$ and the potential $\bm J$.
-This mean that in our simulations given the tree structure and $\bm J=\bm J_{init}$  we could define a threshold value for $\gamma$ above which  sequences in the  alignment are considered  non-independent: $\gamma_c=\lambda_{max}  \Delta t_{max}$, where  $\lambda_{max}$ is the maximum eigenvalue of $\bm J_{init}$ and $\Delta t_{max}$ is the  higher path time between two leaves in the tree.
-Tunning the mutation rate respect $1/\gamma_c$ allow to generate sequences with distinct levels of phylogenetic bias.
-We choose three different values of parameter $\gamma$ in order to simulate  different regimes: with weak  $\gamma_{init}=0.1\gamma_c$ ,intermediate $\gamma_{init}=0.3\gamma_c$ and  strong  $\gamma_{init}=0.5\gamma_c$  phylogenetic signal.
+In order to evaluate our inference procedure, we generate artificial data corresponding to the process described in section \ref{sub:statement_of_the_problem}. 
+We first build a  balanced binary tree $\mathcal{T}$ with $2^9=512$ leaves. 
+The length of each branch of $\mathcal{T}$ is chosen from a uniform distribution in the interval $[0,1]$. 
+We also sample positive semi-definite coupling matrix $\bm{J}$ of size $L\times L$ with $L=4$ or $L=10$, with entries normally distributed with mean $\mu_J=0.8$ and $\sigma_J=0.2$. 
+In order to perform interaction prediction, we randomly set to $0$ off-diagonal elements of $J$ with probability $0.7$, resulting in a coupling matrix with approximative density $0.3$.  
+Null elements of $J$ correspond to variables that do not interact, or to non-contacts in the case of an application to protein sequences.
 
-In order to limit the influence of a particular selection of the initial conditions in our simulations,  30 repetitions with different set of initial parameters were made.
-Sequences with length $L=10$  was drawn in the root from a Gaussian distribution $P^0(\mathbf{ x})\sim N(0,\bm J_{init}^{-1})$.
-The different couplings matrices for each data generation was selected as a positive semidefinite matrix drawn from a normal distribution with mean $\mu_{J_{init}}=0.8$ and standard deviation $\sigma_{J_{init}}=0.2$.
-To make the matrix sparse we dilute the couplings with density $0.3$.
+In order to investigate the different regimes of tree-induced correlation, we vary the parameter $\gamma$ around a reference timescale $\gamma_d$ defined as follows: 
+\begin{equation}
+    \gamma_d = \frac{1}{\Delta t_{av}\rho_{min}} 
+\end{equation}
+where $\Delta t_{av}$ is the average branch length separating two leaves of $\mathcal{T}$. 
+For $\gamma \gg \gamma_d$, leaf configurations are on average well decorrelated, whereas for $\gamma \ll \gamma_d$ all leaves will be strongly correlated. 
+By simulating data using different $\gamma$ in the range $[10^{-2},2]\cdot\gamma_d$, we investigate all relevant temporal regimes.  
+For each value of $\gamma$, we then sample  configurations of leaves of $\mathcal{T}$ using the process described in section \ref{sub:generating_artificial_data} of the supplementary material. 
+To avoid statistical noise when assessing the quality of our inference, we repeat the sampling of leaf configurations 100 times for each value of $\gamma$.
 
-In figures \ref{results_gamma10_n10},  \ref{results_corr_map_J_10_n10}, \ref{results_corr_map_C_10_n10} we compare the maximum likelihood estimators with the empirical ones which were obtained  from the inversion of the empirical covariance matrix ($\bm J_0=\bm C^{-1}_0$) and from minimization of the expression \ref{gamma_inverse_eq} on the parameter $\gamma_0$.
-As can be seen    our method is able of inferring covariance and  coupling matrices closer to the initial  ones for all regimes, and  differences are more clear if there is more phylogenetic contamination in  alignments.
+For each repetition of the sampling process, we perform our maximum likelihood procedure and obtain an inferred covariance matrix $\bm{C}_{max}$. 
+As a means of comparison, we also compute the empirical covariance matrix $\bm{C}_{emp}$ as if leaf configurations were independent. 
+Figure \ref{fig:pears_L4} shows the Pearson correlation between the real covariance matrix $\bm{C}=\bm{J}^{-1}$ and the empirical or inferred ones in the $L=4$ case (similar figures for $L=10$ are in section \ref{sub:supplementary_figures}).
+As expected, both methods perform well in the large $\gamma$ limit with a correlation close to $1$, and worse in the low $\gamma$ limit. 
+In this latter case, correlations due to phylogeny are too strong for our maximum likelihood method to pick up signal, and both methods perform as poorly. 
+However, there exists an intermediate regime where $\bm{C}_{max}$ is much closer to the actual correlation than  $\bm{C}_{emp}$. 
+In figure \ref{fig:error_1_L4}, we plot the relative $l2$-error between either covariance matrices in the left panel or coupling matrices in the right panel. 
+In both cases, our maximum-likelihood method results in a consistent improvement over the empirical estimator. 
+However, the relative error still reaches high values in the low $\gamma$ regime, which is likely due to $\bm{C}_{max}$ and $\bm{C}_{emp}$ being close to singular in this case.
 
-On the other hand we wonder if our inference algorithm make a difference in the task of  reproduce the topology of the co-evolutionary network defined by Hamiltonian \ref{eq:equil}, identifying the strong pairs.
-As our original coupling matrix is sparse  we could explore the ability of maximum likelihood estimators to identify the strongest non-zero entries of the original coupling matrix $\bm J_{init}$.
-Figure \ref{PPV_n10} show , for different levels of phylogeny contamination,  positive predictive values averaged over 30 samples.
-Comparison with the corresponding result from empirical estimators  shows that taking phylogeny into account, as is the case of our inference algorithm, also helps in this direction.
+\begin{figure}[!htb]
+	\centering
+	\includegraphics[keepaspectratio=true,width=0.45\textwidth]{Figures/corr_C_L4_balanced_tree_100_rep.pdf}
+	\caption{ Pearson correlation between empirical /maximum-likelihood   covariance matrices and the true covariance matrix.
+The inset plot represents the ratio between the Pearson correlation for the  maximum-likelihood covariance matrix and the one for the empirical covariance matrix.
+Simulations are performed for a tree of $512$ leaves  and  system size $L=4$.}
+	\label{fig:pears_L4}
+\end{figure}
 
 \begin{figure*}[!htb]
-\includegraphics[keepaspectratio=true,width=1.0\textwidth]{Figures/gamma_panel_L10_M1024.png}
-	\vspace{-1mm}
-\caption{{ Results of inference of the  maximum likelihood ($\gamma_{max}$) and empirical ($\gamma_{0}$)  estimators of parameter $\gamma$.
-Each point correspond to a different sampling-inference process as result of simulations with different initial parameters: $\gamma_{init},\bm J_{init}$.
-The length of the sequence vectors was $L=10$ and the number of sequences  $M=1024$ as consequence of a binary non-homogeneous tree with $K=10$ levels.
-Three different regimes were explored incrementing from left to right the  phylogenetic bias present at leaves sequences where \textbf{A}: weak phylogenetic signal and   $\gamma_{init}=0.1*\gamma_c$, \textbf{B}: intermediate phylogenetic signal  and $\gamma_{init}=0.3*\gamma_c$, \textbf{C}: strong phylogenetic signal and $\gamma_{init}=0.5*\gamma_c$.
-}} \label{results_gamma10_n10}
+		\begin{subfigure}{}
+			\centering\includegraphics[keepaspectratio=true,width=0.45\textwidth]{Figures/epsilon_error_C_L4_balanced_tree_100.pdf}
+		\end{subfigure}
+		\hspace{1mm}
+		\begin{subfigure}{}
+			\centering\includegraphics[keepaspectratio=true,width=0.45\textwidth]{Figures/epsilon_error_J_L4_balanced_tree_100.pdf}
+		\end{subfigure}
+	\caption{\textbf{Left:}Relative $l2$-error between empirical or maximum-likelihood  covariance matrices and the true covariance matrix.
+\textbf{Right:}Relative $l2$-error between empirical /maximum-likelihood  coupling matrices and the true coupling matrix.
+Logarithmic scale is chosen for the $y$-axis because of large values of the error at low $\gamma$.
+The inset in both panels show the ratio between the two errors.}
+	\label{fig:error_1_L4}
 \end{figure*}
 
-\begin{figure*}[!htb]
-	\includegraphics[keepaspectratio=true,width=1.0\textwidth]{Figures/couplings_panel_frob_dist_vs_lin_reg_slope_L10_M1024_same_range.png}
-\vspace{-1mm}
-\caption{{ Fobenious distance and  slope of linear regression between empirical ($\bm J_{0}$) / inferred couplings matrix ($\bm J_{max}$) and initial couplings matrix $\bm J_{init}$.
-Each point correspond to a different sampling-inference process as result of simulations with different initial parameters: $\gamma_{init},\bm J_{init}$.
-The length of the sequence vectors was $L=10$ and the number of sequences  $M=1024$ as consequence of a binary non-homogeneous tree with $K=10$ levels.
-Three different regimes were explored incrementing from left to right the  phylogenetic bias present at leaves sequences where \textbf{A}: weak phylogenetic signal and   $\gamma_{init}=0.1*\gamma_c$, \textbf{B}: intermediate phylogenetic signal  and $\gamma_{init}=0.3*\gamma_c$, \textbf{C}: strong phylogenetic signal and $\gamma_{init}=0.5*\gamma_c$.}}
-\label{results_corr_map_J_10_n10}
-\end{figure*}
+An interesting way to illustrate the benefits of reconstructing the covariance matrix using knowledge of the tree is to evaluate the effective number of samples that is gained by doing so. 
+Figure \sref{fig:error_vs_nseqs} shows the $l2$-error between the true covariances and the empirical covariances computed from an \emph{i.i.d.} sample of size $N$. 
+The error is as expected increasing for decreasing values of $N$. 
+We can use this as a calibration to express values of the $l2$-error in terms of an effective sample size.   
+For example, the error reached by $\bm{C}_{emp}$ at for $\gamma/\gamma_d\in[0.5, 1]$ and $L=4$ corresponds to the one obtained for an \emph{i.i.d.} sample of size $\sim 16$, whereas it corresponds to a sample of size $\sim 32-64$ for $\bm{C}_{max}$.
+Thus, our correction is equivalent to doubling the number of effective samples.
 
-\begin{figure*}[!htb]
-	\includegraphics[keepaspectratio=true,width=1.0\textwidth]{Figures/covariance_panel_pearson_corr_vs_lin_reg_slope_L10_M1024_same_range.png}
-	\vspace{-1mm}
-\caption{{  Pearson correlation and  slope of linear regression between empirical ($\bm C_{0}$) /inferred ($\bm C_{max}$) covariance  matrix and to  initial covariance matrix $\bm C_{init}=\bm J^{-1}_{init}$.
-Each point correspond to a different sampling-inference process as result of simulations with different initial parameters: $\gamma_{init},\bm J_{init}$.
-The length of the sequence vectors was $L=10$ and the number of sequences  $M=1024$ as consequence of a binary non-homogeneous tree with $K=10$ levels.
-Three different regimes were explored incrementing from left to right the  phylogenetic bias present at leaves sequences where \textbf{A}: weak phylogenetic signal and   $\gamma_{init}=0.1*\gamma_c$, \textbf{B}: intermediate phylogenetic signal  and $\gamma_{init}=0.3*\gamma_c$, \textbf{C}: strong phylogenetic signal and $\gamma_{init}=0.5*\gamma_c$.
-\textbf{center}: $\gamma=\gamma_i$, \textbf{right} $\gamma=\gamma_s$.
-}}
-\label{results_corr_map_C_10_n10}
-\end{figure*}
+Finally, we assess the performance of our method in improving the prediction of interactions between gaussian variables $\{x_a\}$. 
+We consider that two variables $x_a$ and $x_b$ interact if the corresponding entry in the coupling matrix is non-zero, that is $J_{ab}\neq 0$. 
+We predict these interactions by taking the largest $n$ elements (in absolute value) of the inferred coupling matrix, resulting in $n$ predictions. 
+The fraction of these $n$ predictions that correspond to non-zero entries in the true matrix defines the positive predictive value (PPV). 
+This problem is equivalent to the one of predicting contacts in a protein structure
 
-\begin{figure*}[!htb]
-	\includegraphics[keepaspectratio=true,width=1.0\textwidth]{Figures/PPV_panel_L10_M1024.png}
-	\vspace{-1mm}
-\caption{{  Positive predictive values (PPV) averaged over 30 samples,  for each of the three different regimes: \textbf{A}: weak phylogenetic signal and   $\gamma_{init}=0.1*\gamma_c$, \textbf{B}: intermediate phylogenetic signal  and $\gamma_{init}=0.3*\gamma_c$, \textbf{C}: strong phylogenetic signal and $\gamma_{init}=0.5*\gamma_c$.
-Positive predictive values are found as a proportion of agreement in the strongest couplings of the matrix $\bm J_{init}$ .
-The length of the sequences was $L=10$ so that couplings matrix has $10^2$ entries of which on average only 16 off diagonals elements  are nonzero  given the sparse character of the initial matrices.
-}} .
-\label{PPV_n10}
-\end{figure*}
+Figure $\ref{fig:PPV_L4}$ shows the PPV as a function of the number of predictions for different values of $\gamma$ and $L=4$ (see Fig. \sref{fig:PPV_L10} for the $L=10$ case). 
+In this case, the coupling matrix only has $6$ independent non-diagonal elements, and only 6 predictions can be made. 
+Our correction systematically outperforms the predictions from the empirical coupling matrix, with an always larger PPV. 
+This gain is negligible in the extreme regimes of very high $\gamma$, where the prediction is close to identical to the one obtained with an \emph{i.i.d.} sample, or very low $\gamma$, where it is essentially random.
+It is however much larger in the intermediate regime, with a significantly improved prediction in the region $\gamma/\gamma_d\in[0.5,1]$.
+
+\begin{figure}[!htb]
+	\centering
+	\includegraphics[keepaspectratio=true,width=1.0\textwidth]{Figures/PPVs_L4.png}
+	\caption{Quality of prediction of interactions for  different values of $\gamma$ and system size $L=4$.
+Interactions are defined as non-zero elements of the coupling matrix.
+In the $L=4$ case, there are $6$ possible interactions.
+Predictions are made by taking the largest elements (in absolute terms) of the inferred coupling matrix.
+The PPV is the fraction of correctly predicted contacts for a given number of predictions.}
+	\label{fig:PPV_L4}
+\end{figure}
 
@@
 \label{sec:discussion}
+
+What is this useful for? 
+-
 
@@
 
-\subsection{Joint probability of leaves is a gaussian} % (fold)
-\label{sec:joint_probability_of_leaves_is_a_gaussian}
-
-	A quick description of the idea. 
-	Let's assume that we have a (sub-)tree $\mathcal{T}$ constituted of nodes $\mathbf{X}=(\vx_1,\ldots,\vx_N)$. $\mathbf{X}$ comprises all the leaves \emph{and} internal nodes. 
-	We assume that $P(X)$ is a normal distribution, that is we can write it as 
-	\begin{equation}
-		P(\mathbf{X}) \propto \curlynormal{\mathbf{X}^{T}\mathbb{C}^{\mathcal{T}}\mathbf{X}},
-		\label{eq:tree_gaussian}
-	\end{equation}
-	with $\mathbb{C}^{\mathcal{T}}$ being the corresponding correlation matrix. 
-	Note that the superscript $\mathcal{T}$ implies that the correlation matrix depends on the shape of the tree .
-	We will now ``graft'' another node $\vec{y}$ on $\mathcal{T}$ by joining it to some node $\vx_r\in \mathbf{X}$. \pierre{Need a small figure to sketch the idea. Will draw it later.}
-	It is assumed that the evolution separating the configurations of $\vx_r$ and of $\vec{y}$ has taken place according to the Ornstein-Uhlenbeck (OU) process.
-	If that is the case, the distribution of $\vec{y}$ conditional on $\vx_r$, $P(\vec{y}\vert\,\vx_r)$ is described by equation~\ref{eq:OUpropagator} of the main text. 
-	We will now prove that the joint distribution $P(\vec{y},X)$ is a normal distribution. 
-	If we achieve this, it follows that if the configurations of all nodes in a tree were generated by the OU process, then their joint distribution is a Gaussian. 
-	Since node $\vec{y}$ attaches only to $\mathcal{T}$ only through node $\vx_r$, we can write
-	\begin{equation*}
-		P(\vec{y},X) = P(\vec{y}\vert\,X)P(X) = P(\vec{y}\vert \,\vx_r)P(X).
-	\end{equation*}
-	Combining equations~\ref{eq:OUpropagator} of the main text and \ref{eq:tree_gaussian}, we obtain
-	\begin{equation}
-		P(\vec{y},X) = \curlynormalpar{\vec{y}^{T}\iSig \vec{y} - 2\vx_r^{T}\Lam\iSig\vec{y} + \vx_r\Lam^2\iSig\vx_r + \mathbf{X}\mathbb{C}^{\mathcal{T}}\mathbf{X}}.
-	\end{equation}
-	If we now define the joint vector $\mathbf{X}'=(\vx_1,\ldots,\vx_N,\vec{y})$, with $\vec{y}$ having index $N+1$, it is clear that $\mathbf{X}'$ is normally distributed with a new correlation matrix $\mathbb{C}'$ having the following properties \pierre{the same if $i,j$ are different from $r$ or $N+1$, and new values if they correspond to $r$ or $N+1$}...
-\\
-	Thus, for a tree on which configuration of nodes have been obtained by the OU process, the joint distribution of configurations of all nodes is a normal distribution. 
-	This implies that the distribution of any subset of nodes is also a Gaussian, and thus that the joint distribution of \emph{leaves} is a Gaussian.
-
-% section joint_probability_of_leaves_is_a_gaussian (end)
-
 % section supplementary_material (end)
 
+\subsection{Parametrizations of  eigenvectors } % (fold)
+\label{sub:parametrizations_of_eigenvectors}
+\subsubsection{Parametrization in term of generalized Eulerian angles}
+The aim is to  parameterize each base vector $\vec{s}_a$ according to the $a-th$ column of an arbitrary orthogonal matrix $\bm S$ parameterized   in terms of $L(L-1)/2$ independent variables  $\theta_ {pq}, p =1,2, ..., L; q =1,2, ..., L; p <q $, named Generalized Eulerian angles (Eulerian angles is for L=3).
+
+To construct this matrix we start from a  transformation in a two-dimensional subspace of an $L-dimensional$ space which is given by an L-dimensional matrix of the  form:
+
+\begin{equation} 
+\bm T_{pq} =  \left(
+\begin{array}{cccccc}
+1 &   &  &  &  &\\
+ & 1& & & & \\
+& & \cos\theta_{pq} & & \sin\theta_{pq}\\
+& &  & 1 & &\\
+& &  -\sin\theta_{pq}& & \cos\theta_{pq} &\\
+& &  & &  & 1\\
+\end{array}
+\right)
+\end{equation}
+where all diagonal elements are unity except the diagonal elements in the $pth$ column and the $qth$ column, which are $\cos\theta_{pq}$ and all off-diagonal elements are zero except the one corresponding to the intersection of the $pth$ row and the $qth$ column, which is $\sin\theta_{pq}$, and that on the intersection of the $qth$ row and the $pth$ column, which is $-\sin\theta_{pq}$.
+There are $L(L - 1)/2$  matrices of the form indicated in Equation (1), corresponding to all choices of $p$ and $q$ with $p <q$.
+
+An arbitrary  $L-dimensional$ orthogonal matrix $\bm S$ can be represented as a product of these $L(L-1)/2$ orthogonal matrices with appropriate values of the $L(L-1)/2$ independent parameters $\theta_{pq}$.
+
+\begin{equation}
+\bm S (\bm \theta)=\prod_{\left\lbrace pq\right\rbrace }T_{pq}(\theta_{pq})
+\end{equation}
+
+ This matrix is in charge of rotations of vectors in $L-dimensional$ space. For $L=3$ we get the classical Eulerian matrix.
+
+ In reference [7] is explained an recursive algorithm to efficiently perform this multiplication,  as well as the   construction of the matrix   derivatives respect to parameters $\theta_{pq}$.
+
+For the first, equation  (2) is transformed in  the recurrence equations:
+
+\begin{equation}
+\bm S(\vec{\theta} )=\bm T^{(L)}
+\end{equation}
+
+with $\theta_ {pq}, p =1,2, ..., L; q =1,2, ..., L; p <q$ and
+
+\begin{equation}
+T^{(n)}_{kl}=\cos\theta_{kn} * t^{(n)}_{kl}-\sin\theta_{kn}* s^{(n)}_{kl}
+\end{equation}
+with
+\begin{equation}
+\begin{split}
+&s^{(n)}_{k+1,l}=\sin\theta_{kn} * t^{(n)}_{kl}+\cos\theta_{kn}*s^{(n)}_{kl}\\
+&s^{(n)}_{k+1,n}=\cos\theta_{kn}*s^{(n)}_{kn}\\
+&s^{(n)}_{1,l}=-\delta_{ln}\\
+&\theta_{nn}=\pi/2\\
+&\end{split}
+\end{equation}
+and
+\begin{equation} 
+\bm t^{(n)} =  \left(
+\begin{array}{cc}
+\bm T^{(n-1)} & \bm 0  \\
+\bm 0 &  \bm I\\
+
+\end{array}
+\right)
+\end{equation}
+
+This mean that the eigenvectors $\vec{s}_a$ of the matrix $\bm C$ with dimensions $L\times L$ can be written as:
+\begin{equation}
+s^{k}_a=T^{(L)}_{.,a}=\cos\theta_{kL} * t^{(L)}_{ka}-\sin\theta_{kL}* s^{(L)}_{ka} \;\;\; \text{for} \;k=1,..,L
+\end{equation}
+
+Then there is another  important problem : how to determine parameters $\bm \theta$  given an orthogonal matrix $\bm Q$ such that all equations
+
+$$S_{ij}(\bm \theta)=Q_{ij}$$
+
+are satisfied.
+This system of nonlinear transcendental equations  can't be algebraically solved, however it's possible overcome  this issue finding the set of $\bm \theta$ which minimize the function:
+\begin{equation}
+f(\bm \theta)=\sum_{i,j}\left[ Q_{ij}-S_{ij}(\bm \theta)\right] ^2
+\end{equation}
+This is useful when we initialize parameters $\bm \theta$  from the matrix formed by eigenvectors of the empirical covariance matrix.
+\subsubsection{Parametrization in term of the exponential of a skew-symmetric matrix }
+The exponential of a skew-symmetric matrix $\bm X=-\bm X^T$  : 
+ \begin{equation}
+ \bm S=\exp (\bm X)
+ \end{equation}
+
+ is  an  orthogonal  matrix : $\exp(X)^T=\exp(X^T)=\exp(-X)=\exp(X)^{-1}$.
+
+ The derivatives on $L(L-1)/2$ independent variables of $\bm X$  is formally defined by
+
+ \begin{equation}
+ \frac{\partial \bm S}{\partial X_{jk}}=\lim_{h\rightarrow 0} \frac{1}{h} \left( \exp(\bm X+ h \bm E^{jk}) - \exp(\bm X)\right) 
+ \end{equation}
+
+ where $\bm E^{jk}$ for $j > k$ is defined as a skew-symmetric matrix that has only two nonzero elements:
+ \begin{equation}
+ E^{pq}_{jk}=\delta_{pj}\delta_{qk}-\delta_{pk}\delta_{qj}
+ \end{equation}
+
+To obtain   a skew-symmetric matrix    $\bm X$ from an orthogonal matrix $\bm Q$    is enough invert the exponential relation :
+ $$\bm X=\log \bm Q$$
+
+  If we write eigenvectors of the likelihood function as the columns of the exponential of the skew-symmetric matrix $\bm X$, then we are able to perform the optimization over it's $L(L-1)$ independent variables of $\bm X$.
+
+ Furthermore, as the eigenvectors were parameterized by an algebraic function, the complete likelihood  is described in terms of arithmetics operation and elementary functions.
+This allow to compute  the gradient via  automatic differentiation (AD), making computations in our optimization process faster.
+ The idea  is to represent a function as a computational graph, also called a Wengert list , where each node  in this list will represent an intermediate result of the computation.
+The intermediate results can then be assembled using the chain rule to get the final derivative we’re looking for.
+There is two    main algorithms  to traverses the chain rule in AD: forward mode and reverse mode, here we used reverse mode  algorithm  which is the of choice for back-propagation in deep learning .
+In particular we implemented  it using the Julia package Zygote.jl .
+
+ \subsection{ Homogeneous and fully balanced tree }
+
+Let assume that the tree is binary and completely homogeneous, all branches have the same length $\Delta t$.
+For instance the sequences covariance matrix for a binary tree with $K=2$ branching events and four leaves is
+\begin{equation} 
+\mathbb{G} =  \left(
+\begin{array}{cccc}
+{ \bm C} & {\bm C} \Lambda ^{2\Delta t} & {\bm C} \Lambda ^{4\Delta t} & {\bm C} \Lambda ^{4\Delta t} \\
+{\bm C} \Lambda ^{2\Delta t} & {\bm C} & {\bm C} \Lambda ^{4\Delta t} & {\bm C} \Lambda ^{4\Delta t} \\
+{\bm C} \Lambda ^{4\Delta t} & {\bm C} \Lambda ^{4\Delta t} & {\bm C} & {\bm C} \Lambda ^{2\Delta t} \\
+{\bm C} \Lambda ^{4\Delta t} & {\bm C} \Lambda ^{4\Delta t} & {\bm C} \Lambda ^{2\Delta t} & {\bm C} \\
+\end{array}\label{eq:G}
+\right)
+\end{equation}
+
+The associatted matrix $\bm G^a=z(\rho_a,\gamma,\Delta t)$ defined by (14) became
+
+\begin{equation} 
+\bm G^a =  \left(
+\begin{array}{cccc}
+{  \rho^{-1}_a } & {\rho^{-1}_a } e ^{-2\gamma\rho_a\Delta t} & {\rho^{-1}_a } e ^{-4\gamma\rho_a\Delta t}& {e ^{-4\gamma\rho_a\Delta t}} \\
+{\rho^{-1}_a } e ^{-2\gamma\rho_a\Delta t} & {\rho^{-1}_a } & {\rho^{-1}_a } e ^{-4\gamma\rho_a\Delta t} & {\rho^{-1}_a } e ^{-4\gamma\rho_a\Delta t} \\
+{\rho^{-1}_a } e ^{-4\gamma\rho_a\Delta t} & {\rho^{-1}_a} e ^{-4\gamma\rho_a\Delta t}& {\rho^{-1}_a } & {\rho^{-1}_a } e ^{-2\gamma\rho_a\Delta t}\\
+{\rho^{-1}_a } e ^{-4\gamma\rho_a\Delta t} & {\rho^{-1}_a } e ^{-4\gamma\rho_a\Delta t} & {\rho^{-1}_a } e ^{-2\gamma\rho_a\Delta t} & {\rho^{-1}_a } \\
+\end{array}\label{eq:G_a}
+\right)
+\end{equation}
+
+For hypergeometric matrices as \ref{eq:G_a} there is $K+1$ different eigenvalues  given by:
+
+\begin{equation}
+\label{eq:lambda_hyper}
+\lambda_k(\rho_a,\gamma) =\rho^{-1}_a *
+\begin{cases}
+ (1+\sum_{l=1}^{k-1} 2^{l-1} e ^{-2l\gamma\rho_a\Delta t}-2^{k-1} e ^{-2k\gamma\rho_a\Delta t})\;\;\; k\in[1,K] \\
+ (1+\sum_{l=1}^{K} 2^{l-1} e ^{-2l\gamma\rho_a\Delta t}) \;\;\;\;\;\;\;k=K+1
+\end{cases}
+\end{equation}
+
+where $\lambda_{K+1}\ge\lambda_{K}\cdots\ge\lambda_{1}$. For $k<K+1$ degenerancy of eigenvalues $\lambda_k$ is given by : $d_k=2^{K-k}$.
+
+The associated eigenvectors are independent of the parameter $\rho_a$ and reflects the events in the phylogenetic tree  where each eigenvector  $k$ captures the duplication events in the $(K+1 -k) 	st$ generation.
+Summarizing :
+
+\begin{equation} 
+\nonumber
+ \vec{u}_k =
+\begin{cases}
+{(\underbrace{\overbrace{1,\ldots,1}^{2^{k-1}},\overbrace{-1,\ldots,-1}^{2^{k-1}}}_Q,0,\ldots,0)\ \bigcup \Gamma(u_k) } \;\;\; k\in[1,K] \\
+(1,1,1,\ldots,1,1,1)\;\;\;\;\;\;\;k=K+1
+\label{eq:eigenvec_simpler_case}
+\end{cases}	
+\end{equation}
+
+where $\Gamma(\vec{u}_k )$ represent $d_k$ combinations resulting of shift the block of length $Q$ without overlapping, generating all eigenvectors with the same corresponding eigenvalue $\lambda_k$.
+The eigenvectors are orthogonal to eachother, and can be  normalized and arranged horizontally into a matrix $U$.
+
+Then to compute the gradient of the likelihood we only need  derivatives of $\lambda_k(\rho_a,\gamma)$  respect to $\rho_a$ and $\gamma$, which can be obtained from direct derivation of expression \ref{eq:lambda_hyper}.
+
+\subsection{Optimization scheme}
+\label{sub:optimization_scheme}
+
+The proposed inference scheme was transformed into a multidimensional nonlinear optimization problem for which a variant  of quasi-Newton Methods (QNMs) was used.
+The main feature in QNMs is that Hessian matrix $\hat{H}$  does not need to be computed instead it’s approximated.
+The Hessian approximation $\hat{H}$ is chosen to satisfy the secant equation:
+
+$$\nabla L(\vec{x}_k+\Delta\vec{x})=\nabla L(\vec{x}_k)+\hat{H} \Delta\vec{x} $$
+
+for $n-dimensions$ $\hat{H}$ is undetermined, the various QNMs differ in their choice of the solution to the secant equation.
+We used LBFGS variant  (limited memory version of BFGS), this particular method is based in choosing $\hat{H}$ as a positive definite matrix where
+
+$$\hat{H}_{k+1}=\hat{H}_{k}+\frac{y_k y^{T}_k}{y_k\Delta x_k} - \frac{\hat{H}_k\Delta x_k(\hat{H}_k\Delta x_k)^T}{\Delta x^T_k\hat{H}_k\Delta x_k} \;\;\;\text{and}\;\;\; y_k=\nabla L(\vec{x}_k+1)-\nabla L(\vec{x}_k)$$
+
+The method implementation was done using   NLopt package (see reference).
+
+\subsection{Supplementary figures}
+\label{sub:supplementary_figures}
+
+\begin{figure}[!htb]
+	\centering
+	\includegraphics[keepaspectratio=true,width=1.0\textwidth]{Figures/error_Ciid_vs_numb_of_seq.pdf}
+	\caption{Relative $l2$-error between the empirical covariance matrix calculated from an \emph{i.i.d.} sample and the true covariance matrix, for system sizes $L=4$ and $L=10$.
+The dashed vertical line corresponds to the number of leaves of the tree used in the simulations.}
+	\label{fig:error_vs_nseqs}
+\end{figure}
+
+\begin{figure}[!htb]
+	\centering
+	\includegraphics[keepaspectratio=true,width=0.45\textwidth]{Figures/scatter_gamma_L4_100.pdf}
+	\caption{Inferred $\gamma$ values as a function of real $\gamma$, for system size $L=4$.
+$\gamma_{emp}$ is the value obtained by the process described in section \ref{sub:initializing_parameters} of the SM.
+$\gamma_{max}$ is the value inferred by the maximum-likelihood calculation.
+	The inset represents the ratio of both inferred parameters $\gamma_{emp}$ or $\gamma_{max}$ to the real $\gamma$.}
+	\label{fig:scatter_gamma_L4}
+\end{figure}
+
+\begin{figure}[!htb]
+	\centering
+	\includegraphics[keepaspectratio=true,width=0.55\textwidth]{Figures/Pearson_correlation_rate_full_range.pdf}
+	\caption{  Ratio between Pearson correlation of the  maximum-likelihood  and true covariance  matrix to the Pearson correlation of the  empirical and true covariances matrices for the two system sizes $L=4$ and $L=10$.}
+	\label{fig:pearson_corr_comp_L4}
+\end{figure}
+
+\begin{figure}[!htb]
+	\centering
+	\includegraphics[keepaspectratio=true,width=0.65\textwidth]{Figures/corr_C_L10_balanced_tree_100_rep.pdf}
+	\caption{ Pearson correlation between empirical /maximum-likelihood   covariance matrices and the true covariance matrix, the inset plot represent the ratio between the person correlation for the  maximum-likelihood covariance matrix and the one for the empirical covariance matrix.}
+	\label{fig:pears_L10}
+\end{figure}
+
+\begin{figure}[!htb]
+	\centering
+	\includegraphics[keepaspectratio=true,width=0.65\textwidth]{Figures/scatter_gamma_L10_100.pdf}
+	\caption{Inferred $\gamma$ values as a function of real $\gamma$, for system size $L=10$.
+$\gamma_{emp}$ is the value obtained by the process described in section \ref{sub:initializing_parameters} of the SM.
+$\gamma_{max}$ is the value inferred by the maximum-likelihood calculation.
+	The inset represents the ratio of both inferred parameters $\gamma_{emp}$ or $\gamma_{max}$ to the real $\gamma$. }
+	\label{fig:scatter_gamma_L10}
+\end{figure}
+
+ \begin{figure*}[!htb]
+ 		\begin{subfigure}{}
+ 			\centering\includegraphics[keepaspectratio=true,width=0.45\textwidth]{Figures/epsilon_error_C_L10_balanced_tree_100.pdf}
+ 		\end{subfigure}
+ 		\hspace{1mm}
+ 		\begin{subfigure}{}
+ 			\centering\includegraphics[keepaspectratio=true,width=0.45\textwidth]{Figures/epsilon_error_J_L10_balanced_tree_100.pdf}
+ 		\end{subfigure}
+ 	\caption{\textbf{Left:} Relative $l2$-error between empirical or maximum-likelihood covariance matrices and the true covariance matrix.
+\textbf{Right}: Relative $l2$-error between empirical /maximum-likelihood coupling matrices and the true coupling matrix.
+Logarithmic scale is chosen for the y-axis because of large values of the error at low $\gamma$.
+The inset in both panels show the ratio between the two errors.
+For system size $L=10$.}
+ 	\label{fig:error_1_L10}
+ \end{figure*}
+
+\begin{figure}[!htb]
+	\centering
+	\includegraphics[keepaspectratio=true,width=1.0\textwidth]{Figures/PPVs_L10.png}
+	\caption{Quality of prediction of interactions for  different values of $\gamma$ and system size $L=4$.
+Interactions are defined as non-zero elements of the coupling matrix.
+In the $L=4$ case, there are $6$ possible interactions.
+Predictions are made by taking the largest elements (in absolute terms) of the inferred coupling matrix.
+The PPV is the fraction of correctly predicted contacts for a given number of predictions.}
+	\label{fig:PPV_L10}
+\end{figure}
+
 \end{document}

### 2021-01-26 main.tex
first version intro - edits

@@
 \title{Global multivariate model learning from hierarchically correlated data}
-\date{}
+% \date{}
 
@@
 %
-\author{Pierre Barrat-Charlaix} 
-\affiliation{Biozentrum, Universität Basel, Basel, Switzerland}
-%
 \author{Alejandro Lage} 
@@
 \author{Martin Weigt} 
-\email{Correspondence to: Martin Weigt, \bf{martin.weigt@upmc.fr}}
 \affiliation{Sorbonne Universit\'{e}, CNRS, Institut de Biologie
@@
   Quantitative -- LCQB, Paris, France}
+%
+
+\author{Pierre Barrat-Charlaix} 
+\email{Correspondence to: Pierre Barrat-Charlaix, \bf{pierre.barrat@unibas.ch}}
+\affiliation{Biozentrum, Universit\"at Basel, Basel, Switzerland}
 
@@
 In the last decades, biology has seen impressive progress in experimental techniques which has resulted in a large increase of available data. 
-This is especially visible in the case of biological sequences,  with databases now harboring a vast amount of high-quality DNA or protein sequences. [UNIPROT/GENBANK]
-This has in turn fuelled a development of quantitative methods to model forces guiding the evolution of sequences or other biological traits [REFS?]. 
-A common idea in this context is that it is possible to use characteristics of homologous genes or organisms to construct quantitative models of the selection acting on them. 
-A successful example is the representation of protein sequences by probabilistic models in the so-called DCA method [REF].
+This is especially visible in the case of biological sequences,  with databases now harboring a vast amount of high-quality DNA or protein sequences \cite{sayers_genbank_2019,uniprotconsortium_uniprot_2018}.
+This has in turn fuelled a development of quantitative methods to model forces guiding the evolution of sequences or other biological traits [REFS/EXAMPLES?]. 
+A common idea in this context is that it is possible to use characteristics of homologous genes or organisms to construct models of the selection acting on them. 
+A successful example in this regard is the representation of protein sequences by probabilistic models in the so-called DCA method \cite{levy_potts_2017,cocco_inverse_2018}.
 
-Models built in this way usually assume that biological characteristics of related sequences or organisms are evolving under the same constraints but are independent from each other in the statistical sense.
+Models built in this way usually assume that biological characteristics of related organisms are evolving under the same constraints but are independent from each other in the statistical sense. 
 This allows simple formulation for model learning. 
-However, phylogenetic relations between evolving entities imply that these cannot be considered independent from each other, and doing so can result in important biases when learning models. 
-For instance, it has been shown that phylogenetic relations between protein sequences induce non-trivial correlations that are not related to protein function [COLWELL paper].
+However, the phylogenetic relations between evolving entities are in direct contradiction with this simplifying hypothesis, and ignoring them can result in biases when learning models. 
+For instance, it has been shown that phylogenetic relations between protein sequences induce non-trivial correlations that are not related to protein function \cite{qin_power_2018}.
 
-In order to correct for these biases, it is necessary to disentangle the statistical effects due to selection forces and those due to phylogeny. 
-However, this becomes a complex problem when dealing with complex models where traits under study are not independent from each other. 
-Accounting for these biases then requires one to understand how correlated traits evolve along a tree. 
-History of modeling this with BM or OU processes [Felsenstein, Hansen]. Also, field of PCM
+In order to correct for these biases, it is necessary to disentangle the effects due to selection forces and those due to phylogeny. 
+However, this becomes a hard problem when dealing with complex models where traits under study are not independent from each other. 
+Accounting for these biases then requires one to understand how interdependent traits under selection evolve along a tree. 
+A historically well-known way to represent such processes is to use Ornstein-Uhlenbeck dynamics (OU), which models traits as a Gaussian vector evolving in a quadratic potential that represents selection forces \cite{uhlenbeck_theory_1930, felsenstein_phylogenies_1988, hansen_stabilizing_1997}.
+This method is commonly used in the field of phylogenetic comparative methods (PCM) \cite{bartoszek_phylogenetic_2012,mitov_fast_2020}.
 
-Increasing data --> new quantitative techniques in biology. 
-One big recent success: pairwise sequence models, blabla
-However, models learned from data with phylogenetic correlations can suffer from biases.
-To account for these biases in the case of complex models, it's necessary to understand how correlated traits evolve along a tree (or: quantities that are modeled and typically not independent evolve along a tree)
+This modeling approach is \emph{a priori} limited to continuous traits, but could potentially be used for protein sequences combined with a continuous-variable approximation, that has successfully been used in the past \cite{jones_psicov_2012,baldassi_fast_2014}.
+In this context, the equilibrium distribution reached by the OU process represents the probability distribution given by the DCA method, which can be used to predict non-trivial structural contacts in the protein fold, effects of amino-acid mutations or even designing novel functional sequences \cite{morcos_direct-coupling_2011, figliuzzi_coevolutionary_2016, russ_evolution-based_2020}.
 
-There is a history of looking at multi parameter complex models and their relation with phylogeny. 
-in this context, it is often assumed that quantitative traits that evolve under selection can be modeled using an OU process. 
-This is usually done in the field of PCMs, with most common models being Brownian motion or OU processes.
-
-Here, we want to understand how to find parameters values of an OU gaussian process from data correlated through a tree. 
-This could be applied to the evolution of any correlated biological traits, but also potentially to protein sequences.
-
-The MS is divided as follows: >>>
-
-\paragraph{Old} % (fold)
-\label{par:old}
-
-Theory and applications of the inverse problem of statistical physics has had great development in recent years since the availability of large amounts of data on a microscopic scale produced by the continuous technological progress.
-In this context statistical physics paths: from model parameters to  observables, is reversed, toward find global models which describe the statistics of  a set of observed  configurations \cite{Inverse_problem_Berg}.
-
-Most of the work has been focus in equilibrium reconstruction , where the aim is to learn max Entropy models, Boltzman distribution, from independent and identically distributed equilibrium configurations.
-Another kind of methods deal with parameters learning from either time series data or from samples of the non-equilibrium steady state: non-equilibrium reconstruction.
-
-Less effort has been done for data biased by population structured signal, where  samples  are not independent because of presence of relatedness between configurations.
-These structure could be for instance of  geographical or demogrhapic  nature in a population or  phylogenetic  nature  related to a evolutionary processes shaping trait variation among species.
-Here we are going to deal with data given by sequences phylogenetically related  through a tree, which describes the hierarchical correlation between configurations.
-An important problem is the  distinction   of patterns induced by species traits or by phylogeny.
-
-Our source of inspiration came from the inverse problem in proteins \cite{Inverse_problem_proteins}, where  covariance signal present in an alignment of homologous proteins is described via Potts model.
-This covariance signal is assumed a consequence of phenotypics constraints like protein structure and function.
-However it is known that phylogeny relationships in homologous sequences  contaminate the statistics of  the alignment used to learn the equilibrium model.
-Therefore a direct inference on this alignments  leads to the existence of Potts parameters that attempt to model the full biased statistics limiting their capacity to predict phenotypic properties.
-
- To define and solve the inverse problem of statistical physics for data samples with hierarchical  relatedness structure is the aim of this work.
-
-Main difficulties to address  this kind of problem  come from the necessity to use  a proper global evolutionary model which describes the relation between sequences edge connected in a tree and the discrete  characters of data variables which makes certain calculations intractable \cite{entropy_paper}.
-
-Here we  adopt an approximation where we forget about the discrete nature of variables and model them by continuous variables instead.
-This transforms the Potts model into a Gaussian distribution \cite{Baldassi}, making the design of a global propagator tractable.
-We propose a model of evolution described by a multivariate Ornstein-Uhlenbeck process \cite{gardiner}, which is used  to generate evolutionary related data given an arbitrary phylogenetic tree.
-The corresponding inverse problem became in the inference of the Gaussian potential (Potts parameters) from the knowledge of tree leaves configurations and the dynamics of evolution.
-
-To this aim, we first review in section \ref{sec:ornstein_uhlenbeck_dynamics}, main characteristics of a multivariate Ornstein-Uhlenbeck process.
-In section \ref{evolutionary process} we describe the evolutionary process, describing the algorithm used to generate data phylogenetically biased.
-The statement of the problem in a Bayesian framework is shown in section \ref{statement of the problem}.
-Section \ref{methods} is dedicated to describe methodology used to solve the inverse problem.
-Results for evolution of continuos variable sequences and  the application for configuration of discrete variables is presented in section \ref{Results}.
-Finally the conclusion and outlook in \ref{conc}.
-
-% paragraph old (end)
+In this work, we are interested in constructing an inference method for parameters of an OU process from data correlated through a tree.
+Our approach is purely methodological, and the data can represent any set of continuous phenotypic traits, \emph{e.g.} from different organisms, with the tree indicating the phylogenetic relations between data points.
+Inferred parameters then represent the selection forces without biasing effects from the phylogeny. 
+The manuscript is divided as follows: we first review in section \ref{sec:ornstein_uhlenbeck_dynamics} the main characteristics of the multivariate OU process. 
+We then describe the setting of the inference problem that we want to solve in section \ref{sub:statement_of_the_problem}, propose a solution in sections \ref{sub:Calculation_of_the_likelihood} and \ref{sub:maximization_of_the_likelihood}.
+Finally, we present results obtained on simulated data in section \ref{sec:Results}, with the context of pairwise models of protein sequences in mind.
 
@@
 We are interested in inferring the coupling matrix from a given amount of observed states $\vx$ of the system.
-If these observations were independent from each other, due to the simple Gaussian form of Eq.~(\ref{eq:eq_distribution}), $\bm{J}$ would simply be equal to the inverse of the empirical \emph{covariance matrix} of the data, written $\bm{C}=\bm{J}^{-1}$.
+If these observations were independent from each other, due to the simple Gaussian form of Eq.~\eqref{eq:eq_distribution}, $\bm{J}$ would simply be equal to the inverse of the empirical \emph{covariance matrix} of the data, written $\bm{C}=\bm{J}^{-1}$.
 
-However, here we are concerned with the case where observations are not independent. 
-On the contrary, they result from a dynamical process taking place during a finite amount of time. 
-Different datapoints are therefore correlated even if individually assumed to be at equilibrium. 
+However, we consider the case where observations are not independent. 
+On the contrary, they result from a dynamical process taking place during a finite amount of time, and different data-points are therefore correlated to each other.
 This dynamical process is described below.
@@
 Here, $\vec{\xi}(t)$ is a vector of uncorrelated white noise, and $\gamma^{-1}$ is the characteristic timescale governing the dynamics. 
-In short, Eq.~(\ref{eq:langevin}) states that the system described by $\vx$ undergoes Brownian motion in a quadratic energy landscape characterized by the coupling matrix $\bm{J}$.
+In short, Eq.~\eqref{eq:langevin} states that the system described by $\vx$ undergoes Brownian motion in a quadratic energy landscape characterized by the coupling matrix $\bm{J}$.
 
-We are not interested in $\vx$ directly, but rather in its probability distribution $P(\vx\vert\,\vx_0,\Delta t)$, \ie~in the probability to find the system in state $\vx$ knowing it was in state $\vx_0$ some time $\Delta t$ before.
-The Fokker-Planck equation corresponding to Eq.~(\ref{eq:langevin}) is straightforward to write, 
+We are not interested in $\vx$ directly, but rather in its probability distribution $P(\vx\vert\,\vx_0,\Delta t)$, \ie~in the probability to find the system in state $\vx$ knowing it was in state $\vx_0$ some time $\Delta t$ in the past.
+The Fokker-Planck equation corresponding to Eq.~\eqref{eq:langevin} is straightforward to write, 
 \begin{equation}
@@
 where the parenthesized expression on the right hand side is understood as an operator acting on $P$. 
-The solution to Eq.~(\ref{eq:FP}) is a multivariate normal distribution~\cite{singh2017multiOU}:
+The solution to Eq.~\eqref{eq:FP} is a multivariate normal distribution~\cite{singh2017multiOU}:
 \begin{equation}
@@
 \end{equation}
-Eqs.~(\ref{eq:OUpropagator}) and~(\ref{eq:def_lambda}) define a multivariate \emph{Ornstein-Uhlenbeck} (OU) process.
+Eqs.~\eqref{eq:OUpropagator} and~\eqref{eq:def_lambda} define a multivariate \emph{Ornstein-Uhlenbeck} (OU) process.
 
@@
 We also underline that $\Sig$ and $\vec{\mu}$ depend on $\Delta t$, although this dependence is not explicitly written in our notation to make it less heavy. 
+By taking $\gamma\Delta t \gg 1$ and using the fact that $\bm{J}$ has strictly positive eigenvalues, one immediately recovers Eq.~\eqref{eq:eq_distribution}, meaning that the OU process converges to the desired equilibrium distribution.
 
-By taking $\gamma\Delta t \gg 1$ and using the fact that $\bm{J}$ has strictly positive eigenvalues, one immediately recovers Eq.~(\ref{eq:eq_distribution}). 
-We can thus compute the joint distribution of two configurations $\vx_1$ and $\vx_2$ separated by time difference $\Delta t$ by multiplying Eqs.~(\ref{eq:eq_distribution}) and~(\ref{eq:OUpropagator}),
+We can compute the joint distribution of two configurations $\vx_1$ and $\vx_2$ separated by a time $\Delta t$ by multiplying Eqs.~\eqref{eq:eq_distribution} and~\eqref{eq:OUpropagator}, 
 \begin{eqnarray}
@@
 This equation illustrates the \emph{time reversibility} of the OU process. 
-Indeed, the distribution does not depend on which configuration $\vx_1$ or $\vx_2$ came first.
+Indeed, the distribution is symmetric in $\vx_1$ or $\vx_2$ and does not depend on which configuration came first.
 
 Equation \eqref{eq:OUjoint} allows for computing the joint covariance of the correlated equilibrium configurations $\vx_1$ and $\vx_2$. 
-The probability distribution in Eq.~(\ref{eq:OUjoint}) is normal with an inverse covariance matrix defined by blocks: $\Sig$ on the diagonal and $-\Lam^{\Delta t}\Sig$ off-diagonal. 
+The probability distribution in Eq.~\eqref{eq:OUjoint} is normal with an inverse covariance matrix defined by blocks: $\Sig$ on the diagonal and $-\Lam^{\Delta t}\Sig$ off-diagonal. 
 By inverting this block matrix, given that $\Lam$ and $\Sig$ commute and are invertible, one obtains the following covariance: 
@@
 
-Eq.~(\ref{eq:pairwisecov}) allows us to readily distinguish two regimes. 
+Eq.~\eqref{eq:pairwisecov} allows us to readily distinguish two regimes. 
 Let us call $\rho_a$ the eigenvalues of $\bm{J}$. 
@@
 Inversely, for $\Delta t / \tau_c \ll 1$, $\vx_1$ and $\vx_2$ are highly correlated, defining a \emph{strongly correlated} regime. 
-It should be noted that for $\Delta t = 0$, the joint correlation matrix of $\vx_1$ and $\vx_2$ becomes non invertible, and Eq.~(\ref{eq:pairwisecov}) becomes irrelevant.
+It should be noted that for $\Delta t = 0$, the joint correlation matrix of $\vx_1$ and $\vx_2$ becomes non invertible, and Eq.~\eqref{eq:pairwisecov} becomes irrelevant.
 Actually, $\vx_1$ and $\vx_2$ coincide at that point, \ie~we have $P(\vx_1,\vx_2|\,\Delta t=0) = P_{eq}(\vx_1) \times \delta(\vx_1 - \vx_2)$ using the $L$-dimensional Dirac distribution.
@@
 Formally, we assume that the data consists of $N$ real-valued vectors of length $L$, denoted $\{\vx_i\}\in\mathbb{R}^L$ with $i=1,...,N$. 
-Taken individually, the $\vx_i$ are distributed according to Eq.~\ref{eq:eq_distribution}, \ie~according to a multivariate Gaussian of zero mean and covariance $\bm C$.
+Taken individually, we assume that the $\vx_i$ are distributed according to Eq.~\eqref{eq:eq_distribution}, \ie~according to a multivariate Gaussian of zero mean and covariance $\bm C$.
 By construction, the equilibrium covariance between any pair of elements of a given vector $\vx=(x^1,...,x^L)$ is given by the inverse of the coupling matrix: $\langle x^a x^b\rangle-\langle x^a\rangle\langle x^b\rangle = \bm C_{ab} = (\bm J^{-1})_{ab}$ for all $a,b=1,...,L$.
@@
     \item The process starts at the root node $r$ with a state vector $\vx_r$ drawn from the equilibrium distribution $P_{eq}$.
-    \item On each branch $(n,a(n))$ of length $\Delta t_{n,a(n)}$ connecting node $n$ with its ancestral node $a(n)$, the dynamics follow Eq.~(\ref{eq:langevin}), starting from initial condition $\vx_{a(n)}$, and running for time $\Delta t_{n,a(n)}$.
-In other words, given the state $\vx_{a(n)}$ of the ancestral node, $\vx_n$ is sampled from $P(\vx_n |\, \vx_{a(n)}, \Delta t_{n,a(n)})$, see Eq.~(\ref{eq:OUpropagator})
+    \item On each branch $(n,a(n))$ of length $\Delta t_{n,a(n)}$ connecting node $n$ with its ancestral node $a(n)$, the dynamics follow Eq.~\eqref{eq:langevin}, starting from initial condition $\vx_{a(n)}$, and running for time $\Delta t_{n,a(n)}$.
+In other words, given the state $\vx_{a(n)}$ of the ancestral node, $\vx_n$ is sampled from $P(\vx_n |\, \vx_{a(n)}, \Delta t_{n,a(n)})$, see Eq.~\eqref{eq:OUpropagator}
     \item As a consequence, OU processes on branches stemming from common ancestral node evolve independently, but from an identical initial condition.
@@
 This process is thought to represent the evolution of biologic traits along a phylogenetic tree, with the leave nodes corresponding to traits observed in today's species.
-Note that due to the reversible nature of our OU process, the joint probability of any pair of leaf configurations $\vx_i$ and $\vx_j$, with $i,j\in\{1,...,N\}$, is given by  $P(\vx_i,\vx_2|\,\Delta t_{ij})$ (Eq.~(\ref{eq:OUjoint}), with $\Delta t_{ij}$ denoting the total branch length of the path connecting $i$ and $j$ in the tree.
+Note that due to the reversible nature of our OU process, the joint probability of any pair of leaf configurations $\vx_i$ and $\vx_j$, with $i,j\in\{1,...,N\}$, is given by  $P(\vx_i,\vx_j|\,\Delta t_{ij})$ (Eq.~\eqref{eq:OUjoint}), with $\Delta t_{ij}$ denoting the total branch length of the path connecting $i$ and $j$ in the tree.
 
@@
 Here, $P(\bm C, \gamma)$ can be any arbitrarily chosen prior distribution.
-The difficulty in Eq.~(\ref{eq:bayesian_likelihood}) lies in the estimation of the likelihood $P(\mathcal{D}\vert\bm C, \gamma, \mathcal{T})$,
+The difficulty in Eq.~\eqref{eq:bayesian_likelihood} lies in the estimation of the likelihood $P(\mathcal{D}\vert\bm C, \gamma, \mathcal{T})$,
 \ie~of the joint probability of the datapoints $\mathcal{D}=\{\vx_i\}_{i=1,...,N}$ for an OU process given by its parameters $\{\bm C, \gamma\}$ and the tree $\mathcal{T}$.
@@
 
-The joint distribution of two configurations $\vx_1$ and $\vx_2$ separated by time $\Delta t$ is given by Eq.~(\ref{eq:OUjoint}) and corresponds to a joint normal distribution. 
-This means that the vector $\vec{X}=[\vx_1, \vx_2]$, \emph{i.e.} the concatenation of vectors $\vx_1$ and $\vx_2$, follows a normal distribution with zero mean and variance described above in Eqs.~(\ref{eq:def_lambda}).
+The joint distribution of two configurations $\vx_1$ and $\vx_2$ separated by time $\Delta t$ is given by Eq.~\eqref{eq:OUjoint} and corresponds to a joint normal distribution. 
+This means that the vector $\vec{X}=[\vx_1, \vx_2]$, \emph{i.e.} the concatenation of vectors $\vx_1$ and $\vx_2$, follows a normal distribution with zero mean and variance described above in Eqs.~\eqref{eq:def_lambda}.
 Of importance here is that this property of the OU process can be extended to the joint distribution of any subset of nodes in a tree. 
@@
 Each block $\mathbb{G}_{ij}$ is describing the connected correlations between two data vectors $\vx_i$ and $\vx_j$, which are separated by time $\Delta t_{ij}$, resulting as the sum of all branch lengths of the path connecting $i$ and $j$ on tree $\mathcal{T}$.
-Because the OU process is time reversible, we can directly apply Eq.~(\ref{eq:pairwisecov}) and give all blocks of $\mathbb{G}$ in closed form,
+Because the OU process is time reversible, we can directly apply Eq.~\eqref{eq:pairwisecov} and give all blocks of $\mathbb{G}$ in closed form,
 \begin{equation}
@@
 
-Eq.~(\ref{eq:multiOU_jointall}) allows us to compute the log-likelihood of the data $\vec{X}$ as a function of $\vec{X}$ itself and of the joint covariance matrix.
+Eq.~\eqref{eq:multiOU_jointall} allows us to compute the log-likelihood of the data $\vec{X}$ as a function of $\vec{X}$ itself and of the joint covariance matrix.
 Indeed, taking its logarithm immediately gives
@@
 but this expression is impractical for any numerical evaluation due to the large dimension of $\mathbb{G}$.
-However, the particular block structure of $\mathbb{G}$ described in Eq.~(\ref{eq:jointcov_2}) allows us to simplify the expression.
+However, the particular block structure of $\mathbb{G}$ described in Eq.~\eqref{eq:jointcov_2} allows us to simplify the expression.
 To do so, we first introduce the eigenvalues and eigenvectors
@@
 By definition, we have $\rho_a>0$ for all $a$.
-Using now Eq.~(\ref{eq:jointcov_2}), we immediately see that the vectors $\vsa$ are also eigenvectors of the individual blocks $\mathbb{G}_{ij}$ with eigenvalues $z(\rho_a, \Delta t_{ij})$ where we introduced
+Using now Eq.~\eqref{eq:jointcov_2}, we immediately see that the vectors $\vsa$ are also eigenvectors of the individual blocks $\mathbb{G}_{ij}$ with eigenvalues $z(\rho_a, \Delta t_{ij})$ where we introduced
 \begin{equation}
@@
 
-Note that this decomposition of the eigenvectors leads to a drastic decrease in computational complexity for diagonalizing $\mathbb{G}$ (at given $\bm C$, $\gamma$ and $\cal T$), and in consequence also for calculating the likelihood according to Eq.~(\ref{eq:likelihood_2}), which depends on the inverse covariance matrix $\mathbb{G}^{-1}$.
+Note that this decomposition of the eigenvectors leads to a drastic decrease in computational complexity for diagonalizing $\mathbb{G}$ (at given $\bm C$, $\gamma$ and $\cal T$), and in consequence also for calculating the likelihood according to Eq.~\eqref{eq:likelihood_2}, which depends on the inverse covariance matrix $\mathbb{G}^{-1}$.
 Matrix $\mathbb{G}$ has linear dimension $LN$, so the numerical diagonalization or inversion takes time ${\cal O}( (L N)^3)$.
@@
 This observation is essential for inference, since we need to redo this calculation for many realizations of $\bm C$ and $\gamma$, in order to find the ones maximizing the likelihood given the data $\cal D$ and the tree $\cal T$.
-\martin{As is shown in App.
-XXX, this calculation simplifies even more when considering a fully balanced and homogeneous tree.
+As is shown in section \sref{sub:homogeneous_and_fully_balanced_tree}, this calculation simplifies even more when considering a fully balanced and homogeneous tree.
 In this case, the matrices $\bm G^a$ commute and can be diagonalized simultaneously and analytically for any value of $\rho^a$.
 
-For the case of arbitrary trees, }Eq.~(\ref{eq:likelihood_2}) can now be rewritten using the eigendecomposition of $\mathbb{G}$: 
+For the case of arbitrary trees, Eq.~\eqref{eq:likelihood_2} can now be rewritten using the eigen-decomposition of $\mathbb{G}$: 
 \begin{equation}
@@
 \end{equation}
-Eq.~(\ref{eq:likelihood_3}) expresses the likelihood as a function of $\vuka$, $\lambda_{ka}$ (resulting from the tree $\cal T$ and given $\rho^a$) and $\vsa$ (resulting from $\bm C$). 
-However, the definition of $\bm G^a$ Eq.~(\ref{eq:subG_def}) makes clear that its eigenmodes $\{\lambda_{ka}, \vuka \}$ depend only of the eigenvalues $\rho_a$ of $\iC$, on $\gamma$, as well as of the structure of the tree through the quantities $\Delta t_{ij}$, although this dependence cannot be analytically expressed in a simple manner.
-This means that the likelihood in equation~\ref{eq:likelihood_3} is in fact a function of $\{\rho_a, \vsa\}$, \emph{i.e.} the eigenmodes of $\iC$ and of the time scale parameter $\gamma$.
+Eq.~\eqref{eq:likelihood_3} expresses the likelihood as a function of $\vuka$, $\lambda_{ka}$ (resulting from the tree $\cal T$ and given $\rho^a$) and $\vsa$ (resulting from $\bm C$). 
+However, the definition of $\bm G^a$ in Eq.~\eqref{eq:subG_def} makes clear that its eigenmodes $\{\lambda_{ka}, \vuka \}$ depend only of the eigenvalues $\rho_a$ of $\iC$, on $\gamma$, as well as of the structure of the tree through the quantities $\Delta t_{ij}$, although this dependence cannot be analytically expressed in a simple manner.
+This means that the likelihood in equation~\eqref{eq:likelihood_3} is in fact a function of $\{\rho_a, \vsa\}$, \emph{i.e.} the eigenmodes of $\iC$, of the time scale parameter $\gamma$ and of the pairwise distances on the tree $\Delta t_{ij}$.
 
@@
 We also need to find the optimal time scale $\gamma$. 
-In Eq.~(\ref{eq:likelihood_3}), the likelihood is expressed as a function of $\gamma$ and $\{\rho_a, \vsa\}$, \ie~the eigenvalues and eigenvectors of $\iC$, either directly or through the quantities $\{\lambda_{ka}, \vuka \}$.
+In Eq.~(\eqref{eq:likelihood_3}), the likelihood is expressed as a function of $\gamma$ and $\{\rho_a, \vsa\}$, \ie~the eigenvalues and eigenvectors of $\iC$, either directly or through the quantities $\{\lambda_{ka}, \vuka \}$.
 We know attempt to maximize the likelihood with respect to the eigenmodes $\{\rho_a, \vsa\}$ and to the time scale $\gamma$.
@@
 To make explicit the dependences of eigenvalues and eigenvectors of the matrices $\bm G^a$ on these parameters, we introduce the notation $\vec{u}_k(\rho_a,\gamma) = \vuka$ and $\lambda_k(\rho_a,\gamma) = \lambda_{ka}$.
-Note that from the definition of $\bm G^a$ in Eq.~(\ref{eq:subG_def}), its eigenvalues and vectors depend only on the eigenvalues of $\iC$ and not on its eigenvectors. 
+Note that from the definition of $\bm G^a$ in Eq.~\eqref{eq:subG_def}, its eigenvalues and vectors depend only on the eigenvalues of $\iC$ and not on its eigenvectors. 
 In the same way, we will now write $\bm G(\rho_a,\gamma)$ instead of $\bm G^a$.
 
-The gradient of the likelihood is obtained by differentiating Eq.~(\ref{eq:likelihood_3}) with respect to the parameters of interest. 
+The gradient of the likelihood is obtained by differentiating Eq.~\eqref{eq:likelihood_3} with respect to the parameters of interest. 
 This gives us  three equations: 
@@
 For each eigenvalue $\rho_a$, we compute and diagonalize matrix $\bm{G}(\rho_a)$ to obtain its eigenmodes $\vec{u}_k(\rho_a)$ and $\lambda_k(\rho_a)$. 
-Using equations~\ref{eq:grad_eigval} and ~\ref{eq:grad_eigvec} and their equivalent form for $\gamma$, we also numerically compute their derivatives with respect to $\rho_a$ and $\gamma$. 
-This gives us all the quantities to estimate the gradient of $\mathcal{L}$ with respect to $\rho_a$ using equation~\ref{eq:gradlikelihood_1}.
+Using equations~\eqref{eq:grad_eigval} and ~\eqref{eq:grad_eigvec} and their equivalent form for $\gamma$, we also numerically compute their derivatives with respect to $\rho_a$ and $\gamma$. 
+This gives us all the quantities to estimate the gradient of $\mathcal{L}$ with respect to $\rho_a$ using equation~\eqref{eq:gradlikelihood_1}.
 
@@
 We also sample positive semi-definite coupling matrix $\bm{J}$ of size $L\times L$ with $L=4$ or $L=10$, with entries normally distributed with mean $\mu_J=0.8$ and $\sigma_J=0.2$.
-In order to perform interaction prediction, we randomly set to $0$ off-diagonal elements of $J$ with probability $0.7$, resulting in a coupling matrix with approximative density $0.3$.  
+
+In the case of statistical models of protein sequence, a major achievement is the ability of pairwise models to predict contacts in the three-dimensional structure of the protein from an inferred coupling matrix.
+In order to replicate this setting and to perform interaction prediction, we randomly set to $0$ off-diagonal elements of $J$ with probability $0.7$, resulting in a coupling matrix with approximative density $0.3$.
 Null elements of $J$ correspond to variables that do not interact, or to non-contacts in the case of an application to protein sequences.
@@
 As a means of comparison, we also compute the empirical covariance matrix $\bm{C}_{emp}$ as if leaf configurations were independent. 
-Figure \ref{fig:pears_L4} shows the Pearson correlation between the real covariance matrix $\bm{C}=\bm{J}^{-1}$ and the empirical or inferred ones in the $L=4$ case (similar figures for $L=10$ are in section \ref{sub:supplementary_figures}).
+Figure \ref{fig:pears_L4} shows the Pearson correlation between the real covariance matrix $\bm{C}=\bm{J}^{-1}$ and the empirical or inferred ones in the $L=4$ case (similar figures for $L=10$ are in section \ref{sub:supplementary_figures} of the smallest).
 As expected, both methods perform well in the large $\gamma$ limit with a correlation close to $1$, and worse in the low $\gamma$ limit. 
@@
 Figure \sref{fig:error_vs_nseqs} shows the $l2$-error between the true covariances and the empirical covariances computed from an \emph{i.i.d.} sample of size $N$. 
-The error is as expected increasing for decreasing values of $N$. 
+As expected the error is increasing for decreasing values of $N$. 
 We can use this as a calibration to express values of the $l2$-error in terms of an effective sample size.   
-For example, the error reached by $\bm{C}_{emp}$ at for $\gamma/\gamma_d\in[0.5, 1]$ and $L=4$ corresponds to the one obtained for an \emph{i.i.d.} sample of size $\sim 16$, whereas it corresponds to a sample of size $\sim 32-64$ for $\bm{C}_{max}$.
+For example, the error reached by $\bm{C}_{emp}$ for $\gamma/\gamma_d\in[0.5, 1]$ and $L=4$ corresponds to the one obtained for an \emph{i.i.d.} sample of size $\sim 16$, whereas it corresponds to a sample of size $\sim 32-64$ for $\bm{C}_{max}$.
 Thus, our correction is equivalent to doubling the number of effective samples.
@@
 	Therefore, we have to be able to simulate the OU process on a tree. 
-	In practice, given a rooted tree such as the one shown in figure~\ref{fig:sample_tree} of the main text, we want to sample a configuration $\vec{x}$ for every node in such a way that equation~\ref{eq:OUjoint} holds for every pair of nodes, the time $\Delta t$ then being the branch length connecting them.
+	In practice, given a rooted tree such as the one shown in figure~\ref{fig:sample_tree} of the main text, we want to sample a configuration $\vec{x}$ for every node in such a way that equation~\eqref{eq:OUjoint} holds for every pair of nodes, the time $\Delta t$ then being the branch length connecting them.
 \\
 	We use a simple methodology to achieve this. 
-	First, note that given an arbitrary configuration $\vec{x}_0$ and a time $\Delta t$, we can generate a new configuration $\vec{x}$ distributed according to the propagator in equation~\ref{eq:OUpropagator} by the transformation
+	First, note that given an arbitrary configuration $\vec{x}_0$ and a time $\Delta t$, we can generate a new configuration $\vec{x}$ distributed according to the propagator in equation~\eqref{eq:OUpropagator} by the transformation
 	\begin{equation}
@@
 	\end{equation}
-	where $\Lam$ and $\Sig$ are defined in equation~\ref{eq:def_lambda}, and $\vec{\eta}$ is a vector of uncorrelated variables with distributions $\mathcal{N}(0,1)$. 
-	Moreover, if $\vec{x_0}$ is distributed according to equation~\ref{eq:eq_distribution}, then $\vx$ and $\vx_0$ are distributed according to equation~\ref{eq:OUjoint}. 
-	Note that equation \ref{eq:OUsample} is quite different from the Langevin equation~\ref{eq:langevin}, even though they have similar forms. 
-	Whereas the Langevin equation describes the motion of $\vx$ in the potential $\bm{J}$, eq.~\ref{eq:OUsample} directly samples from the OU process. \\
-	Given an already sampled internal node in the tree, equation~\ref{eq:OUsample} allows to sample a configuration for each of its children. 
-	To sample the whole tree, we first sample the root note $\vx_0$ from the equilibrium distribution~\ref{eq:eq_distribution}. 
-	By recursive applications of~\ref{eq:OUsample}, we then simply work our way down the tree until all leaves are sampled.
+	where $\Lam$ and $\Sig$ are defined in equation~\eqref{eq:def_lambda}, and $\vec{\eta}$ is a vector of uncorrelated variables with distributions $\mathcal{N}(0,1)$. 
+	Moreover, if $\vec{x_0}$ is distributed according to equation~\eqref{eq:eq_distribution}, then $\vx$ and $\vx_0$ are distributed according to equation~\eqref{eq:OUjoint}. 
+	Note that equation \eqref{eq:OUsample} is quite different from the Langevin equation~\eqref{eq:langevin}, even though they have similar forms. 
+	While the Langevin equation describes the motion of $\vx$ in the potential $\bm{J}$, eq.~\eqref{eq:OUsample} directly samples from the OU process. \\
+	Given an already sampled internal node in the tree, equation~\eqref{eq:OUsample} allows to sample a configuration for each of its children. 
+	To sample the whole tree, we first sample the root note $\vx_0$ from the equilibrium distribution~\eqref{eq:eq_distribution}. 
+	By recursive applications of~\eqref{eq:OUsample}, we then simply work our way down the tree until all leaves are sampled.
 
@@
 
-	The probability distribution $P$ for the configurations of two leaves $\vx_i$ and  $\vx_j$  separated by time $\Delta t_{ij}$ is given by equation~\ref{eq:OUjoint} of the main text. 
+	The probability distribution $P$ for the configurations of two leaves $\vx_i$ and  $\vx_j$  separated by time $\Delta t_{ij}$ is given by equation~\eqref{eq:OUjoint} of the main text. 
 	With this distribution we can calculate analytically the average of the scalar product $\vx_i^{\,T}\vx_j$:
@@
 	\end{align*}
-	The covariance $\langle  x_i^{a} x_j^a \rangle_{P}$ between observations separated by a time $\Delta t_{ij}$ is given by equation~\ref{eq:pairwisecov} of the main text.
+	The covariance $\langle  x_i^{a} x_j^a \rangle_{P}$ between observations separated by a time $\Delta t_{ij}$ is given by equation~\eqref{eq:pairwisecov} of the main text.
 	Using this, we now have
@@
 	\end{align}
-	Having initialized the covariance matrix $\mathbf{C}$ at its empirical value, we know the values of all members of the r.h.s. of equation~\ref{eq:scalar_product} except that of $\gamma$. 
-	On the other hand, equivalent versions of equation~\ref{eq:scalar_product} can be written for all pairs of configurations $i$ and $j$. 
+	Having initialized the covariance matrix $\mathbf{C}$ at its empirical value, we know the values of all members of the r.h.s. of equation~\eqref{eq:scalar_product} except that of $\gamma$. 
+	On the other hand, equivalent versions of equation~\eqref{eq:scalar_product} can be written for all pairs of configurations $i$ and $j$. 
 	To find an initial value of $\gamma$ which is consistent with the data and the empirical covariance matrix, we search for one that best explains the observed scalar products between configurations. 
@@
 
- \subsection{ Homogeneous and fully balanced tree }
+ \subsection{Homogeneous and fully balanced tree}
+ \label{sub:homogeneous_and_fully_balanced_tree}
 
-Let assume that the tree is binary and completely homogeneous, all branches have the same length $\Delta t$.
-For instance the sequences covariance matrix for a binary tree with $K=2$ branching events and four leaves is
+Let's assume that the tree is binary, symmetric and completely homogeneous with all branches having the same length $\Delta t$. 
+As an example, the covariance matrix for such a tree with $K=2$ branching events and four leaves is
 \begin{equation} 
@@
 \end{array}\label{eq:G}
-\right)
+\right).
 \end{equation}
 
-The associatted matrix $\bm G^a=z(\rho_a,\gamma,\Delta t)$ defined by (14) became
+The associated matrix $\bm G^a=z(\rho_a,\gamma,\Delta t)$  defined in Eq.~\eqref{eq:G} becomes
 
@@
 \end{array}\label{eq:G_a}
-\right)
+\right).
 \end{equation}
 
-For hypergeometric matrices as \ref{eq:G_a} there is $K+1$ different eigenvalues  given by:
+For hyper-geometric matrices as \eqref{eq:G_a} of dimension $2^K$, there are $K+1$ different eigenvalues  given by:
 
@@
 
-where $\lambda_{K+1}\ge\lambda_{K}\cdots\ge\lambda_{1}$. For $k<K+1$ degenerancy of eigenvalues $\lambda_k$ is given by : $d_k=2^{K-k}$.
-
-The associated eigenvectors are independent of the parameter $\rho_a$ and reflects the events in the phylogenetic tree  where each eigenvector  $k$ captures the duplication events in the $(K+1 -k) 	st$ generation.
-Summarizing :
+where $\lambda_{K+1}\ge\lambda_{K}\cdots\ge\lambda_{1}$. For $k<K+1$, the degeneracy of eigenvalue $\lambda_k$ is $d_k=2^{K-k}$.  
+The associated eigenvectors are independent of the parameter $\rho_a$and reflect the events in the phylogenetic tree.
+Each eigenvector $\vec{u}_k$ captures the duplication events in the $(K+1 -k) st$ generation:
 
@@
 
-where $\Gamma(\vec{u}_k )$ represent $d_k$ combinations resulting of shift the block of length $Q$ without overlapping, generating all eigenvectors with the same corresponding eigenvalue $\lambda_k$.
-The eigenvectors are orthogonal to eachother, and can be  normalized and arranged horizontally into a matrix $U$.
+where $\Gamma(\vec{u}_k )$ represents the $d_k$ combinations obtained by shifting the block of length $Q$, generating all eigenvectors corresponding to the eigenvalue $\lambda_k$.
+The eigenvectors are orthogonal to each other, and can be  normalized and arranged horizontally into a matrix $U$.
 
-Then to compute the gradient of the likelihood we only need  derivatives of $\lambda_k(\rho_a,\gamma)$  respect to $\rho_a$ and $\gamma$, which can be obtained from direct derivation of expression \ref{eq:lambda_hyper}.
+To compute the gradient of the likelihood, we need  derivatives of $\lambda_k(\rho_a,\gamma)$ with respect to $\rho_a$ and $\gamma$, which can be directly obtained from expression \eqref{eq:lambda_hyper}.
 
@@
 
-The method implementation was done using   NLopt package (see reference).
+The method implementation was done using   NLopt package (see reference) \pierre{Add reference?}.
 

### 2021-01-26 main.tex
discussion v1

@@
 
-This modeling approach is \emph{a priori} limited to continuous traits, but could potentially be used for protein sequences combined with a continuous-variable approximation, that has successfully been used in the past \cite{jones_psicov_2012,baldassi_fast_2014}.
+This modeling approach is \emph{a priori} limited to continuous traits, but could potentially be used for protein sequences combined with a continuous-variable approximation, that has successfully been used in the past \cite{jones_psicov_2012,barton_large_2014,baldassi_fast_2014}.
 In this context, the equilibrium distribution reached by the OU process represents the probability distribution given by the DCA method, which can be used to predict non-trivial structural contacts in the protein fold, effects of amino-acid mutations or even designing novel functional sequences \cite{morcos_direct-coupling_2011, figliuzzi_coevolutionary_2016, russ_evolution-based_2020}.
@@
 
-What is this useful for? 
--
+In this work, we proposed a method for inferring parameters of an Ornstein-Uhlenbeck process using data that is correlated through an evolutionary tree. 
+We kept a very general setting in which data can in principle represent any set of continuous phenotypic traits or potentially discrete sequences if a continuous approximation is made. 
+As such, our approach is only methodological, and does not directly investigate any particular application.
+
+We showed that due to the Gaussian and time reversible nature of the OU process, it is possible to write the joint covariance matrix of all data vectors in a simple way. 
+The resulting matrix $\mathbb{G}$ consists of block entries that represent covariances between pairs of leaves.
+The dependence of these blocks on the coupling matrix $\bm{J}$ characterizing the OU process and on the tree structure can be written explicitly. 
+Interestingly, $\mathbb{G}$ only depends on the tree structure through the pairwise branch length $\Delta t_{ij}$ separating leaves.
+
+We then proposed a way to compute the likelihood of the data given the tree and the parameters of the OU process, namely the coupling matrix $\mathbf{J}$ and timescale $\gamma$.
+This method relies on computing the eigenvalues and vectors of the joint covariance matrix in an efficient manner. 
+Indeed, it is possible to separate this calculation in two steps: the first in which we perform the eigen-decomposition of the matrix $\mathbf{J}$, and the second in which we compute eigenvalues and vectors of matrices $\mathbf{G}^a$ that embed the tree structure.
+This reduces the computational complexity from $\mathcal{O}(L^3N^3)$ for a naive inversion of $\mathbb{G}$ to $\mathcal{O}(L^3) + \mathcal{O}(LN^3)$. 
+We also show that this method can be used to compute the gradient of the likelihood with respect to parameters with the same complexity.
+This makes the problem of inferring $\mathbf{J}$ amenable to maximum likelihood methods using a gradient ascent approach.
+
+Finally, we showed that this process gives expected results on simulated data, with a more accurate reconstruction of parameters than if empirical estimation was performed. 
+These simulations highlight the fact that this method is only useful in the intermediate regime of phylogenetic correlations. 
+If $\gamma$ is too high, correlation of data points through the tree is weak and an empirical estimation performs well. 
+On the other hand, a very low $\gamma$ results in strong phylogenetic biases that make recovering $\mathbf{J}$ impossible.
+However, in the intermediate regime, our re-construction of $\mathbf{J}$ that takes the tree into account results in clear benefits over an empirical estimation.
+
+A limitation of our approach remains the long computational time. 
+Even with the efficient computation of the gradient, it was necessary to use small system sizes, $L=10$ at most, to repeat our inference process many times with simulated data in a reasonable time. 
+For this reason, the framework proposed here is limited to a small number of variables. 
+In this respect, it is interesting to note that a different manner of computing the likelihood developed in \cite{mitov_fast_2020} and based on Gaussian integrations on every branch of the tree results in an asymptotic complexity of $\mathcal{O}(NL^3)$.
+
+Although our method can in principle be used for any set of traits, a major motivation in developing it is its potential application to model of proteins sequences. 
+Several results in the last years have shown that selection forces shaping the evolution of protein sequences are well described by a pairwise potential. 
+The estimation of this potential is performed using homologous sequences, and is therefore biased by the phylogenetic relations between these sequences.  
+Results presented here are a first step in disentangling effects due to phylogeny from effects due to selection in a principled way.
+
+However, there remain several challenges in using this framework for protein sequences. 
+First, the computational power required to process actual sequences is much larger than what was needed for the small simulated systems presented here. 
+As an example, a protein of length $L=100$ will be represented by $q\times 100=2000$ Gaussian variables, where $q=20$ is the number of amino acids. 
+This is of course much larger than the $L=10$ system used as an example to test our approach.
+
+A second question is the capacity of a continuous variable approximation, necessary when using Ornstein-Uhlenbeck dynamics, to represent dynamical properties of the landscape protein sequences evolve in.
+This type of approximation has been successfully used before, but in quite different contexts \cite{jones_psicov_2012,barton_large_2014,baldassi_fast_2014}. 
+Its use in the context of modelling the evolutionary dynamics of protein sequences remains an open question.
 

### 2021-02-05 main.tex
Martin&Edwin changes

@@
 \begin{abstract}
-	The Inverse problem of Statistical Physics infers maximum-entropy  models  compatibles with a corresponding set of empirical averages extracted from a high-dimensional dataset of equilibrium configurations.
-Practical interest extended these methods to non-equilibrium data  like  time series, where detailed balance does not hold.
-However, in several applications, data samples  result from evolutionary processes and the relation between configurations is conditioned by a hierarchical structure of the population.
-This makes sample statistics a superposition of  signals coming from both: internal configuration correlations and relatedness correlations produced by common evolutionary  history, typically represented by a tree.
-How to disentangle both sources of correlation for a better parameter inference is an open question to explore.
-Here we propose an approach  where the  evolutive process through a phylogenetic tree is described by  a  multivariate Ornstein-Uhlenbeck dynamics characterized by  gaussian propagator and stationary distributions.
-Inference of these distributions from  phylogenetic non-independent samples is solved in a Bayesian framework.
-This procedure can be  extended to discrete variables by using a binary representation of data and approximating  binary sates by continuous variables.
-Our method proposes a possible way for  a better estimation of both, equilibrium and dynamic parameters, making  applications more accurate.
+The inverse statistical physics infers maximum-entropy models compatible with a set of empirical averages estimated from a high-dimensional dataset of independently distributed equilibrium configurations.
+However, in several applications, data result from stochastic evolutionary processes, and the relation between configurations is characterized by a hierarchical correlation structure, typically represented by a tree.
+In turn, empirical averages of observables superpose intrinsic signals related to the equilibrium distribution of the studied system, and spurious historical (or phylogenetic) signals characterizing the specific correlated data-generating process.
+The naive application of maximum-entropy techniques therefore leads to systematic biases and an effective reduction of the sample size.
+To advance on the currently open task of extracting intrinsic signals from correlated data, we study a system described by a multivariate Ornstein-Uhlenbeck process defined on a finite tree.
+Using a Bayesian framework, we can disentangle covariances in the data, which are result of their multivariate Gaussian equilibrium distribution, from those resulting from the historical correlations.
+Our approach leads to a clear gain in accuracy in the inferred equilibrium distribution, which corresponds to an effective two- to fourfold increase in sample size.
 \end{abstract}
@@
 \label{sec:int}
+
+With the emergence of large, high-dimensional datasets for complex systems across disciplines, methods of {\it inverse statistical physics} have seen rapidly growing interest during the last years.
+In the most standard setting , the data provide observational samples of the "microscopic" degrees of freedom of the system under study -- this can be biological sequences, firing patterns of neurons, individuals in animal or human groups, stock prices etc.
+Within a static modeling approach, frequently based on the maximum-entropy approach, data $\vec x$ are assumed to be generated independently from some unknown probability distribution $P(\vec x)$.
+This distribution describes the underlying interaction patterns between the observed degrees of freedom, and has to be learned from data to unveil the rules governing the system.
+In some, more rare cases, data correspond to observed time series , but the theoretical and algorithmic development is much less advanced than in the case of independent static data, assuming, e.g., that measurements are taken at the microscopic time scale.
+
+One of the biggest application areas of inverse statistical mechanics is the modeling of biological sequences, and we will use them to motivate our otherwise theoretical-methodological study.
+Typical datasets are so called homologous protein families, collecting ensembles of proteins of common evolutionary ancestry.
+Despite their strong divergence in amino-acid sequences, the proteins in such families have highly similar biological functions and three-dimensional fold structures.
+The different sequences can thereby, in a statistical-physics perspective, be seen as microscopically distinct solutions for the same macroscopic behavior.
+
+In such protein families, which are typically represented by large multiple-sequence alignments, at least two complementary types of biological information are contained:
+\begin{itemize}
+    \item {\it Phylogenetic information:} the distances between sequences carry information about the evolutionary time since the common ancestor.
+Using phylogenetic methods we may reconstruct the evolutionary history of a protein family (and of the species carrying the proteins), and also the sequences of the ancestral proteins.
+    \item {\it Coevolutionary information} considers the correlated evolution of positions in proteins, frequently related to contacts in the folded protein, even for positions distant along the sequence.
+This information has been massively used in protein-structure prediction, most recently in combination with structurally supervised deep learning, but also to infer mutational landscapes or networks of interacting proteins.
+\end{itemize}
+
+These two types of information are contained in two complementary features of the data set: phylogenetic inference is based on the comparative analysis of the rows of the best, i.e., of entire proteins; and coevolutionary inference is based on the comparative analysis of MSA columns, which describe the amino-acid usage of the individual positions across proteins.
+
+For methodological reasons, each type of inference assumes the absence of the other information: phylogenetic inference assumes independent evolution of all positions; coevolutionary inference assumes that proteins form an independently and identically distributed sample of the model to be inferred.
+
+It has turned out to be hard to combine the two types of information in a single approach, even if the existence of, e.g, phylogenetic correlations between sequences influences the results of coevolutionary analysis.
+Here we propose to advance on the question of phylogeny-aware coevolutionary inference using the simplest nontrivial model setting: correlated multivariate Ornstein-Uhlenbeck (OU) processes.
 
@@
 A second possibility, that we have found faster in practice, is to express the matrix of the $\{\vsa\}$ as the exponential of a skew-symetric matrix, see section \ref{sub:parametrizations_of_eigenvectors} of the SM.
-However, this parametrization does not allow a simple analytical expression of the gradient of the likelihood, and we use it along with automatic differentiation.
-\pierre{Edwin, can you add a reference to the package you used in the .bib and here?}
+However, this parametrization does not allow a simple analytical expression of the gradient of the likelihood, and we use it along with automatic differentiation \cite{Zygote.jl-2018}. 
 For this reason, we use the Eulerian angles below to express the gradient of the likelihood.
@@
 In the case of statistical models of protein sequence, a major achievement is the ability of pairwise models to predict contacts in the three-dimensional structure of the protein from an inferred coupling matrix.
-In order to replicate this setting and to perform interaction prediction, we randomly set to $0$ off-diagonal elements of $J$ with probability $0.7$, resulting in a coupling matrix with approximative density $0.3$.
-Null elements of $J$ correspond to variables that do not interact, or to non-contacts in the case of an application to protein sequences.
+In order to replicate this setting and to perform interaction prediction, we randomly set to $0$ off-diagonal elements of $J$ with probability $0.7$, resulting in a sparsified coupling matrix of approximate density $0.3$.
+Zero elements of $J$ correspond to variables that do not interact, in analogy to non-contacts in the case of an application to protein sequences.
 
@@
 As a means of comparison, we also compute the empirical covariance matrix $\bm{C}_{emp}$ as if leaf configurations were independent. 
-Figure \ref{fig:pears_L4} shows the Pearson correlation between the real covariance matrix $\bm{C}=\bm{J}^{-1}$ and the empirical or inferred ones in the $L=4$ case (similar figures for $L=10$ are in section \ref{sub:supplementary_figures} of the smallest).
+Fig.~\ref{fig:pears_L4} shows the Pearson correlation between the real covariance matrix $\bm{C}=\bm{J}^{-1}$ and the empirical or inferred ones in the $L=4$ case (similar figures for $L=10$ are in Appendix \ref{sub:supplementary_figures}).
 As expected, both methods perform well in the large $\gamma$ limit with a correlation close to $1$, and worse in the low $\gamma$ limit. 
-In this latter case, correlations due to phylogeny are too strong for our maximum likelihood method to pick up signal, and both methods perform as poorly. 
+In this latter case, correlations due to phylogeny are too strong for our maximum likelihood method to pick up signal, and both methods perform equally poorly. 
 However, there exists an intermediate regime where $\bm{C}_{max}$ is much closer to the actual correlation than  $\bm{C}_{emp}$. 
-In figure \ref{fig:error_1_L4}, we plot the relative $l2$-error between either covariance matrices in the left panel or coupling matrices in the right panel. 
+In Fig.~\ref{fig:error_1_L4}, we plot the relative $l2$-error between either covariance matrices in the left panel or coupling matrices in the right panel. 
 In both cases, our maximum-likelihood method results in a consistent improvement over the empirical estimator. 
@@
 	\centering
-	\includegraphics[keepaspectratio=true,width=0.45\textwidth]{Figures/corr_C_L4_balanced_tree_100_rep.pdf}
+	\includegraphics[keepaspectratio=true,width=0.5\textwidth]{Figures/corr_C_L4_balanced_tree_100_rep.pdf}
 	\caption{ Pearson correlation between empirical /maximum-likelihood   covariance matrices and the true covariance matrix.
@@
 \begin{figure*}[!htb]
-		\begin{subfigure}{}
-			\centering\includegraphics[keepaspectratio=true,width=0.45\textwidth]{Figures/epsilon_error_C_L4_balanced_tree_100.pdf}
-		\end{subfigure}
+	%	\begin{subfigure}{}
+			\centering\includegraphics[keepaspectratio=true,width=1.0\textwidth]{Figures/epsilon_error_L4_balanced_tree_100.pdf}
+	%	\end{subfigure}
 		\hspace{1mm}
-		\begin{subfigure}{}
-			\centering\includegraphics[keepaspectratio=true,width=0.45\textwidth]{Figures/epsilon_error_J_L4_balanced_tree_100.pdf}
-		\end{subfigure}
+	%	\begin{subfigure}{}
+	%		\centering\includegraphics[keepaspectratio=true,width=0.45\textwidth]{Figures/epsilon_error_J_L4_balanced_tree_100.pdf}
+	%	\end{subfigure}
 	\caption{\textbf{Left:}Relative $l2$-error between empirical or maximum-likelihood  covariance matrices and the true covariance matrix.
@@
 
-An interesting way to illustrate the benefits of reconstructing the covariance matrix using knowledge of the tree is to evaluate the effective number of samples that is gained by doing so. 
-Figure \sref{fig:error_vs_nseqs} shows the $l2$-error between the true covariances and the empirical covariances computed from an \emph{i.i.d.} sample of size $N$. 
-As expected the error is increasing for decreasing values of $N$. 
-We can use this as a calibration to express values of the $l2$-error in terms of an effective sample size.   
+An interesting way to illustrate the benefits of reconstructing the covariance matrix using knowledge of the tree is to evaluate the gain in {\em effective sample size}.
+Intuitively, the use of correlated samples reduces the information contained in the data, as compared to an equally large dataset of \emph{i.i.d.} configurations.
+It is therefore interesting to compare the accuracy of our inferences with the accuracy obtained on smaller but \emph{i.i.d.} samples.
+To do so, we report in Fig.~\sref{fig:error_vs_nseqs} the $l2$-error between true and empirical covariances computed from a \emph{i.i.d.} samples of variable sizes $N$.
+As expected,  the error increases with decreasing values of $N$.
+We can use this in turn to express values of the $l2$-error in correlated samples in terms of effective \emph{i.i.d.} sample sizes.   
 For example, the error reached by $\bm{C}_{emp}$ for $\gamma/\gamma_d\in[0.5, 1]$ and $L=4$ corresponds to the one obtained for an \emph{i.i.d.} sample of size $\sim 16$, whereas it corresponds to a sample of size $\sim 32-64$ for $\bm{C}_{max}$.
-Thus, our correction is equivalent to doubling the number of effective samples.
+Thus, our correction is equivalent to increasing by a factor 2-4 the number of effective samples.
 
-Finally, we assess the performance of our method in improving the prediction of interactions between gaussian variables $\{x_a\}$. 
+Finally, we assess the performance of our method in improving the prediction of the network of interactions between the Gaussian variables $\{x_a\}$. 
 We consider that two variables $x_a$ and $x_b$ interact if the corresponding entry in the coupling matrix is non-zero, that is $J_{ab}\neq 0$. 
-We predict these interactions by taking the largest $n$ elements (in absolute value) of the inferred coupling matrix, resulting in $n$ predictions. 
-The fraction of these $n$ predictions that correspond to non-zero entries in the true matrix defines the positive predictive value (PPV). 
+Using the data, we predict these interactions by taking the largest $n$ elements (in absolute value) of the inferred coupling matrix, resulting in $n$ predictions. 
+The fraction TP$/n$ of these $n$ predictions that correspond to non-zero entries in the true matrix (TP = true positives) defines the positive predictive value (PPV). 
 This problem is equivalent to the one of predicting contacts in a protein structure
 
-Figure $\ref{fig:PPV_L4}$ shows the PPV as a function of the number of predictions for different values of $\gamma$ and $L=4$ (see Fig. \sref{fig:PPV_L10} for the $L=10$ case). 
+Fig.~$\ref{fig:PPV_L4}$ shows the PPV as a function of the number of predictions for different values of $\gamma$ and $L=4$ (see Fig. \sref{fig:PPV_L10} for the $L=10$ case). 
 In this case, the coupling matrix only has $6$ independent non-diagonal elements, and only 6 predictions can be made. 
@@
 	\centering
-	\includegraphics[keepaspectratio=true,width=1.0\textwidth]{Figures/PPVs_L4.png}
+	\includegraphics[keepaspectratio=true,width=1.0\textwidth]{Figures/PPVs_L4.pdf}
 	\caption{Quality of prediction of interactions for  different values of $\gamma$ and system size $L=4$.
@@
 We kept a very general setting in which data can in principle represent any set of continuous phenotypic traits or potentially discrete sequences if a continuous approximation is made. 
-As such, our approach is only methodological, and does not directly investigate any particular application.
+As such, our approach is purely methodological, and does not directly investigate any particular application.
 
@@
 The dependence of these blocks on the coupling matrix $\bm{J}$ characterizing the OU process and on the tree structure can be written explicitly. 
-Interestingly, $\mathbb{G}$ only depends on the tree structure through the pairwise branch length $\Delta t_{ij}$ separating leaves.
+Interestingly, $\mathbb{G}$ only depends on the tree structure through the pairwise path length $\Delta t_{ij}$ separating leaves along the tree.
 
@@
 
-Finally, we showed that this process gives expected results on simulated data, with a more accurate reconstruction of parameters than if empirical estimation was performed. 
+Finally, we showed that this process gives encouraging results on simulated data, with a more accurate reconstruction of parameters than if empirical estimation was performed. 
 These simulations highlight the fact that this method is only useful in the intermediate regime of phylogenetic correlations. 
-If $\gamma$ is too high, correlation of data points through the tree is weak and an empirical estimation performs well. 
-On the other hand, a very low $\gamma$ results in strong phylogenetic biases that make recovering $\mathbf{J}$ impossible.
-However, in the intermediate regime, our re-construction of $\mathbf{J}$ that takes the tree into account results in clear benefits over an empirical estimation.
+If the timescale $\gamma$ characterizing the branch lengths of the tree is too large, correlation of data points through the tree is weak and an empirical estimation performs well. 
+On the other hand, a very low $\gamma$ results in strong phylogenetic biases that make recovering $\mathbf{J}$ impossible, basically due to a strong reduction of the information in a too redundant dataset.
+However, in an intermediate regime where intrinsic and historical correlations in the dataset coexist, our tree-aware re-construction of $\mathbf{J}$ results in clear benefits over a tree-unaware empirical estimation.
 
@@
 
-\textbf{Acknowledgments:}
+\textbf{Acknowledgments:} We acknowledge interesting discussions with Roberto Mulet.
+PBC and MW acknowlege the hospitality of the Department of  Theoretical  Physics of University  of  Havana, where part of this work was done.
+Our work was partially funded by the EU H2020 Research and Innovation Programme MSCA-RISE-2016 under Grant
+Agreement No. 734439 InferNet.
 
@@
 
-The method implementation was done using   NLopt package (see reference) \pierre{Add reference?}.
+The method implementation was done using   NLopt package \cite{NLopt} .
 
@@
 	\centering
-	\includegraphics[keepaspectratio=true,width=1.0\textwidth]{Figures/error_Ciid_vs_numb_of_seq.pdf}
+	\includegraphics[keepaspectratio=true,width=0.5\textwidth]{Figures/error_Ciid_vs_numb_of_seq.pdf}
 	\caption{Relative $l2$-error between the empirical covariance matrix calculated from an \emph{i.i.d.} sample and the true covariance matrix, for system sizes $L=4$ and $L=10$.
@@
 	\centering
-	\includegraphics[keepaspectratio=true,width=0.45\textwidth]{Figures/scatter_gamma_L4_100.pdf}
+	\includegraphics[keepaspectratio=true,width=0.5\textwidth]{Figures/scatter_gamma_L4_100.pdf}
 	\caption{Inferred $\gamma$ values as a function of real $\gamma$, for system size $L=4$.
@@
 	\centering
-	\includegraphics[keepaspectratio=true,width=0.55\textwidth]{Figures/Pearson_correlation_rate_full_range.pdf}
+	\includegraphics[keepaspectratio=true,width=0.5\textwidth]{Figures/Pearson_correlation_rate_full_range.pdf}
 	\caption{  Ratio between Pearson correlation of the  maximum-likelihood  and true covariance  matrix to the Pearson correlation of the  empirical and true covariances matrices for the two system sizes $L=4$ and $L=10$.}
@@
 	\centering
-	\includegraphics[keepaspectratio=true,width=0.65\textwidth]{Figures/corr_C_L10_balanced_tree_100_rep.pdf}
+	\includegraphics[keepaspectratio=true,width=0.5\textwidth]{Figures/corr_C_L10_balanced_tree_100_rep.pdf}
 	\caption{ Pearson correlation between empirical /maximum-likelihood   covariance matrices and the true covariance matrix, the inset plot represent the ratio between the person correlation for the  maximum-likelihood covariance matrix and the one for the empirical covariance matrix.}
@@
 	\centering
-	\includegraphics[keepaspectratio=true,width=0.65\textwidth]{Figures/scatter_gamma_L10_100.pdf}
+	\includegraphics[keepaspectratio=true,width=0.5\textwidth]{Figures/scatter_gamma_L10_100.pdf}
 	\caption{Inferred $\gamma$ values as a function of real $\gamma$, for system size $L=10$.
@@
  \begin{figure*}[!htb]
- 		\begin{subfigure}{}
- 			\centering\includegraphics[keepaspectratio=true,width=0.45\textwidth]{Figures/epsilon_error_C_L10_balanced_tree_100.pdf}
- 		\end{subfigure}
+ 	%	\begin{subfigure}{}
+ 			\centering\includegraphics[keepaspectratio=true,width=1.0\textwidth]{Figures/epsilon_error_L10_balanced_tree_100.pdf}
+ %		\end{subfigure}
  		\hspace{1mm}
- 		\begin{subfigure}{}
- 			\centering\includegraphics[keepaspectratio=true,width=0.45\textwidth]{Figures/epsilon_error_J_L10_balanced_tree_100.pdf}
- 		\end{subfigure}
+ %		\begin{subfigure}{}
+ %			\centering\includegraphics[keepaspectratio=true,width=0.45\textwidth]{Figures/epsilon_error_J_L10_balanced_tree_100.pdf}
+ %		\end{subfigure}
  	\caption{\textbf{Left:} Relative $l2$-error between empirical or maximum-likelihood covariance matrices and the true covariance matrix.
@@
 	\centering
-	\includegraphics[keepaspectratio=true,width=1.0\textwidth]{Figures/PPVs_L10.png}
+	\includegraphics[keepaspectratio=true,width=1.0\textwidth]{Figures/PPVs_L10.pdf}
 	\caption{Quality of prediction of interactions for  different values of $\gamma$ and system size $L=4$.

### 2021-02-06 main.tex
Change abstract

@@
 \begin{abstract}
-The inverse statistical physics infers maximum-entropy models compatible with a set of empirical averages estimated from a high-dimensional dataset of independently distributed equilibrium configurations.
-However, in several applications, data result from stochastic evolutionary processes, and the relation between configurations is characterized by a hierarchical correlation structure, typically represented by a tree.
+Inverse statistical physics aims at inferring models compatible with a set of empirical averages estimated from a high-dimensional dataset of independently distributed equilibrium configurations.
+However, in several applications such as biology, data result from stochastic evolutionary processes, and the relation between configurations is characterized by a hierarchical correlation structure, typically represented by a tree.
 In turn, empirical averages of observables superpose intrinsic signals related to the equilibrium distribution of the studied system, and spurious historical (or phylogenetic) signals characterizing the specific correlated data-generating process.
-The naive application of maximum-entropy techniques therefore leads to systematic biases and an effective reduction of the sample size.
+The naive application of inverse statistical physics techniques therefore leads to systematic biases and an effective reduction of the sample size.
 To advance on the currently open task of extracting intrinsic signals from correlated data, we study a system described by a multivariate Ornstein-Uhlenbeck process defined on a finite tree.
-Using a Bayesian framework, we can disentangle covariances in the data, which are result of their multivariate Gaussian equilibrium distribution, from those resulting from the historical correlations.
+Using a Bayesian framework, we can disentangle covariances in the data, which are a result of their multivariate Gaussian equilibrium distribution, from those resulting from the historical correlations.
 Our approach leads to a clear gain in accuracy in the inferred equilibrium distribution, which corresponds to an effective two- to fourfold increase in sample size.
@@
 \label{sec:int}
+% Keep
+With the emergence of large, high-dimensional datasets for complex systems across disciplines, methods of {\it inverse statistical physics} have seen rapidly growing interest during the last years. 
+In the most standard setting, the data provide observational samples of the ``microscopic''degrees of freedom of the system under study -- this can be biological sequences, firing patterns of neurons, individuals in animal or human groups, stock prices etc...
+Within a static modeling approach, frequently based on the maximum-entropy principle, data $\vec x$ are assumed to be generated independently from some unknown probability distribution $P(\vec x)$. 
+This distribution describes the underlying interaction patterns between the observed degrees of freedom, and has to be learned from data to unveil the rules governing the system. 
+In more rare cases where data correspond to observed time series, theoretical and algorithmic development is much less advanced than in the case of independent static data.
+%, assuming, e.g., that measurements are taken at the microscopic time scale.
 
-With the emergence of large, high-dimensional datasets for complex systems across disciplines, methods of {\it inverse statistical physics} have seen rapidly growing interest during the last years.
-In the most standard setting , the data provide observational samples of the "microscopic" degrees of freedom of the system under study -- this can be biological sequences, firing patterns of neurons, individuals in animal or human groups, stock prices etc.
-Within a static modeling approach, frequently based on the maximum-entropy approach, data $\vec x$ are assumed to be generated independently from some unknown probability distribution $P(\vec x)$.
-This distribution describes the underlying interaction patterns between the observed degrees of freedom, and has to be learned from data to unveil the rules governing the system.
-In some, more rare cases, data correspond to observed time series , but the theoretical and algorithmic development is much less advanced than in the case of independent static data, assuming, e.g., that measurements are taken at the microscopic time scale.
+% Merge with my first par, don't be too specific with protein sequences
+One of the biggest application areas of inverse statistical mechanics is the modeling of biological processes.
+These applications are fuelled by the large amount of available data resulting from the impressive progress in experimental techniques in biology. 
+This is especially visible in the case of biological sequences,  with databases now harboring a vast amount of high-quality DNA or protein sequences \cite{sayers_genbank_2019,uniprotconsortium_uniprot_2018}.
+A common idea in this context is that it is possible to use characteristics of genes or organisms related by a common ancestry -- called \emph{homologous} -- to construct models of the selection acting on them.
+A successful example in this regard is the representation of protein sequences by probabilistic models in the so-called DCA method \cite{levy_potts_2017,cocco_inverse_2018}. 
+We will use this type of application to motivate our otherwise purely methodological study.
 
-One of the biggest application areas of inverse statistical mechanics is the modeling of biological sequences, and we will use them to motivate our otherwise theoretical-methodological study.
-Typical datasets are so called homologous protein families, collecting ensembles of proteins of common evolutionary ancestry.
-Despite their strong divergence in amino-acid sequences, the proteins in such families have highly similar biological functions and three-dimensional fold structures.
-The different sequences can thereby, in a statistical-physics perspective, be seen as microscopically distinct solutions for the same macroscopic behavior.
-
-In such protein families, which are typically represented by large multiple-sequence alignments, at least two complementary types of biological information are contained:
+% Keep (shorten?)
+When modeling homologous genes or organisms, one has to distinguish between two complementary types of information: 
 \begin{itemize}
-    \item {\it Phylogenetic information:} the distances between sequences carry information about the evolutionary time since the common ancestor.
-Using phylogenetic methods we may reconstruct the evolutionary history of a protein family (and of the species carrying the proteins), and also the sequences of the ancestral proteins.
-    \item {\it Coevolutionary information} considers the correlated evolution of positions in proteins, frequently related to contacts in the folded protein, even for positions distant along the sequence.
-This information has been massively used in protein-structure prediction, most recently in combination with structurally supervised deep learning, but also to infer mutational landscapes or networks of interacting proteins.
+    \item {\it Phylogenetic information:} the distances between gene sequences or traits in organisms carry information about the evolutionary time since their common ancestor. 
+    Using phylogenetic methods we may reconstruct the evolutionary history of a set of genes or species, and also the sequences or traits of their ancestors. 
+    \item {\it Co-evolutionary information:} positions in a sequence or traits in organisms typically do not evolve independently, but rather in a correlated way. 
+    This co-evolution carries important information about the selection forces acting on evolving entities.  
+    This fact has been extensively studied in the case of protein sequences, and used to predict structure, mutational landscapes or networks of interacting proteins.  
 \end{itemize}
 
-These two types of information are contained in two complementary features of the data set: phylogenetic inference is based on the comparative analysis of the rows of the best, i.e., of entire proteins; and coevolutionary inference is based on the comparative analysis of MSA columns, which describe the amino-acid usage of the individual positions across proteins.
+These two types of information are contained in two complementary features of the data: phylogenetic inference is based on the comparative analysis of different species or genes, while co-evolutionary information is contained in the correlation of different trait values in the same gene or organism.
 
-For methodological reasons, each type of inference assumes the absence of the other information: phylogenetic inference assumes independent evolution of all positions; coevolutionary inference assumes that proteins form an independently and identically distributed sample of the model to be inferred.
+Models built using biological data usually ignore one of the two types of information. 
+For instance, inference of phylogenies typically assumes that all positions in a sequence evolve independently, while co-evolutionary models of proteins assume that their sequences are independently distributed.
+This choice is motivated by the fact that taking the two types of correlations into account, \emph{i.e.} through time with phylogeny and accross trait values for co-evolution, results in very hard inference procedures.
+However, this can lead to biases in the model parameters: it has for instance been shown that phylogenetic relations between protein sequences induce non-trivial correlations that are not related to protein function \cite{qin_power_2018}.
 
-It has turned out to be hard to combine the two types of information in a single approach, even if the existence of, e.g, phylogenetic correlations between sequences influences the results of coevolutionary analysis.
-Here we propose to advance on the question of phylogeny-aware coevolutionary inference using the simplest nontrivial model setting: correlated multivariate Ornstein-Uhlenbeck (OU) processes.
-
-In the last decades, biology has seen impressive progress in experimental techniques which has resulted in a large increase of available data. 
-This is especially visible in the case of biological sequences,  with databases now harboring a vast amount of high-quality DNA or protein sequences \cite{sayers_genbank_2019,uniprotconsortium_uniprot_2018}.
-This has in turn fuelled a development of quantitative methods to model forces guiding the evolution of sequences or other biological traits [REFS/EXAMPLES?]. 
-A common idea in this context is that it is possible to use characteristics of homologous genes or organisms to construct models of the selection acting on them. 
-A successful example in this regard is the representation of protein sequences by probabilistic models in the so-called DCA method \cite{levy_potts_2017,cocco_inverse_2018}.
-
-Models built in this way usually assume that biological characteristics of related organisms are evolving under the same constraints but are independent from each other in the statistical sense. 
-This allows simple formulation for model learning. 
-However, the phylogenetic relations between evolving entities are in direct contradiction with this simplifying hypothesis, and ignoring them can result in biases when learning models. 
-For instance, it has been shown that phylogenetic relations between protein sequences induce non-trivial correlations that are not related to protein function \cite{qin_power_2018}.
-
-In order to correct for these biases, it is necessary to disentangle the effects due to selection forces and those due to phylogeny. 
-However, this becomes a hard problem when dealing with complex models where traits under study are not independent from each other. 
-Accounting for these biases then requires one to understand how interdependent traits under selection evolve along a tree. 
+Accounting for both co-evolution and phylogeny in models requires one to understand how interdependent traits under selection evolve along a tree. 
 A historically well-known way to represent such processes is to use Ornstein-Uhlenbeck dynamics (OU), which models traits as a Gaussian vector evolving in a quadratic potential that represents selection forces \cite{uhlenbeck_theory_1930, felsenstein_phylogenies_1988, hansen_stabilizing_1997}.
-This method is commonly used in the field of phylogenetic comparative methods (PCM) \cite{bartoszek_phylogenetic_2012,mitov_fast_2020}.
-
+OU processes are commonly used in the field of phylogenetic comparative methods (PCM) \cite{bartoszek_phylogenetic_2012,mitov_fast_2020}. 
 This modeling approach is \emph{a priori} limited to continuous traits, but could potentially be used for protein sequences combined with a continuous-variable approximation, that has successfully been used in the past \cite{jones_psicov_2012,barton_large_2014,baldassi_fast_2014}.

### 2021-02-09 main.tex
Edit comments
martin changes

@@
 \label{sec:int}
-% Keep
-With the emergence of large, high-dimensional datasets for complex systems across disciplines, methods of {\it inverse statistical physics} have seen rapidly growing interest during the last years. 
-In the most standard setting, the data provide observational samples of the ``microscopic''degrees of freedom of the system under study -- this can be biological sequences, firing patterns of neurons, individuals in animal or human groups, stock prices etc...
-Within a static modeling approach, frequently based on the maximum-entropy principle, data $\vec x$ are assumed to be generated independently from some unknown probability distribution $P(\vec x)$. 
+
+With the emergence of large, high-dimensional datasets for complex systems across disciplines, methods of {\it inverse statistical physics} have seen rapidly growing interest during the last years \cite{Inverse_problem_Berg}.
+In the most standard setting, the data provide observational samples of the ``microscopic''degrees of freedom of the system under study -- this can be biological sequences \cite{levy_potts_2017,cocco_inverse_2018}, firing patterns of neurons \cite{schneidman2006weak,roudi2009ising}, individuals in animal groups \cite{bialek2012statistical,cavagna2018physics}, stock markets \cite{bury2013market,borysov2015us} etc.
+Within a static modeling approach, frequently based on the maximum-entropy principle \cite{jaynes1957information}, data $\vec x$ are assumed to be generated independently from some unknown probability distribution $P(\vec x)$.
 This distribution describes the underlying interaction patterns between the observed degrees of freedom, and has to be learned from data to unveil the rules governing the system.
-In more rare cases where data correspond to observed time series, theoretical and algorithmic development is much less advanced than in the case of independent static data.
-%, assuming, e.g., that measurements are taken at the microscopic time scale.
+In more rare cases where data correspond to observed time series, theoretical and algorithmic development is much less advanced than in the case of independent static data \cite{Inverse_problem_Berg}.
 
-% Merge with my first par, don't be too specific with protein sequences
-One of the biggest application areas of inverse statistical mechanics is the modeling of biological processes.
-These applications are fuelled by the large amount of available data resulting from the impressive progress in experimental techniques in biology. 
-This is especially visible in the case of biological sequences,  with databases now harboring a vast amount of high-quality DNA or protein sequences \cite{sayers_genbank_2019,uniprotconsortium_uniprot_2018}.
-A common idea in this context is that it is possible to use characteristics of genes or organisms related by a common ancestry -- called \emph{homologous} -- to construct models of the selection acting on them.
-A successful example in this regard is the representation of protein sequences by probabilistic models in the so-called DCA method \cite{levy_potts_2017,cocco_inverse_2018}. 
-We will use this type of application to motivate our otherwise purely methodological study.
+Here we address a different case, the inverse problem for high-dimensional data, which show hierarchical correlations due to a data-generating process defined via a branching process.
+The motivation for this purely methodological study comes from the modeling of biological sequence data, which is one of the major fields of application of inverse statistical mechanics \cite{levy_potts_2017,cocco_inverse_2018}, but even more of other bioinformatic inference approaches, cf.~\cite{durbin1998biological,felsenstein2004inferring}.
+Sequence data are the result of natural evolution, {\em i.e.}~a branching process initiating in some unknown common ancestor of extant and thus observable sequences \cite{sayers_genbank_2019,uniprotconsortium_uniprot_2018,el2019pfam}.
+The prototypical datasets in this case are multiple-sequence alignments (MSA), with lines being a so-called homologous, {\em i.e.} evolutionarily related sequences, and columns specific positions deriving from some common ancestral position.
+Such MSA contain at least two kinds of information:
+\begin{itemize}
+    \item {\it Phylogenetic information:} the distances between sequences carry information about the evolutionary time since their common ancestor.
+Using phylogenetic methods \cite{felsenstein_phylogenies_1988,felsenstein2004inferring} we may reconstruct the evolutionary history of our dataset, represented by a phylogenetic tree.
+    \item {\it Co-evolutionary information:} positions in a sequence typically do not evolve independently, but rather in a correlated way.
+This co-evolution carries important information about the selection forces acting on evolving entities.
+This fact has been extensively studied in the case of protein sequences, and used to predict structure, mutational landscapes or networks of interacting proteins \cite{levy_potts_2017,cocco_inverse_2018}.
+\end{itemize}
+These two types of information are contained in two complementary features of the data: phylogenetic inference is based on the comparative analysis of different sequences, while co-evolutionary information is contained in the correlation of different columns of the MSA.
+Approaches aiming at one type of information typically neglect the other one: inference of phylogenies generally assumes that all positions in a sequence evolve independently, while co-evolutionary models of proteins assume that sequences in the MSA are independently distributed.
+This choice is motivated by the fact that taking the two types of correlations into account, \emph{i.e.} through time with phylogeny and accross trait values for co-evolution, results in very hard inference procedures, cf.~\cite{obermayer2014inverse,rodriguez2019toward}.
+However, this can lead to biases in the model parameters: it has for instance been shown that phylogenetic relations between protein sequences induce non-trivial correlations that are not related to protein function \cite{qin_power_2018,horta2020phylogenetic}.
 
-% Keep (shorten?)
-When modeling homologous genes or organisms, one has to distinguish between two complementary types of information: 
-\begin{itemize}
-    \item {\it Phylogenetic information:} the distances between gene sequences or traits in organisms carry information about the evolutionary time since their common ancestor. 
-    Using phylogenetic methods we may reconstruct the evolutionary history of a set of genes or species, and also the sequences or traits of their ancestors. 
-    \item {\it Co-evolutionary information:} positions in a sequence or traits in organisms typically do not evolve independently, but rather in a correlated way. 
-    This co-evolution carries important information about the selection forces acting on evolving entities.  
-    This fact has been extensively studied in the case of protein sequences, and used to predict structure, mutational landscapes or networks of interacting proteins.  
-\end{itemize}
+We have used phylogeny and co-evolution of biological sequences to explain our motivation, but the underlying problem is much more general.
+Instead of sequences of discrete characters, like amino acids or nucleotides, we may consider continuous phenotypic traits.
+The branching process is not necessarily the phylogeny of species, but it may be the genealogy of populations of the same species, or other branching processes like epidemics spreading or geographic migration.
 
-These two types of information are contained in two complementary features of the data: phylogenetic inference is based on the comparative analysis of different species or genes, while co-evolutionary information is contained in the correlation of different trait values in the same gene or organism.
-
-Models built using biological data usually ignore one of the two types of information. 
-For instance, inference of phylogenies typically assumes that all positions in a sequence evolve independently, while co-evolutionary models of proteins assume that their sequences are independently distributed.
-This choice is motivated by the fact that taking the two types of correlations into account, \emph{i.e.} through time with phylogeny and accross trait values for co-evolution, results in very hard inference procedures.
-However, this can lead to biases in the model parameters: it has for instance been shown that phylogenetic relations between protein sequences induce non-trivial correlations that are not related to protein function \cite{qin_power_2018}.
-
-Accounting for both co-evolution and phylogeny in models requires one to understand how interdependent traits under selection evolve along a tree. 
-A historically well-known way to represent such processes is to use Ornstein-Uhlenbeck dynamics (OU), which models traits as a Gaussian vector evolving in a quadratic potential that represents selection forces \cite{uhlenbeck_theory_1930, felsenstein_phylogenies_1988, hansen_stabilizing_1997}.
+Here we use a simple and very general model for the temporal evolution of correlated variables: a historically well-known way to represent such processes is to use Ornstein-Uhlenbeck dynamics (OU), which models configurations as Gaussian vectors evolving in a quadratic potential that represents selection forces \cite{uhlenbeck_theory_1930, felsenstein_phylogenies_1988, hansen_stabilizing_1997}.
 OU processes are commonly used in the field of phylogenetic comparative methods (PCM) \cite{bartoszek_phylogenetic_2012,mitov_fast_2020}. 
@@
 \bibliography{bib_phylo}
-
 \bibliographystyle{unsrt}
@@
 \appendix
-\setcounter{figure}{0}
-\renewcommand{\figurename}{Figure S}
-\setcounter{table}{0}
-\renewcommand{\tablename}{Table S}
+%\setcounter{figure}{0}
+%\renewcommand{\figurename}{Figure S}
+%\setcounter{table}{0}
+%\renewcommand{\tablename}{Table S}
 
-\section{Supplementary material} % (fold)
+\section{Description of technical details} % (fold)
 \label{sec:supplementary_material}
@@
 
-	We are interested in the case where the described Ornstein-Uhlenbeck process takes place on a tree. 
+We are interested in the case where the dynamics of the $L$-dimensional Ornstein-Uhlenbeck process takes place on a tree.
 For example, if configurations $\vec{x}$ represent quantitative traits of some organisms, the tree can represent the genealogy or phylogeny of these organisms.
-	Therefore, we have to be able to simulate the OU process on a tree. 
-	In practice, given a rooted tree such as the one shown in figure~\ref{fig:sample_tree} of the main text, we want to sample a configuration $\vec{x}$ for every node in such a way that equation~\eqref{eq:OUjoint} holds for every pair of nodes, the time $\Delta t$ then being the branch length connecting them.
-\\
+Therefore, to generate our datasets, we have to be able to simulate the OU process on a tree.
+In practice, given a rooted tree such as the one shown in Fig.~\ref{fig:sample_tree} above, we want to sample a configuration $\vec{x}$ for every node in such a way that Eq.~\eqref{eq:OUjoint} holds for every pair of nodes, the time $\Delta t$ being the path length connecting the nodes along the tree.
+
 We use a simple methodology to achieve this.
-	First, note that given an arbitrary configuration $\vec{x}_0$ and a time $\Delta t$, we can generate a new configuration $\vec{x}$ distributed according to the propagator in equation~\eqref{eq:OUpropagator} by the transformation
+First, note that given an arbitrary configuration $\vec{x}_0$ and a time $\Delta t$, we can generate a new configuration $\vec{x}$ distributed according to the propagator Eq.~\eqref{eq:OUpropagator} by exploiting the transformation
 \begin{equation}
-		\vx = \Lam^{\Delta t}\vx_0 + \Sig^{1/2}\vec{\eta}
+	\vx = \Lam^{\Delta t}\vx_0 + \Sig^{1/2}\vec{\eta}\ ,
 	\label{eq:OUsample}
 \end{equation}
-	where $\Lam$ and $\Sig$ are defined in equation~\eqref{eq:def_lambda}, and $\vec{\eta}$ is a vector of uncorrelated variables with distributions $\mathcal{N}(0,1)$. 
-	Moreover, if $\vec{x_0}$ is distributed according to equation~\eqref{eq:eq_distribution}, then $\vx$ and $\vx_0$ are distributed according to equation~\eqref{eq:OUjoint}. 
-	Note that equation \eqref{eq:OUsample} is quite different from the Langevin equation~\eqref{eq:langevin}, even though they have similar forms. 
-	While the Langevin equation describes the motion of $\vx$ in the potential $\bm{J}$, eq.~\eqref{eq:OUsample} directly samples from the OU process. \\
-	Given an already sampled internal node in the tree, equation~\eqref{eq:OUsample} allows to sample a configuration for each of its children. 
-	To sample the whole tree, we first sample the root note $\vx_0$ from the equilibrium distribution~\eqref{eq:eq_distribution}. 
-	By recursive applications of~\eqref{eq:OUsample}, we then simply work our way down the tree until all leaves are sampled.
+where $\Lam$ and $\Sig$ are defined in Eq.~\eqref{eq:def_lambda}, and $\vec{\eta}$ is a vector of uncorrelated variables drawn individually from the normal distribution $\mathcal{N}(0,1)$.
+Moreover, if $\vec{x_0}$ is distributed according to the equilibrium distribution Eq.~\eqref{eq:eq_distribution}, then $\vx$ and $\vx_0$ are distributed according to the joint distribution Eq.~\eqref{eq:OUjoint} describing two equilibrium configurations at finite time difference.
+Note that Eq.~\eqref{eq:OUsample} is quite different from the Langevin Eq.~\eqref{eq:langevin}, which describes the instantaneous dynamics of  $\vx$ in the potential given by $\bm{J}$, and which could also be simulated in more complicated situation where no analytical expression for the propagator can be derived.
+
+Given any already sampled internal node in the tree, Eq.~\eqref{eq:OUsample} allows to emit a configuration for each of its child nodes.
+To sample the whole tree, we first draw the root configuration $\vx_0$ from the equilibrium distribution Eq.~\eqref{eq:eq_distribution}.
+By recursive applications of Eq.~\eqref{eq:OUsample}, we then simply work our way down the tree until all leaves are sampled.
+Only the configurations in the leaves form the data set, the internal configuration remain hidden to our model-learning task.
 
@@
 
-	\subsubsection{Eigenvalues and eigenvectors of $\mathbf{C}^{-1}$} % (fold)
+\subsubsection{Eigenvalues and eigenvectors of $\mathbf{C}^{-1}$}
 \label{ssub:eigenvalues_and_eigenvectors}
 
-	The initial value that we take for the covariance matrix is the empirical one
+We initialize the covariance matrix using the empirical one:
 \begin{equation*}
-		\mathbf{C}^{emp} = \frac{1}{N}\sum_{i=1}^{N}\vx_i \cdot\vx_i^{\,T}.
+	\mathbf{C}^{emp} = \frac{1}{N}\sum_{i=1}^{N}\vx_i \cdot\vx_i^{\,T}\ .
 \end{equation*}
 Its eigenmodes $\{\rho^0_a, \vsa^{\,0}\}$ determine the starting point of the optimization.
-
-	[Insert here how we go from $\vsa^{\,0}$ to the corresponding set of Eulerian angles or to the corresponding skew symmetric matrix.]
+A suitable parametrization of $\vsa^{\,0}$ in terms of generalized Eulerian angles or a skew symmetric matrix is described below in Sec.~\ref{sub:parametrizations_of_eigenvectors}.
 
@@
 The optimization also requires that we initialize the time scale $\gamma$.
-	For this, we try to find the optimal $\gamma$ given the data $\mathbf{X}$, the tree, and the OU process defined by the empirical covariance matrix.
+For coherence with the last section, we need to find the optimal $\gamma$ given the data $\mathbf{X}$, the tree, and the OU process defined by the empirical covariance matrix.
 
-	The probability distribution $P$ for the configurations of two leaves $\vx_i$ and  $\vx_j$  separated by time $\Delta t_{ij}$ is given by equation~\eqref{eq:OUjoint} of the main text. 
-	With this distribution we can calculate analytically the average of the scalar product $\vx_i^{\,T}\vx_j$:
-	\begin{align*}
-		\langle \vx_i^{\,T}\vx_j \rangle_P &= \int \ddroit \vx_i \ddroit \vx_j P(\vx_i,\vx_j|\Delta t_{ij}) \sum_{a=1}^L   x_i^{a}x_j^{a}  \\
-		&=\sum_{i=a}^L \langle  x_i^{a} x_j^a \rangle_{P}.\\
-	\end{align*}
-	The covariance $\langle  x_i^{a} x_j^a \rangle_{P}$ between observations separated by a time $\Delta t_{ij}$ is given by equation~\eqref{eq:pairwisecov} of the main text.
-	Using this, we now have
+The probability distribution $P$ for the configurations of two leaves $\vx_i$ and  $\vx_j$  separated by time $\Delta t_{ij}$ is given by Eq.~\eqref{eq:OUjoint} of the main text.
+With this distribution we can analytically calculate the average of the scalar product $\vx_i^{\,T}\cdot\vx_j$ of two equilibrium configurations at given time separation:
 \begin{align}
-		\nonumber\langle \vx_i^{\,T}\vx_j \rangle_P &= \sum_{a=1}^L \left( \Lam^{\Delta t_{ij}} \mathbf{C} \right)_{aa}\\
+\langle \vx_i^{\,T}\cdot\vx_j \rangle_P &= \int \ddroit \vx_i \ddroit \vx_j P(\vx_i,\vx_j|\Delta t_{ij}) \sum_{a=1}^L   x_i^{a}x_j^{a}  \nonumber \\
+	&=\sum_{a=1}^L \langle  x_i^{a} x_j^a \rangle_{P}.
+\end{align}
+The covariance $\langle  x_i^{a} x_j^a \rangle_{P}$ of two observations separated by time $\Delta t_{ij}$ is given by Eq.~\eqref{eq:pairwisecov}. Using this, we find
+\begin{align}
+		\nonumber\langle \vx_i^{\,T}\cdot\vx_j \rangle_P &= \sum_{a=1}^L \left( \Lam^{\Delta t_{ij}} \mathbf{C} \right)_{aa}\\
 		\nonumber&= \Tr\left( \Lam^{\Delta t_{ij}} \mathbf{C} \right)\\
@@
 \end{align}
-	Having initialized the covariance matrix $\mathbf{C}$ at its empirical value, we know the values of all members of the r.h.s. of equation~\eqref{eq:scalar_product} except that of $\gamma$. 
-	On the other hand, equivalent versions of equation~\eqref{eq:scalar_product} can be written for all pairs of configurations $i$ and $j$. 
-	To find an initial value of $\gamma$ which is consistent with the data and the empirical covariance matrix, we search for one that best explains the observed scalar products between configurations. 
+Having initialized the covariance matrix $\mathbf{C}$ with its empirical value, we know the values of all members of the r.h.s.
+of Eq.~\eqref{eq:scalar_product} except the one of $\gamma$.
+To find an initial value of $\gamma$ which is consistent with the data and the empirical covariance matrix for all pairs of data configurations $i<j$, we search for one that best explains the observed scalar products between configurations.
 We thus define $\gamma^0$ to be the argument minimizing the functional $F(\gamma)$:
 \begin{equation}
-	 	F(\gamma) = \sum_{1\leq i<j\leq N}\left[ \vx_i^{\,T}\vx_j - \sum_{a=1}^L \rho_a^{-1}e^{-\gamma\rho_a\Delta t_{ij}} \right].
+    F(\gamma) = \sum_{1\leq i<j\leq N}\left[ \vx_i^{\,T}\cdot\vx_j - \sum_{a=1}^L \rho_a^{-1}e^{-\gamma\rho_a\Delta t_{ij}} \right].
 \end{equation} 
-	As $F$ depends on one scalar parameter, it is straightforward to minimize it, allowing us to initialize $\gamma$ to a reasonable value.
-
-% subsection initializing_parameters (end)
-
-% section supplementary_material (end)
+Since $F$ depends on a single scalar parameter, it is straightforward to minimize it and thereby to initialize $\gamma$ to an empirically reasonable value.
 
@@
 \label{sub:parametrizations_of_eigenvectors}
-\subsubsection{Parametrization in term of generalized Eulerian angles}
-The aim is to  parameterize each base vector $\vec{s}_a$ according to the $a-th$ column of an arbitrary orthogonal matrix $\bm S$ parameterized   in terms of $L(L-1)/2$ independent variables  $\theta_ {pq}, p =1,2, ..., L; q =1,2, ..., L; p <q $, named Generalized Eulerian angles (Eulerian angles is for L=3).
+\subsubsection{Parametrization using generalized Eulerian angles}
+The idea is to write the base vectors $\vec{s}_a$ as columns of an orthogonal matrix $\bm S$, and to parameterize this matrix in terms of $L(L-1)/2$ independent variables  $\theta_ {pq}$ with $1 \leq p <q \leq L $.
+These parameters are called generalized Eulerian angles, since they generalize Eulerian angles to $L>3$.
 
-To construct this matrix we start from a  transformation in a two-dimensional subspace of an $L-dimensional$ space which is given by an L-dimensional matrix of the  form:
-
+To construct this matrix we start from a rotational transformation in a two-dimensional subspace of an $L$-dimensional space. It is given by an $L$-dimensional matrix of the form
 \begin{equation} 
@@
 \end{array}
-\right)
+\right) \ ,
 \end{equation}
-where all diagonal elements are unity except the diagonal elements in the $pth$ column and the $qth$ column, which are $\cos\theta_{pq}$ and all off-diagonal elements are zero except the one corresponding to the intersection of the $pth$ row and the $qth$ column, which is $\sin\theta_{pq}$, and that on the intersection of the $qth$ row and the $pth$ column, which is $-\sin\theta_{pq}$.
-There are $L(L - 1)/2$  matrices of the form indicated in Equation (1), corresponding to all choices of $p$ and $q$ with $p <q$.
+where all diagonal elements are unity except for the diagonal elements in the $p$th and the $q$th column, which equal $\cos\theta_{pq}$.
+All off-diagonal elements are zero except for the one corresponding to the intersection of the $p$th row and the $q$th column, which is $\sin\theta_{pq}$, and that on the intersection of the $q$th row and the $p$th column, which equals $-\sin\theta_{pq}$.
+There are $L(L - 1)/2$  matrices of this form, corresponding to all choices of $p$ and $q$ with $1\leq p <q\leq L$.
 
-An arbitrary  $L-dimensional$ orthogonal matrix $\bm S$ can be represented as a product of these $L(L-1)/2$ orthogonal matrices with appropriate values of the $L(L-1)/2$ independent parameters $\theta_{pq}$.
+An arbitrary $L$-dimensional orthogonal matrix $\bm S$ can be represented as a product of these $L(L-1)/2$ orthogonal matrices with appropriate values of the $L(L-1)/2$ independent parameters $\theta_{pq}$:
+\begin{equation}
+\label{eq:eulerprod}
+\bm S (\bm \theta)=\prod_{\left\lbrace pq\right\rbrace }T_{pq}(\theta_{pq}) \ .
+\end{equation}
+The basic idea is to express a general rotation in $L$ dimensions as the result of $L(L-1)/2$ successive rotations by angles $\theta_ {pq}$ in the two-dimensional subspaces given by the components $p, q$.
+For $L=3$ we get the classical Eulerian matrix.
 
+Ref.~{\color{red} [7] CHECK, is it [18]?
+use ref} exposes a recursive algorithm to efficiently perform this multiplication, as well as the   construction of the matrix derivatives respect to parameters $\theta_{pq}$.
+To start, Eq.~\eqref{eq:eulerprod} is transformed into a set of recursive equations, with $\bm S$ being equal to the final step
 \begin{equation}
-\bm S (\bm \theta)=\prod_{\left\lbrace pq\right\rbrace }T_{pq}(\theta_{pq})
+\bm S(\bm \theta)=\bm T^{(L)} \ .
 \end{equation}
-
- This matrix is in charge of rotations of vectors in $L-dimensional$ space. For $L=3$ we get the classical Eulerian matrix.
-
- In reference [7] is explained an recursive algorithm to efficiently perform this multiplication,  as well as the   construction of the matrix   derivatives respect to parameters $\theta_{pq}$.
-
-For the first, equation  (2) is transformed in  the recurrence equations:
-
-\begin{equation}
-\bm S(\vec{\theta} )=\bm T^{(L)}
-\end{equation}
-
-with $\theta_ {pq}, p =1,2, ..., L; q =1,2, ..., L; p <q$ and
-
+For all $1\leq k < l\leq L$ we iterate over $n$: {\color{red} [7] What is the initialization?
+Please check carefully, also the correct ranges of the $kl$ which seem to go to $n$ only.
+It would also be better to avoid $s$, which we use for the eigenvectors.}
 \begin{equation}
@@
     \begin{split}
-&s^{(n)}_{k+1,l}=\sin\theta_{kn} * t^{(n)}_{kl}+\cos\theta_{kn}*s^{(n)}_{kl}\\
-&s^{(n)}_{k+1,n}=\cos\theta_{kn}*s^{(n)}_{kn}\\
-&s^{(n)}_{1,l}=-\delta_{ln}\\
-&\theta_{nn}=\pi/2\\
-&\end{split}
+        s^{(n)}_{k+1,l}&=\sin\theta_{kn} * t^{(n)}_{kl}+\cos\theta_{kn}*s^{(n)}_{kl}\\
+        s^{(n)}_{k+1,n}&=\cos\theta_{kn}*s^{(n)}_{kn}\\
+        s^{(n)}_{1,l}&=-\delta_{ln}\\
+        \theta_{nn}&=\pi/2
+    \end{split}
 \end{equation}
@@
 \bm T^{(n-1)} & \bm 0  \\
-\bm 0 &  \bm I\\
-
+\bm 0 &  \bm I
 \end{array}
@@
 
-This mean that the eigenvectors $\vec{s}_a$ of the matrix $\bm C$ with dimensions $L\times L$ can be written as:
+As a consequence, the eigenvectors $\vec{s}_a$ of the matrix $\bm C$ with dimensions $L\times L$ can be written as:
 \begin{equation}
-s^{k}_a=T^{(L)}_{.,a}=\cos\theta_{kL} * t^{(L)}_{ka}-\sin\theta_{kL}* s^{(L)}_{ka} \;\;\; \text{for} \;k=1,..,L
+s^{k}_a=T^{(L)}_{.,a}=\cos\theta_{kL} * t^{(L)}_{ka}-\sin\theta_{kL}* {\color{red} s}^{(L)}_{ka} \;\;\; \text{for} \;k=1,..,L \ .
 \end{equation}
+To use this expression, we still need to determine parameters $\bm \theta$  given an orthogonal matrix $\bm S$, such that all equations
+\begin{equation}
+    S_{ij}(\bm \theta)=S_{ij}
+\end{equation}
+are satisfied.
+This system of nonlinear transcendental equations  cannot be solved algebraically.
+However, it is possible to overcome this issue finding the set of $\bm \theta$ which minimize the square distance between the target and the parametrized matrices:
+\begin{equation}
+ {\bm{\hat \theta}} = \text{argmin}_{\bm \theta} \sum_{i<j}\left[ S_{ij}-S_{ij}(\bm \theta)\right] ^2 \ .
+\end{equation}
+This is useful when we initialize parameters $\bm \theta$ for the matrix formed by the eigenvectors of the empirical covariance matrix.
 
-Then there is another  important problem : how to determine parameters $\bm \theta$  given an orthogonal matrix $\bm Q$ such that all equations
+\subsubsection{Parametrization in term of the exponential of a skew-symmetric matrix }
 
-$$S_{ij}(\bm \theta)=Q_{ij}$$
+Any orthogonal matrix can be written as the exponential of a skew-symmetric matrix $\bm X=-\bm X^T$: 
+\begin{equation}
+    \bm S=\exp (\bm X)\ ,
+\end{equation}
+since $\exp(\bm X)^T=\exp(\bm X^T)=\exp(-\bm X)=\exp(\bm X)^{-1}$.
 
-are satisfied.
-This system of nonlinear transcendental equations  can't be algebraically solved, however it's possible overcome  this issue finding the set of $\bm \theta$ which minimize the function:
-\begin{equation}
-f(\bm \theta)=\sum_{i,j}\left[ Q_{ij}-S_{ij}(\bm \theta)\right] ^2
-\end{equation}
-This is useful when we initialize parameters $\bm \theta$  from the matrix formed by eigenvectors of the empirical covariance matrix.
-\subsubsection{Parametrization in term of the exponential of a skew-symmetric matrix }
-The exponential of a skew-symmetric matrix $\bm X=-\bm X^T$  : 
- \begin{equation}
- \bm S=\exp (\bm X)
- \end{equation}
-
- is  an  orthogonal  matrix : $\exp(X)^T=\exp(X^T)=\exp(-X)=\exp(X)^{-1}$.
-
- The derivatives on $L(L-1)/2$ independent variables of $\bm X$  is formally defined by
-
+The derivatives with respect to the $L(L-1)/2$ independent variables of $\bm X$ is formally defined by
 \begin{equation}
@@
 \end{equation}
-
- where $\bm E^{jk}$ for $j > k$ is defined as a skew-symmetric matrix that has only two nonzero elements:
+where $\bm E^{jk}$ for $j > k$ is defined as a skew-symmetric matrix that has only two nonzero entries in positions $(j,k)$ and $(k,j)$:
 \begin{equation}
@@
 
-To obtain   a skew-symmetric matrix    $\bm X$ from an orthogonal matrix $\bm Q$    is enough invert the exponential relation :
- $$\bm X=\log \bm Q$$
+To obtain a skew-symmetric matrix $\bm X$ from an orthogonal matrix $\bm Q$, it is sufficient to invert the exponential relation,
+$\bm X=\log \bm Q$.
+If we write eigenvectors of the covariance matrix as the columns of the exponential of the skew-symmetric matrix $\bm X$, we are able to perform the optimization over its $L(L-1)/2$ independent entries.
 
-  If we write eigenvectors of the likelihood function as the columns of the exponential of the skew-symmetric matrix $\bm X$, then we are able to perform the optimization over it's $L(L-1)$ independent variables of $\bm X$.
-
- Furthermore, as the eigenvectors were parameterized by an algebraic function, the complete likelihood  is described in terms of arithmetics operation and elementary functions.
-This allow to compute  the gradient via  automatic differentiation (AD), making computations in our optimization process faster.
- The idea  is to represent a function as a computational graph, also called a Wengert list , where each node  in this list will represent an intermediate result of the computation.
-The intermediate results can then be assembled using the chain rule to get the final derivative we’re looking for.
-There is two    main algorithms  to traverses the chain rule in AD: forward mode and reverse mode, here we used reverse mode  algorithm  which is the of choice for back-propagation in deep learning .
-In particular we implemented  it using the Julia package Zygote.jl .
+Furthermore, as the eigenvectors are parametrized by an algebraic function, the complete likelihood is described in terms of arithmetic operations and elementary functions.
+This allow to compute the gradient via automatic differentiation (AD)
+{\color{red} CITE}, making computations in our optimization process faster.
+The idea  is to represent a function as a computational graph, also called Wengert list {\color{red} CITE}, where each node in the list represents an intermediate result of the computation.
+The intermediate results can be assembled using the chain rule to get the final derivative we are looking for.
+There are two main algorithms to traverse the chain rule in AD: a forward mode and a reverse mode.
+Here, we used the reverse-mode algorithm, which is also the one of choice for back-propagation in deep learning.
+In particular we implemented it using the Julia package Zygote.jl.
 
@@
 
-\subsection{Supplementary figures}
+\subsection{Supporting figures}
 \label{sub:supplementary_figures}

### 2021-02-09 main.tex
changes and comments to SI

@@
 We are interested in the case where the dynamics of the $L$-dimensional Ornstein-Uhlenbeck process takes place on a tree.
-For example, if configurations $\vec{x}$ represent quantitative traits of some organisms, the tree can represent the genealogy or phylogeny of these organisms.
+For example, if configurations $\{\vec{x}\}$ represent quantitative traits of some organisms, the tree can represent the genealogy or phylogeny of these organisms.
 Therefore, to generate our datasets, we have to be able to simulate the OU process on a tree.
-In practice, given a rooted tree such as the one shown in Fig.~\ref{fig:sample_tree} above, we want to sample a configuration $\vec{x}$ for every node in such a way that Eq.~\eqref{eq:OUjoint} holds for every pair of nodes, the time $\Delta t$ being the path length connecting the nodes along the tree.
+In practice, given a rooted tree such as the one shown in Fig.~\ref{fig:sample_tree}  of the main text, we want to sample a configuration $\vec{x}$ for every node in such a way that Eq.~\eqref{eq:OUjoint} holds for every pair of nodes, with time $\Delta t$ being the path length connecting the nodes along the tree.
 
@@
 Given any already sampled internal node in the tree, Eq.~\eqref{eq:OUsample} allows to emit a configuration for each of its child nodes.
-To sample the whole tree, we first draw the root configuration $\vx_0$ from the equilibrium distribution Eq.~\eqref{eq:eq_distribution}.
+To sample the whole tree, we first draw the root configuration $\vx_r$ from the equilibrium distribution Eq.~\eqref{eq:eq_distribution}.
 By recursive applications of Eq.~\eqref{eq:OUsample}, we then simply work our way down the tree until all leaves are sampled.
-Only the configurations in the leaves form the data set, the internal configuration remain hidden to our model-learning task.
+Only the configurations at the leaves form the data set, and the internal configuration remain hidden to our model-learning task.
 
@@
 \end{equation*}
-Its eigenmodes $\{\rho^0_a, \vsa^{\,0}\}$ determine the starting point of the optimization.
-A suitable parametrization of $\vsa^{\,0}$ in terms of generalized Eulerian angles or a skew symmetric matrix is described below in Sec.~\ref{sub:parametrizations_of_eigenvectors}.
+Its eigenmodes $\{\rho^0_a, \vsa^{\;0}\}$ determine the starting point of the optimization.
+A suitable parametrization of $\vsa^{\;0}$ in terms of generalized Eulerian angles or a skew symmetric matrix is described below in Sec.~\ref{sub:parametrizations_of_eigenvectors}.
 
@@
 With this distribution we can analytically calculate the average of the scalar product $\vx_i^{\,T}\cdot\vx_j$ of two equilibrium configurations at given time separation:
-\begin{align}
-\langle \vx_i^{\,T}\cdot\vx_j \rangle_P &= \int \ddroit \vx_i \ddroit \vx_j P(\vx_i,\vx_j|\Delta t_{ij}) \sum_{a=1}^L   x_i^{a}x_j^{a}  \nonumber \\
-	&=\sum_{a=1}^L \langle  x_i^{a} x_j^a \rangle_{P}.
-\end{align}
+% \begin{align}
+% \langle \vx_i^{\,T}\cdot\vx_j \rangle_P &= \int \ddroit \vx_i \ddroit \vx_j P(\vx_i,\vx_j|\Delta t_{ij}) \sum_{a=1}^L   x_i^{a}x_j^{a}  \nonumber \\
+% 	&=\sum_{a=1}^L \langle  x_i^{a} x_j^a \rangle_{P}.
+% \end{align}
+\begin{equation}
+	\langle \vx_i^{\,T}\cdot\vx_j \rangle_P = \sum_{a=1}^L \langle  x_i^{a} x_j^a \rangle_{P}.
+\end{equation}
 The covariance $\langle  x_i^{a} x_j^a \rangle_{P}$ of two observations separated by time $\Delta t_{ij}$ is given by Eq.~\eqref{eq:pairwisecov}. Using this, we find
@@
 Ref.~{\color{red} [7] CHECK, is it [18]?
-use ref} exposes a recursive algorithm to efficiently perform this multiplication, as well as the   construction of the matrix derivatives respect to parameters $\theta_{pq}$.
+use ref} exposes a recursive algorithm to efficiently perform the matrix multiplication of Eq.~\eqref{eq:eulerprod}, as well as the construction of the derivatives of $\bm S$ with respect to parameters $\theta_{pq}$.
 To start, Eq.~\eqref{eq:eulerprod} is transformed into a set of recursive equations, with $\bm S$ being equal to the final step
@@
 Please check carefully, also the correct ranges of the $kl$ which seem to go to $n$ only.
-It would also be better to avoid $s$, which we use for the eigenvectors.}
+It would also be better to avoid $s$, which we use for the eigenvectors.} \pierre{Adding on Martin's comment above: I don't understand the equations below.
+Matrices $T_{pq}$ defined above are not used at all (I'm guessing they're used in the initialization but it's not said).
+It's not said that $1\leq n \leq L$ (is it the case?).
+I don't really understand what $s^{(n)}_{kl}$ is (Eq.
+(A9) gives some information but it's quite confusing) and as Martin points it has the same name as the eigenvectors we want to compute.
+We also use several different set of indices to go from $1$ to $L$: $(p,q)$, $n$, $(k,l)$ and $(i,j)$.
+In the maintext we use $(a,b)$ for this, making this confusing.
+Finally, I think the $*$ symbol should not be here if it just stands for normal multiplication, or maybe replaced by $\cdot$ if needed.
+Finally, this does not say how we compute the derivative of $\bm S$ w.r.
+to $\{\theta_{pq}\}$, which is the most important part.
+Much more explanations needed (at least for me...)!
+}
 \begin{equation}
@@
 \begin{equation}
+	\label{eq:sksym_def}
     \bm S=\exp (\bm X)\ ,
 \end{equation}
-since $\exp(\bm X)^T=\exp(\bm X^T)=\exp(-\bm X)=\exp(\bm X)^{-1}$.
+since $\exp(\bm X)^T=\exp(\bm X^T)=\exp(-\bm X)=\exp(\bm X)^{-1}$. \pierre{This does not prove the statement above but the other implication: the exponential of a skew-symmetric is orthogonal.}
+To obtain a skew-symmetric matrix $\bm X$ from an orthogonal matrix $\bm S$, it is sufficient to invert the exponential relation, $\bm X=\log \bm S$.
+\pierre{What needs to be proven is that $\log \bm S$ is always defined if $\bm S$ is orthogonal, and that $\bm$ X is skew-symetric.}
 
-The derivatives with respect to the $L(L-1)/2$ independent variables of $\bm X$ is formally defined by
+The advantage of expressing $\bm S$ in this form is that $\bm X$ has $L(L-1)/2$ entries that can be varied independently. 
+This allows us to perform the optimization over $L(L-1)/2$ independent parameters, with derivatives with respect to the independent entries of $\bm X$ being defined by
 \begin{equation}
 	\frac{\partial \bm S}{\partial X_{jk}}=\lim_{h\rightarrow 0} \frac{1}{h} \left( \exp(\bm X+ h \bm E^{jk}) - \exp(\bm X)\right) 
+	\label{eq:derivative_sksym}
 \end{equation}
@@
 
-To obtain a skew-symmetric matrix $\bm X$ from an orthogonal matrix $\bm Q$, it is sufficient to invert the exponential relation,
-$\bm X=\log \bm Q$.
-If we write eigenvectors of the covariance matrix as the columns of the exponential of the skew-symmetric matrix $\bm X$, we are able to perform the optimization over its $L(L-1)/2$ independent entries.
-
-Furthermore, as the eigenvectors are parametrized by an algebraic function, the complete likelihood is described in terms of arithmetic operations and elementary functions.
-This allow to compute the gradient via automatic differentiation (AD)
-{\color{red} CITE}, making computations in our optimization process faster.
+It is not possible to give a simple analytical form to Eq.~\eqref{eq:derivative_sksym}. 
+However, since $\bm S$ is obtained through a simple algebraic expression (Eq.~\eqref{eq:sksym_def}), it is possible to compute its derivative with respect to entries of $\bm X$ through automatic differentiation techniques {\color{red} CITE}.
+\pierre{About what is below: I don't think it's necessary to explain how AD works. It's a technique at the base of all deep-learning methods, so we could just say that plus cite something.}
 The idea  is to represent a function as a computational graph, also called Wengert list {\color{red} CITE}, where each node in the list represents an intermediate result of the computation. 
@@
 Here, we used the reverse-mode algorithm, which is also the one of choice for back-propagation in deep learning.
-In particular we implemented it using the Julia package Zygote.jl.
+In particular we implemented it using the Julia package Zygote.jl \cite{Zygote.jl-2018}.
 
@@
 Let's assume that the tree is binary, symmetric and completely homogeneous with all branches having the same length $\Delta t$. 
-As an example, the covariance matrix for such a tree with $K=2$ branching events and four leaves is
+As an example, the covariance matrix for such a tree with $K=2$ levels with branching and four leaves is
 \begin{equation} 
@@
 
+% \begin{equation} 
+% \bm G^a =  \left(
+% \begin{array}{cccc}
+% {  \rho^{-1}_a } & {\rho^{-1}_a } e ^{-2\gamma\rho_a\Delta t} & {\rho^{-1}_a } e ^{-4\gamma\rho_a\Delta t}& {e ^{-4\gamma\rho_a\Delta t}} \\
+% {\rho^{-1}_a } e ^{-2\gamma\rho_a\Delta t} & {\rho^{-1}_a } & {\rho^{-1}_a } e ^{-4\gamma\rho_a\Delta t} & {\rho^{-1}_a } e ^{-4\gamma\rho_a\Delta t} \\
+% {\rho^{-1}_a } e ^{-4\gamma\rho_a\Delta t} & {\rho^{-1}_a} e ^{-4\gamma\rho_a\Delta t}& {\rho^{-1}_a } & {\rho^{-1}_a } e ^{-2\gamma\rho_a\Delta t}\\
+% {\rho^{-1}_a } e ^{-4\gamma\rho_a\Delta t} & {\rho^{-1}_a } e ^{-4\gamma\rho_a\Delta t} & {\rho^{-1}_a } e ^{-2\gamma\rho_a\Delta t} & {\rho^{-1}_a } \\
+% \end{array}\label{eq:G_a}
+% \right).
+% \end{equation}
 \begin{equation} 
-\bm G^a =  \left(
+\bm G^a =  \rho_a^{-1}\left(
 \begin{array}{cccc}
-{  \rho^{-1}_a } & {\rho^{-1}_a } e ^{-2\gamma\rho_a\Delta t} & {\rho^{-1}_a } e ^{-4\gamma\rho_a\Delta t}& {e ^{-4\gamma\rho_a\Delta t}} \\
-{\rho^{-1}_a } e ^{-2\gamma\rho_a\Delta t} & {\rho^{-1}_a } & {\rho^{-1}_a } e ^{-4\gamma\rho_a\Delta t} & {\rho^{-1}_a } e ^{-4\gamma\rho_a\Delta t} \\
-{\rho^{-1}_a } e ^{-4\gamma\rho_a\Delta t} & {\rho^{-1}_a} e ^{-4\gamma\rho_a\Delta t}& {\rho^{-1}_a } & {\rho^{-1}_a } e ^{-2\gamma\rho_a\Delta t}\\
-{\rho^{-1}_a } e ^{-4\gamma\rho_a\Delta t} & {\rho^{-1}_a } e ^{-4\gamma\rho_a\Delta t} & {\rho^{-1}_a } e ^{-2\gamma\rho_a\Delta t} & {\rho^{-1}_a } \\
+1 & e ^{-2\gamma\rho_a\Delta t} & e ^{-4\gamma\rho_a\Delta t}& {e ^{-4\gamma\rho_a\Delta t}} \\
+e ^{-2\gamma\rho_a\Delta t} & 1 & e ^{-4\gamma\rho_a\Delta t} & e ^{-4\gamma\rho_a\Delta t} \\
+e ^{-4\gamma\rho_a\Delta t} & e ^{-4\gamma\rho_a\Delta t}& 1 & e ^{-2\gamma\rho_a\Delta t}\\
+e ^{-4\gamma\rho_a\Delta t} & e ^{-4\gamma\rho_a\Delta t} & e ^{-2\gamma\rho_a\Delta t} & 1 \\
 \end{array}\label{eq:G_a}
@@
 
-For hyper-geometric matrices as \eqref{eq:G_a} of dimension $2^K$, there are $K+1$ different eigenvalues  given by:
+Matrices such as the one in \eqref{eq:G_a} are called \emph{hyper-geometric}.
+For dimensions $2^K$, they have $K+1$ different eigenvalues given by: \pierre{ref?}
 
@@
 \begin{cases}
- (1+\sum_{l=1}^{k-1} 2^{l-1} e ^{-2l\gamma\rho_a\Delta t}-2^{k-1} e ^{-2k\gamma\rho_a\Delta t})\;\;\; k\in[1,K] \\
- (1+\sum_{l=1}^{K} 2^{l-1} e ^{-2l\gamma\rho_a\Delta t}) \;\;\;\;\;\;\;k=K+1
+ 1+\sum_{l=1}^{k-1} 2^{l-1} e ^{-2l\gamma\rho_a\Delta t}-2^{k-1} e ^{-2k\gamma\rho_a\Delta t}, \;\text{for}\; k\in[1,K] \\
+ 1+\sum_{l=1}^{K} 2^{l-1} e ^{-2l\gamma\rho_a\Delta t}, \;\text{for}\; k=K+1
 \end{cases}
@@
 
-where $\lambda_{K+1}\ge\lambda_{K}\cdots\ge\lambda_{1}$. For $k<K+1$, the degeneracy of eigenvalue $\lambda_k$ is $d_k=2^{K-k}$.  
-The associated eigenvectors are independent of the parameter $\rho_a$and reflect the events in the phylogenetic tree.
-Each eigenvector $\vec{u}_k$ captures the duplication events in the $(K+1 -k) st$ generation:
+where $\lambda_{K+1}\ge\lambda_{K}\cdots\ge\lambda_{1}$. 
+For $k<K+1$, the degeneracy of eigenvalue $\lambda_k$ is $d_k=2^{K-k}$.  
+The associated eigenvectors are independent of the parameter $\rho_a$ and reflect the events in the phylogenetic tree.
+Each eigenvector $\vec{u}_k$ of length $2^K$ captures the duplication events in the $(K+1 -k) st$ generation:
 
@@
 \begin{cases}
-{(\underbrace{\overbrace{1,\ldots,1}^{2^{k-1}},\overbrace{-1,\ldots,-1}^{2^{k-1}}}_Q,0,\ldots,0)\ \bigcup \Gamma(u_k) } \;\;\; k\in[1,K] \\
-(1,1,1,\ldots,1,1,1)\;\;\;\;\;\;\;k=K+1
+{(\underbrace{\overbrace{1,\ldots,1}^{2^{k-1}},\overbrace{-1,\ldots,-1}^{2^{k-1}}}_Q,0,\ldots,0)\ \bigcup \Gamma(u_k) }, \;\text{for}\; k\in[1,K] \\
+(1,1,1,\ldots,1,1,1), \;\text{for}\; k=K+1
 \label{eq:eigenvec_simpler_case}
@@
 
-where $\Gamma(\vec{u}_k )$ represents the $d_k$ combinations obtained by shifting the block of length $Q$, generating all eigenvectors corresponding to the eigenvalue $\lambda_k$.
+where $\Gamma(\vec{u}_k )$ represents the $d_k$ combinations obtained by shifting the block of length $Q=2^k$, generating all eigenvectors corresponding to the eigenvalue $\lambda_k$.  
 The eigenvectors are orthogonal to each other, and can be  normalized and arranged horizontally into a matrix $U$.
 
-To compute the gradient of the likelihood, we need  derivatives of $\lambda_k(\rho_a,\gamma)$ with respect to $\rho_a$ and $\gamma$, which can be directly obtained from expression \eqref{eq:lambda_hyper}.
+To compute the gradient of the likelihood, derivatives of $\lambda_k(\rho_a,\gamma)$ with respect to $\rho_a$ and $\gamma$ can be directly obtained from expression \eqref{eq:lambda_hyper}.
 
@@
 
-The proposed inference scheme was transformed into a multidimensional nonlinear optimization problem for which a variant  of quasi-Newton Methods (QNMs) was used.
-The main feature in QNMs is that Hessian matrix $\hat{H}$  does not need to be computed instead it’s approximated.
-The Hessian approximation $\hat{H}$ is chosen to satisfy the secant equation:
+The proposed inference scheme was transformed into a multidimensional nonlinear optimization problem for which we can compute the gradient of the optimized quantity $\mathcal{L}$.
+To solve it, we used a variant of the quasi-Newton methods (QNM).  
+The main feature in QNM when compared to standard Newton method is that the Hessian matrix $\bm H$ is approximated instead of computed exactly. 
+The Hessian approximation $\mathbf{\hat{H}}$ is chosen to satisfy the secant equation: \pierre{What's $\vec{x}_k$? I suppose it stands for parameters after the $k$th iteration? }
 
-$$\nabla L(\vec{x}_k+\Delta\vec{x})=\nabla L(\vec{x}_k)+\hat{H} \Delta\vec{x} $$
+$$\nabla \mathcal{L}(\vec{x}_k+\Delta\vec{x})=\nabla \mathcal{L}(\vec{x}_k)+\mathbf{\hat{H}} \Delta\vec{x} $$.
 
-for $n-dimensions$ $\hat{H}$ is undetermined, the various QNMs differ in their choice of the solution to the secant equation.
-We used LBFGS variant  (limited memory version of BFGS), this particular method is based in choosing $\hat{H}$ as a positive definite matrix where
+for $n-dimensions$ $\mathbf{\hat{H}}$ is undetermined \pierre{what do you mean "for $n$ dimensions"?}, the various QNMs differ in their choice of the solution to the secant equation.
+We used LBFGS variant \pierre{maybe a reference}.
+This particular method is based on choosing $\mathbf{\hat{H}}$ as a positive definite matrix where
 
-$$\hat{H}_{k+1}=\hat{H}_{k}+\frac{y_k y^{T}_k}{y_k\Delta x_k} - \frac{\hat{H}_k\Delta x_k(\hat{H}_k\Delta x_k)^T}{\Delta x^T_k\hat{H}_k\Delta x_k} \;\;\;\text{and}\;\;\; y_k=\nabla L(\vec{x}_k+1)-\nabla L(\vec{x}_k)$$
+$$\mathbf{\hat{H}}_{k+1}=\mathbf{\hat{H}}_{k}+\frac{y_k y^{T}_k}{y_k\Delta x_k} - \frac{\mathbf{\hat{H}}_k\Delta x_k(\mathbf{\hat{H}}_k\Delta x_k)^T}{\Delta x^T_k\mathbf{\hat{H}}_k\Delta x_k} \;\;\;\text{and}\;\;\; y_k=\nabla \mathcal{L}(\vec{x}_k+1)-\nabla \mathcal{L}(\vec{x}_k)$$
+\pierre{What's $H_k$? I'm guessing Hessian matrix at iteration $k$, but then all above equations should have been with $H_k$ as well instead of just $H$.}
 
-The method implementation was done using   NLopt package \cite{NLopt} .
+Implementation of the method was done using NLopt package \cite{NLopt} .
 

### 2021-02-11 main.tex
arXiv submission

@@
 \newcommand{\pierre}[1]{{\color{red}Pierre: #1}}
+\newcommand{\edwin}[1]{{\color{red}Edwin: #1}}
 \newcommand{\martin}[1]{{\color{blue}Martin: #1}}
@@
 
-\author{Edwin Rodriguez Horta} 
+\author{Edwin Rodr\'{i}guez Horta} 
 \affiliation{Sorbonne Universit\'{e}, CNRS, Institut de Biologie
@@
 \begin{abstract}
-Inverse statistical physics aims at inferring models compatible with a set of empirical averages estimated from a high-dimensional dataset of independently distributed equilibrium configurations.
-However, in several applications such as biology, data result from stochastic evolutionary processes, and the relation between configurations is characterized by a hierarchical correlation structure, typically represented by a tree.
-In turn, empirical averages of observables superpose intrinsic signals related to the equilibrium distribution of the studied system, and spurious historical (or phylogenetic) signals characterizing the specific correlated data-generating process.
+Inverse statistical physics aims at inferring models compatible with a set of empirical averages estimated from a high-dimensional dataset of independently distributed equilibrium configurations of a given system.
+However, in several applications such as biology, data result from stochastic evolutionary processes, and configurations are related through a hierarchical structure, typically represented by a tree, and therefore not independent.
+In turn, empirical averages of observables superpose intrinsic signal related to the equilibrium distribution of the studied system and spurious historical (or phylogenetic) signal resulting from the structure underlying the data-generating process.
 The naive application of inverse statistical physics techniques therefore leads to systematic biases and an effective reduction of the sample size.
 To advance on the currently open task of extracting intrinsic signals from correlated data, we study a system described by a multivariate Ornstein-Uhlenbeck process defined on a finite tree.
-Using a Bayesian framework, we can disentangle covariances in the data, which are a result of their multivariate Gaussian equilibrium distribution, from those resulting from the historical correlations.
+Using a Bayesian framework, we can disentangle covariances in the data corresponding to their multivariate Gaussian equilibrium distribution from those resulting from the historical correlations.
 Our approach leads to a clear gain in accuracy in the inferred equilibrium distribution, which corresponds to an effective two- to fourfold increase in sample size.
@@
 This distribution describes the underlying interaction patterns between the observed degrees of freedom, and has to be learned from data to unveil the rules governing the system.
-In more rare cases where data correspond to observed time series, theoretical and algorithmic development is much less advanced than in the case of independent static data \cite{Inverse_problem_Berg}.
-
-Here we address a different case, the inverse problem for high-dimensional data, which show hierarchical correlations due to a data-generating process defined via a branching process.
-The motivation for this purely methodological study comes from the modeling of biological sequence data, which is one of the major fields of application of inverse statistical mechanics \cite{levy_potts_2017,cocco_inverse_2018}, but even more of other bioinformatic inference approaches, cf.~\cite{durbin1998biological,felsenstein2004inferring}.
-Sequence data are the result of natural evolution, {\em i.e.}~a branching process initiating in some unknown common ancestor of extant and thus observable sequences \cite{sayers_genbank_2019,uniprotconsortium_uniprot_2018,el2019pfam}.
-The prototypical datasets in this case are multiple-sequence alignments (MSA), with lines being a so-called homologous, {\em i.e.} evolutionarily related sequences, and columns specific positions deriving from some common ancestral position.
-Such MSA contain at least two kinds of information:
+In more rare cases where data correspond to observed time series, theoretical and algorithmic development is much less advanced than for independent static data \cite{Inverse_problem_Berg}.
+
+One of the biggest application areas of inverse statistical mechanics is the modeling of biological processes.
+These applications are fuelled by the large amount of available data resulting from the impressive progress in experimental techniques in biology. 
+This is especially visible in the case of biological sequences,  with databases now harboring a vast amount of high-quality DNA or protein sequences \cite{sayers_genbank_2019,uniprotconsortium_uniprot_2018}.
+A common idea in this context is that it is possible to use characteristics of genes or organisms related by a common ancestry -- called \emph{homologous} -- to construct models of the selection acting on them.
+A successful example in this regard is the representation of protein sequences by probabilistic models in the so-called DCA method \cite{levy_potts_2017,cocco_inverse_2018}. 
+The prototypical datasets in this context are multiple-sequence alignments (MSA), with lines being a so-called homologous, {\em i.e.} evolutionarily related sequences, and columns specific positions deriving from some common ancestral position \cite{durbin1998biological}.
+The MSA contains at least two kinds complementary information:
 \begin{itemize}
     \item {\it Phylogenetic information:} the distances between sequences carry information about the evolutionary time since their common ancestor.
-Using phylogenetic methods \cite{felsenstein_phylogenies_1988,felsenstein2004inferring} we may reconstruct the evolutionary history of our dataset, represented by a phylogenetic tree.
+Using phylogeny inference methods \cite{felsenstein_phylogenies_1988,felsenstein2004inferring} we may reconstruct the evolutionary history of our dataset, represented by a phylogenetic tree.
     \item {\it Co-evolutionary information:} positions in a sequence typically do not evolve independently, but rather in a correlated way.
@@
 These two types of information are contained in two complementary features of the data: phylogenetic inference is based on the comparative analysis of different sequences, while co-evolutionary information is contained in the correlation of different columns of the MSA.
-Approaches aiming at one type of information typically neglect the other one: inference of phylogenies generally assumes that all positions in a sequence evolve independently, while co-evolutionary models of proteins assume that sequences in the MSA are independently distributed.
+Modeling approaches using one type of information typically neglect the other one: inference of phylogenies generally assumes that all positions in a sequence evolve independently, while co-evolutionary models of proteins assume that sequences in the MSA are independently distributed.
 This choice is motivated by the fact that taking the two types of correlations into account, \emph{i.e.} through time with phylogeny and accross trait values for co-evolution, results in very hard inference procedures, cf.~\cite{obermayer2014inverse,rodriguez2019toward}.
@@
 
-We have used phylogeny and co-evolution of biological sequences to explain our motivation, but the underlying problem is much more general.
+In this work, we consider the case of the inverse problem for high-dimensional data showing hierarchical correlations due to a branching generating process.
+Our motivation for this purely methodological study comes from the modeling of protein sequences discussed above, but the underlying problem is much more general. 
 Instead of sequences of discrete characters, like amino acids or nucleotides, we may consider continuous phenotypic traits. 
@@
 
-Here we use a simple and very general model for the temporal evolution of correlated variables: a historically well-known way to represent such processes is to use Ornstein-Uhlenbeck dynamics (OU), which models configurations as Gaussian vectors evolving in a quadratic potential that represents selection forces \cite{uhlenbeck_theory_1930, felsenstein_phylogenies_1988, hansen_stabilizing_1997}.
+To address this problem, we use a simple and very general model for the temporal evolution of correlated variables: a historically well-known way to represent such processes is to use Ornstein-Uhlenbeck dynamics (OU), which models configurations as Gaussian vectors evolving in a quadratic potential that represents selection forces \cite{uhlenbeck_theory_1930, felsenstein_phylogenies_1988, hansen_stabilizing_1997}.
 OU processes are commonly used in the field of phylogenetic comparative methods (PCM) \cite{bartoszek_phylogenetic_2012,mitov_fast_2020}. 
@@
 Taken individually, we assume that the $\vx_i$ are distributed according to Eq.~\eqref{eq:eq_distribution}, \ie~according to a multivariate Gaussian of zero mean and covariance $\bm C$.
-By construction, the equilibrium covariance between any pair of elements of a given vector $\vx=(x^1,...,x^L)$ is given by the inverse of the coupling matrix: $\langle x^a x^b\rangle-\langle x^a\rangle\langle x^b\rangle = \bm C_{ab} = (\bm J^{-1})_{ab}$ for all $a,b=1,...,L$.
+By construction, the equilibrium covariance between any pair of elements of a given vector $\vx=(x^1,...,x^L)$ is given by the inverse of the coupling matrix: $\langle x^a x^b\rangle-\langle x^a\rangle\langle x^b\rangle = \bm C_{ab} = (\bm J^{-1})_{ab}$ for all $a,b\in\{1,...,L\}$.
 This implies that inferring the coupling matrix defining the probability distribution amounts to finding the \emph{equilibrium} covariance matrix $\mathbf{C}$.
@@
 In other words, given the state $\vx_{a(n)}$ of the ancestral node, $\vx_n$ is sampled from $P(\vx_n |\, \vx_{a(n)}, \Delta t_{n,a(n)})$, see Eq.~\eqref{eq:OUpropagator}
-    \item As a consequence, OU processes on branches stemming from common ancestral node evolve independently, but from an identical initial condition.
+    \item As a consequence, OU processes on branches stemming from a common ancestral node evolve independently, but from an identical initial condition.
     \item Observed data vectors correspond to the states of the leaves of the tree at the end of this process. The states of the internal nodes are not part of the observed data and remain unknown. 
 \end{itemize}
-This process is thought to represent the evolution of biologic traits along a phylogenetic tree, with the leave nodes corresponding to traits observed in today's species.
+This process is thought to represent the evolution of biological traits along a phylogenetic tree, with the leaf nodes corresponding to traits observed in today's species.
 Note that due to the reversible nature of our OU process, the joint probability of any pair of leaf configurations $\vx_i$ and $\vx_j$, with $i,j\in\{1,...,N\}$, is given by  $P(\vx_i,\vx_j|\,\Delta t_{ij})$ (Eq.~\eqref{eq:OUjoint}), with $\Delta t_{ij}$ denoting the total branch length of the path connecting $i$ and $j$ in the tree.
@@
 Since $\mathbb{G}$ is the covariance matrix of a Gaussian distribution, we conclude the $\lambda_{ka}$ to be strictly positive.
+Interestingly, the definition of $\vec{S}_{ka}$ as a direct product between eigenvectors $\vsa$ of the energy potential and eigenvectors $\vuka$ reflecting the correlation structure mediated by the tree illustrates how these two types of information are entangled in the covariance matrix of the data.
 
@@
 With the $L$ eigenvalues $\rho_a$, this results in $L(L+1)/2$ independent values that fully parametrize the $L(L+1)/2$ values of $\iC$. 
-A second possibility, that we have found faster in practice, is to express the matrix of the $\{\vsa\}$ as the exponential of a skew-symetric matrix, see section \ref{sub:parametrizations_of_eigenvectors} of the SM.
+A second possibility, that we have found faster in practice, is to express the matrix of the $\{\vsa\}$ as the exponential of a skew-symetric matrix with $L(L-1)/2$ independent values, see section \ref{sub:parametrizations_of_eigenvectors} of the appendix.
 However, this parametrization does not allow a simple analytical expression of the gradient of the likelihood, and we use it along with automatic differentiation \cite{Zygote.jl-2018}. 
@@
 
- The optimization is performed by a quasi-Newton method \cite{NLopt}. Details are presented in section \ref{sub:optimization_scheme} of the SM.
+ The optimization is performed by a quasi-Newton method \cite{NLopt}. Details are presented in section \ref{sub:optimization_scheme} of the apprendix.
 
@@
 \subsubsection{Parametrization using generalized Eulerian angles}
-The idea is to write the base vectors $\vec{s}_a$ as columns of an orthogonal matrix $\bm S$, and to parameterize this matrix in terms of $L(L-1)/2$ independent variables  $\theta_ {pq}$ with $1 \leq p <q \leq L $.
+The idea is to write the base vectors $\vec{s}_a$ as columns of an orthogonal matrix $\bm{T}$, and to parameterize this matrix in terms of $L(L-1)/2$ independent variables  $\theta_ {pq}$ with $1 \leq p <q \leq L $.
 These parameters are called generalized Eulerian angles, since they generalize Eulerian angles to $L>3$.
@@
 \begin{equation} 
-\bm T_{pq} =  \left(
+\bm a_{pq} =  \left(
 \begin{array}{cccccc}
@@
 \right) \ ,
+\label{EP_eq:1}
 \end{equation}
@@
 
-An arbitrary $L$-dimensional orthogonal matrix $\bm S$ can be represented as a product of these $L(L-1)/2$ orthogonal matrices with appropriate values of the $L(L-1)/2$ independent parameters $\theta_{pq}$:
-\begin{equation}
-\label{eq:eulerprod}
-\bm S (\bm \theta)=\prod_{\left\lbrace pq\right\rbrace }T_{pq}(\theta_{pq}) \ .
-\end{equation}
-The basic idea is to express a general rotation in $L$ dimensions as the result of $L(L-1)/2$ successive rotations by angles $\theta_ {pq}$ in the two-dimensional subspaces given by the components $p, q$.
-For $L=3$ we get the classical Eulerian matrix.
-
-Ref.~{\color{red} [7] CHECK, is it [18]?
-use ref} exposes a recursive algorithm to efficiently perform the matrix multiplication of Eq.~\eqref{eq:eulerprod}, as well as the construction of the derivatives of $\bm S$ with respect to parameters $\theta_{pq}$.
-To start, Eq.~\eqref{eq:eulerprod} is transformed into a set of recursive equations, with $\bm S$ being equal to the final step
-\begin{equation}
-\bm S(\bm \theta)=\bm T^{(L)} \ .
-\end{equation}
-For all $1\leq k < l\leq L$ we iterate over $n$: {\color{red} [7] What is the initialization?
-Please check carefully, also the correct ranges of the $kl$ which seem to go to $n$ only.
-It would also be better to avoid $s$, which we use for the eigenvectors.} \pierre{Adding on Martin's comment above: I don't understand the equations below.
-Matrices $T_{pq}$ defined above are not used at all (I'm guessing they're used in the initialization but it's not said).
-It's not said that $1\leq n \leq L$ (is it the case?).
-I don't really understand what $s^{(n)}_{kl}$ is (Eq.
-(A9) gives some information but it's quite confusing) and as Martin points it has the same name as the eigenvectors we want to compute.
-We also use several different set of indices to go from $1$ to $L$: $(p,q)$, $n$, $(k,l)$ and $(i,j)$.
-In the maintext we use $(a,b)$ for this, making this confusing.
-Finally, I think the $*$ symbol should not be here if it just stands for normal multiplication, or maybe replaced by $\cdot$ if needed.
-Finally, this does not say how we compute the derivative of $\bm S$ w.r.
-to $\{\theta_{pq}\}$, which is the most important part.
-Much more explanations needed (at least for me...)!
-}
-\begin{equation}
-    T^{(n)}_{kl}=\cos\theta_{kn} * t^{(n)}_{kl}-\sin\theta_{kn}* s^{(n)}_{kl}
-\end{equation}
-with
-\begin{equation}
-    \begin{split}
-        s^{(n)}_{k+1,l}&=\sin\theta_{kn} * t^{(n)}_{kl}+\cos\theta_{kn}*s^{(n)}_{kl}\\
-        s^{(n)}_{k+1,n}&=\cos\theta_{kn}*s^{(n)}_{kn}\\
-        s^{(n)}_{1,l}&=-\delta_{ln}\\
-        \theta_{nn}&=\pi/2
-    \end{split}
-\end{equation}
-and
+An arbitrary $L$-dimensional orthogonal matrix $\bm T$ can be represented as a product of these $L(L-1)/2$ orthogonal matrices with appropriate values of the $L(L-1)/2$ independent parameters $\theta_{pq}$.
+Ref.
+\cite{Raffenetti1970GEA} exposes a recursive algorithm to efficiently perform the matrix multiplication, as well as the construction of the derivatives of $\bm T$ with respect to parameters $\theta_{pq}$.
+The main equations are presented below.
+
+The matrix multiplication  can be done by a sequence of $L$ steps  implied by the following recurrence relations where $n$ goes from 1 to $L$:
+\begin{equation}
+	\bm T=\bm T^{(L)},
+	\label{EP_eq:2}
+\end{equation}
+
+\begin{equation}
+\bm T^{(n)}= \bm A^{(n)} \bm t^{(n)},
+\label{EP_eq:3}
+\end{equation}
+
 \begin{equation} 
@@
 \bm T^{(n-1)} & \bm 0  \\
-\bm 0 &  \bm I
+\bm 0 &  \bm I\\
 \end{array}
-\right)
-\end{equation}
-
-As a consequence, the eigenvectors $\vec{s}_a$ of the matrix $\bm C$ with dimensions $L\times L$ can be written as:
-\begin{equation}
-s^{k}_a=T^{(L)}_{.,a}=\cos\theta_{kL} * t^{(L)}_{ka}-\sin\theta_{kL}* {\color{red} s}^{(L)}_{ka} \;\;\; \text{for} \;k=1,..,L \ .
-\end{equation}
-To use this expression, we still need to determine parameters $\bm \theta$  given an orthogonal matrix $\bm S$, such that all equations
-\begin{equation}
-    S_{ij}(\bm \theta)=S_{ij}
+\right),
+\label{EP_eq:4}
+\end{equation}
+
+\begin{equation} 
+\bm T^{(1)}=1,
+\label{EP_eq:init}
+\end{equation}
+
+\begin{equation}
+A^{(n)}= \bm a_{n,n}\,\bm a_{n-1,n}\dotsc\bm a_{2,n} \,\bm a_{1,n} ,
+\label{EP_eq:5}
+\end{equation}
+
+where the $\bm a_{n,m}$ matrices are defined by \eqref{EP_eq:1} for $n\neq m$, and $\bm a_{n,n}$ is the identity matrix of dimension $n$.
+
+The recurrence equations given by \eqref{EP_eq:3},  can be explicitly written as
+
+\begin{equation}
+T^{(n)}_{kl}=\cos\theta_{kn} \cdot t^{(n)}_{kl}-\sin\theta_{kn}\cdot z^{(n)}_{kl}\;\;\; \text{with} \;\;\; k,l=1,...,n
+\label{EP_eq:6}
+\end{equation}
+
+where
+
+\begin{equation}
+z^{(n)}_{kl} =
+\begin{cases}
+\delta_{ln} & \text{for $k=1$}\\
+\sin\theta_{k-1 n}\cdot t^{(n)}_{k-1 l}+\cos\theta_{k-1 n}\cdot z^{(n)}_{k-1 l} & \text{for $k=2,...,n$}  
+\end{cases}
+\label{EP_eq:7}
+\end{equation}
+with $\theta_{nn}=\pi/2$.
+
+Thus, if $\bm T^{(n-1)}$ is given, we find $\bm t^{(n)}$ from equation \eqref{EP_eq:4}.
+Then from elements $t_{kl}^{(n)}$ we get $z_{kl}^{(n)}$ using \eqref{EP_eq:7} and finally from $z_{kl}^{(n)}$ and $t_{kl}^{(n)}$ we obtain $T^{(n)}_{kl}$.
+
+Therefore eigenvectors $\vec{s}_a$  can be chosen as $a$th column  of matrix $\bm{T}$:
+
+\begin{equation}
+s^{k}_a=T^{(L)}_{.,a}=\cos\theta_{kL} \cdot t^{(L)}_{ka}-\sin\theta_{kL}\cdot z^{(L)}_{ka} \;\;\; \text{for} \;k=1,..,L
+ \label{EP_eq:8}
+\end{equation}
+
+\textbf{Determination of parameters for a given matrix}
+
+To use this expression, we still need to determine parameters $\bm \theta$  given an orthogonal matrix $\bm T$, such that all equations
+\begin{equation}
+    T_{ij}(\bm \theta)=T_{ij}
 \end{equation}
@@
 \begin{equation}
- {\bm{\hat \theta}} = \text{argmin}_{\bm \theta} \sum_{i<j}\left[ S_{ij}-S_{ij}(\bm \theta)\right] ^2 \ .
+ {\bm{\hat \theta}} = \text{argmin}_{\bm \theta} \sum_{i<j}\left[ T_{ij}-T_{ij}(\bm \theta)\right] ^2 \ .
 \end{equation}
@@
 
+\textbf{Derivatives with respect to the angular parameters}
+
+To compute the derivatives of $\bm T$ with respect to the angular parameters $\theta_{pq}$ we first note that it is possible to rewrite the recurrence step of equation \eqref{EP_eq:3} as the following matrix product
+
+ \begin{equation}
+ \bm T^{(L)}= \bm B^{(L)} \bm B^{(L-1)}\cdots\bm B^{(3)}\bm B^{(2)}   
+ \label{EP_eq:10}
+ \end{equation}
+
+where
+
+\begin{equation} 
+\bm B^{(n)} =  \left(
+\begin{array}{cc}
+\bm A^{(n)} & \bm 0  \\
+\bm 0 &  \bm I ^{(L-n)}\\
+\end{array}
+\right)
+\label{EP_eq:11}
+\end{equation}
+
+is block diagonal, $\bm A^{(n)} $ was defined by Eq.~\eqref{EP_eq:5} and $\bm I ^{(L-n)}$ is unit matrix of $(L-n)$ dimensions.
+
+From the definition of $\bm A^{(n)} $ we note that the terms $\theta_{pq}$ for $p=1,2,...,q-1$ only occur in the factor $\bm B^{(q)}$ of equation \eqref{EP_eq:10}. 
+This allows to write  the derivative of $\bm T$ with respect to $\theta_{pq}$  as the matrix product:
+
+\begin{equation}
+ \frac{\partial \bm T}{\partial \theta_{pq}}= \bm B^{(L)} \bm B^{(L-1)}\cdots \frac{\partial \bm B^{(q)}}{\partial \theta_{pq} }\cdots\bm B^{(3)}\bm B^{(2)}   
+\label{EP_eq:12}
+\end{equation}
+
+where
+
+\begin{equation} 
+ \frac{\partial \bm B^{(q)}}{\partial \theta_{pq} } =  \left(
+\begin{array}{cc}
+ \frac{\partial\bm A^{(q)}}{\partial \theta_{pq}  } & \bm 0  \\
+ \bm 0 &  \bm 0\\
+\end{array}
+\right)
+\label{EP_eq:13}
+\end{equation}
+
+Therefore  the calculation of the derivative of $\bm T$ could be done with the following three steps:
+\begin{enumerate}
+\item Calculate the product $\bm B^{(q-1)}\bm B^{(q-2)}\cdots\bm B^{(3)}\bm B^{(2)}   $
+\item  Calculate \begin{equation}
+\frac{\partial \bm B^{(q)}}{\partial \theta_{pq} }\bm B^{(q-1)}\cdots\bm B^{(3)}\bm B^{(2)}   = \left(
+\begin{array}{cc}
+\frac{\partial\bm T^{(q)}}{\partial \theta_{pq}  } & \bm 0  \\
+\bm 0 &  \bm 0\\
+\end{array}
+\right) 
+\label{EP_eq:14}
+\end{equation}
+
+\item Calculate \begin{equation}
+\frac{\partial\bm T}{\partial \theta_{pq}  }=\bm B^{(L)}\bm B^{(L-1)}\cdots \bm B^{(q+1)} \left(
+\begin{array}{cc}
+\frac{\partial\bm T^{(q)}}{\partial \theta_{pq}  } & \bm 0  \\
+\bm 0 &  \bm 0\\
+\end{array}
+\right) 
+\label{EP_eq:15}
+\end{equation}
+
+\end{enumerate}
+
+The $q-3$ recurrence steps for step 1  can be carried out using the same recurrence scheme described before for matrix $\bm T$ construction. For step 2, we need to evaluate 
+\begin{equation}
+\frac{\partial\bm T^{(q)}}{\partial \theta_{pq} }  =
+\begin{cases}
+- \sin\theta_{pq} \sigma^{(q)}_{kl} & k>p\\
+- \sin\theta_{pq} t^{(q)}_{kl} - \cos\theta_{pq} z^{(q)}_{kl} & k=p \\
+0 & k<p
+\end{cases}
+\label{EP_eq:17}
+\end{equation}
+
+where the quantities
+ \begin{equation}
+\sigma^{(q)}_{kl}=\frac{\partial\bm z^{(q)}_{kl}}{\partial \theta_{pq} }  
+\label{EP_eq:18}
+\end{equation}
+
+can be obtained from \eqref{EP_eq:7}. 
+Finally for step 3 we follow $L-q$ recurrence steps described by equations \eqref{EP_eq:6} and \eqref{EP_eq:7} with the two exceptions:
+
+ \begin{equation}
+z^{(n)}_{1l}=0\;\;\;\;\;1\leq l < n
+\label{EP_eq:19}
+\end{equation}
+ \begin{equation}
+z^{(n)}_{k+1,n}=0  \;\;\;\;\;1\leq k \leq n-1
+\label{EP_eq:20}
+\end{equation}
+
 \subsubsection{Parametrization in term of the exponential of a skew-symmetric matrix }
-
-Any orthogonal matrix can be written as the exponential of a skew-symmetric matrix $\bm X=-\bm X^T$: 
+The exponential of a skew-symmetric matrix   $\bm X=-\bm X^T$ is  a  special orthogonal  matrix  : 
+%Any orthogonal matrix can be written as the exponential of a skew-symmetric matrix $\bm X=-\bm X^T$: 
 \begin{equation}
 	\label{eq:sksym_def}
-    \bm S=\exp (\bm X)\ ,
-\end{equation}
-since $\exp(\bm X)^T=\exp(\bm X^T)=\exp(-\bm X)=\exp(\bm X)^{-1}$. \pierre{This does not prove the statement above but the other implication: the exponential of a skew-symmetric is orthogonal.}
-To obtain a skew-symmetric matrix $\bm X$ from an orthogonal matrix $\bm S$, it is sufficient to invert the exponential relation, $\bm X=\log \bm S$.
-\pierre{What needs to be proven is that $\log \bm S$ is always defined if $\bm S$ is orthogonal, and that $\bm$ X is skew-symetric.}
+    \bm S=\exp (\bm X).
+\end{equation}
+This is simply shown by the fact that $\exp(\bm X)^T=\exp(\bm X^T)=\exp(-\bm X)=\exp(\bm X)^{-1}$ and $\det (\bm S)=\exp{(\Tr \bm X)}=1$ since $\Tr \bm X=0$ for a skew-symmetric matrix.
+Furthermore, it is always possible to obtain a skew-symmetric matrix $\bm X$ from a special orthogonal matrix $\bm S$ by inverting the exponential relation, $\bm X=\log \bm S$ \cite{Shepard_param_OM}.
 
@@
 It is not possible to give a simple analytical form to Eq.~\eqref{eq:derivative_sksym}. 
-However, since $\bm S$ is obtained through a simple algebraic expression (Eq.~\eqref{eq:sksym_def}), it is possible to compute its derivative with respect to entries of $\bm X$ through automatic differentiation techniques {\color{red} CITE}.
-\pierre{About what is below: I don't think it's necessary to explain how AD works. It's a technique at the base of all deep-learning methods, so we could just say that plus cite something.}
-The idea  is to represent a function as a computational graph, also called Wengert list {\color{red} CITE}, where each node in the list represents an intermediate result of the computation. 
-The intermediate results can be assembled using the chain rule to get the final derivative we are looking for.
-There are two main algorithms to traverse the chain rule in AD: a forward mode and a reverse mode.
-Here, we used the reverse-mode algorithm, which is also the one of choice for back-propagation in deep learning.
-In particular we implemented it using the Julia package Zygote.jl \cite{Zygote.jl-2018}.
+However, since $\bm S$ is obtained through a simple algebraic expression (Eq.~\eqref{eq:sksym_def}), it is possible to compute its derivative with respect to entries of $\bm X$ through automatic differentiation techniques \cite{griewank_automatic_1989}.
+We implemented this process using the Julia package Zygote.jl \cite{Zygote.jl-2018}.
 
@@
 Matrices such as the one in \eqref{eq:G_a} are called \emph{hyper-geometric}.
-For dimensions $2^K$, they have $K+1$ different eigenvalues given by: \pierre{ref?}
+For dimensions $2^K$, they have $K+1$ different eigenvalues given by:
 
@@
 The main feature in QNM when compared to standard Newton method is that the Hessian matrix $\bm H$ is approximated instead of computed exactly. 
-The Hessian approximation $\mathbf{\hat{H}}$ is chosen to satisfy the secant equation: \pierre{What's $\vec{x}_k$? I suppose it stands for parameters after the $k$th iteration? }
-
-$$\nabla \mathcal{L}(\vec{x}_k+\Delta\vec{x})=\nabla \mathcal{L}(\vec{x}_k)+\mathbf{\hat{H}} \Delta\vec{x} $$.
-
-for $n-dimensions$ $\mathbf{\hat{H}}$ is undetermined \pierre{what do you mean "for $n$ dimensions"?}, the various QNMs differ in their choice of the solution to the secant equation.
-We used LBFGS variant \pierre{maybe a reference}.
-This particular method is based on choosing $\mathbf{\hat{H}}$ as a positive definite matrix where
-
-$$\mathbf{\hat{H}}_{k+1}=\mathbf{\hat{H}}_{k}+\frac{y_k y^{T}_k}{y_k\Delta x_k} - \frac{\mathbf{\hat{H}}_k\Delta x_k(\mathbf{\hat{H}}_k\Delta x_k)^T}{\Delta x^T_k\mathbf{\hat{H}}_k\Delta x_k} \;\;\;\text{and}\;\;\; y_k=\nabla \mathcal{L}(\vec{x}_k+1)-\nabla \mathcal{L}(\vec{x}_k)$$
-\pierre{What's $H_k$? I'm guessing Hessian matrix at iteration $k$, but then all above equations should have been with $H_k$ as well instead of just $H$.}
+When maximizing the likelihood $\mathcal{L}$ with respect to parameters $\vec{\theta}$, the direction of the change of parameters $\Delta\vec{\theta}$ is determined by
+
+$$ \Delta\vec{\theta} = \mathbf{\hat{H}}_k \nabla \mathcal{L}(\vec{\theta}_k), $$
+
+where $\vec{\theta}_k$ and $\mathbf{\hat{H}}_k$ respectively represent the parameter values and the approximation of the Hessian at the $k$-th iteration. 
+% $$\nabla \mathcal{L}(\vec{\theta}_k+\Delta\vec{\theta})=\nabla \mathcal{L}(\vec{\theta}_k)+\mathbf{\hat{H}} \Delta\vec{\theta} $$,
+Various QNMs differ in their approximation of the Hessian matrix.
+We used the LBFGS variant which chooses $\mathbf{\hat{H}}_k$ as a positive definite matrix where
+
+$$\mathbf{\hat{H}}_{k+1}=\mathbf{\hat{H}}_{k}+\frac{y_k y^{T}_k}{y_k\Delta \theta_k} - \frac{\mathbf{\hat{H}}_k\Delta \theta_k(\mathbf{\hat{H}}_k\Delta \theta_k)^T}{\Delta \theta^T_k\mathbf{\hat{H}}_k\Delta \theta_k} \;\;\;\text{with}\;\;\; y_k=\nabla \mathcal{L}(\vec{x}_k+1)-\nabla \mathcal{L}(\vec{x}_k).$$
 
@@
 	\caption{Inferred $\gamma$ values as a function of real $\gamma$, for system size $L=4$.
-$\gamma_{emp}$ is the value obtained by the process described in section \ref{sub:initializing_parameters} of the SM.
+$\gamma_{emp}$ is the value obtained by the process described in section \ref{sub:initializing_parameters} of the appendix.
 $\gamma_{max}$ is the value inferred by the maximum-likelihood calculation.
@@
 
-\begin{figure}[!htb]
-	\centering
-	\includegraphics[keepaspectratio=true,width=0.5\textwidth]{Figures/Pearson_correlation_rate_full_range.pdf}
-	\caption{  Ratio between Pearson correlation of the  maximum-likelihood  and true covariance  matrix to the Pearson correlation of the  empirical and true covariances matrices for the two system sizes $L=4$ and $L=10$.}
-	\label{fig:pearson_corr_comp_L4}
-\end{figure}
+%\begin{figure}[!htb]
+%	\centering
+%	\includegraphics[keepaspectratio=true,width=0.5\textwidth]{Figures/Pearson_correlation_rate_full_range.pdf}
+%	\caption{  Ratio between Pearson correlation of the  maximum-likelihood  and true covariance  matrix to the Pearson correlation of the  empirical and true covariances matrices for the two system sizes $L=4$ and $L=10$.}
+%	\label{fig:pearson_corr_comp_L4}
+%\end{figure}
 
@@
 	\caption{Inferred $\gamma$ values as a function of real $\gamma$, for system size $L=10$.
-$\gamma_{emp}$ is the value obtained by the process described in section \ref{sub:initializing_parameters} of the SM.
+$\gamma_{emp}$ is the value obtained by the process described in section \ref{sub:initializing_parameters} of the appendix.
 $\gamma_{max}$ is the value inferred by the maximum-likelihood calculation.
@@
 	\includegraphics[keepaspectratio=true,width=1.0\textwidth]{Figures/PPVs_L10.pdf}
-	\caption{Quality of prediction of interactions for  different values of $\gamma$ and system size $L=4$.
+	\caption{Quality of prediction of interactions for  different values of $\gamma$ and system size $L=10$.
 Interactions are defined as non-zero elements of the coupling matrix.
-In the $L=4$ case, there are $6$ possible interactions.
+In the $L=10$ case, there are $45$ possible interactions.
 Predictions are made by taking the largest elements (in absolute terms) of the inferred coupling matrix.

### 2021-04-20 submit/main.tex
First submission

--- 2021-02-11 main.tex
@@
 \author{Edwin Rodr\'{i}guez Horta} 
-\affiliation{Sorbonne Universit\'{e}, CNRS, Institut de Biologie
+\email{Correspondence to: Edwin Rodr\'{i}guez Horta, \bf{edwin@fisica.uh.cu}}\affiliation{Sorbonne Universit\'{e}, CNRS, Institut de Biologie
   Paris-Seine, Laboratoire de Biologie Computationnelle et
@@
 
+\textbf{Code availability:} The code used to compute and optimize the likelihood is available at https://github.com/ed-rodh/GaussianPhylogeny,
+
 \textbf{Acknowledgments:} We acknowledge interesting discussions with Roberto Mulet.

### 2021-04-26 revision/response_to_referees.tex
answer comments of R3 on main text

%\documentclass[12pt]{article}
\documentclass[aps,rmp,onecolumn]{revtex4-1}
\usepackage{color}
\usepackage{enumitem}
\usepackage{amsmath,amsfonts,amssymb,amsthm}
\usepackage{bm}% bold math

\newcommand{\gene}[1]{{\it #1}}
\newcommand{\comment}[1]{{\color{red}#1}}
\definecolor{response}{rgb}{0.1, 0.1, 0.1}
\definecolor{drab}{rgb}{0.59, 0.44, 0.09}
\definecolor{celestialblue}{rgb}{0.29, 0.59, 0.82}
\definecolor{purple}{rgb}{0.459,0.109,0.538}
\definecolor{deepsaffron}{rgb}{1.0, 0.6, 0.2}
\newcommand{\pierre}[1]{{\color{red}Pierre: #1}}
\newcommand{\refa}[1]{\textbf{R1:} #1\vskip 5mm}
\newcommand{\refb}[1]{\textbf{R2:} #1\vskip 5mm}
\newcommand{\refc}[1]{\textbf{R3:} #1\vskip 5mm}

\newcommand{\editor}[1]{\textbf{Editor:} #1\vskip 5mm}
%\newcommand{\criticism}[1]{\textbf{Criticism:} #1}
\newcommand{\response}[1]{{\it {\color{response}\textbf{Response:} #1}}\vskip 5mm}
\newcommand{\responsedraft}[1]{{\it {\color{purple}\textbf{ResponseDraft:} #1}}\vskip 5mm}

\begin{document}
\section*{Response to reviewers}

\subsection*{Editor}

\editor{}

\response{}

\subsection*{Reviewer: 1}

\refa{}

\response{}

\subsection*{Reviewer: 2}

\refb{
}

\response{}

\subsection*{Reviewer: 3 - Main text}

\refc{I think the authors should add some comparison to other approaches of inference for OU processes (for example in the introduction), to allow the reader to understand how the work sits in the context of dynamical inference methods}
\responsedraft{Link with comments of Reviewer 1: we can probably group the two answers and modify the intro in a way that satisfies both.}

\refc{I think that ‘statement of the problem’ is an important part that probably deserves to be a section in itself instead of a subsection of ‘Methods’.}
\response{``Statement of the problem'' is now a section in itself.}

\refc{The extreme cases discussed at lines 193-195 seem the same discussed at lines 144-147. If this is the case, the authors could re-think whether it is worth to repeat this point or to establish explicitly the connection to what has been discussed before.}
\response{We agree with reviewer 3 that this was redundant, and modified the text accordingly. 
The last paragraph of section II (former lines 144-147) now introduces timescales of the OU process without referring to the inference procedure. 
Former lines 193-195 then detail what these timescales imply for the inference procedure (the two extreme cases), referring to section II.}

\refc{Is it necessary to introduce a new symbol, $z$, for the expression (13)? 
Note that $z$ is used also in the Appendix for something with a different meaning (e.g. line 637)}
\response{The symbol $z$ is only used in the couple of equations that follow its introduction, and we agree that it is not strictly necessary. 
However, we felt that it made the derivation clearer by separating the calculation of eigenvalues and eigenvectors of $\mathbb{G}$ from the specific functional form of eigenvalues of $\mathbf{\Lambda}$.}

\refc{I think it is not immediately clear what the ‘Results’ section will address. I think one could immediately state (like at line 316) that the goal is to assess the accuracy of Cmax compared to the real C and that such estimate will be compared to the alternativa estimate provided by the empirical C. The definition of empirical C could be also explicitly written down. Later on, at line 374, the authors mention the ‘empirical coupling matrix’ but it is not clear how it is derived, is this simply then the inverse of the empirical C? Clarifying this would help better understand why the fact that C is close to singular (line 349) has an impact especially on the error on J (Fig. 3 right).}
\response{We agree with the reviewer that the Results section lacked explanation. 
We added an introductory paragraph that states the goal (evaluating the accuracy of the inference procedure) and describes how we proceed.
The empirical covariance matrix as well as the different coupling matrices are also clearly defined in this new paragraph. }

\refc{I thought also that it is not clear how the predictions for an ‘i.i.d sample’ (lines 355, 376) are obtained: do they serve as a random control giving the random expectation in for the errors in e.g. Fig. 3? I think this should be stated more clearly, also in the caption, maybe along with the numerical value itself of the i.i.d. error (that is difficult to read from the plots’s scale).}
\responsedraft{{\bf{Edwin}: any way that you can get and add to the text the exact numerical values for the error of the i.i.d. of sizes 16 / 32 / 64? Those are the examples we give in the text.}}
\response{The predictions of the \emph{i.i.d.} sample can indeed be used as a control to compute the minimal error that we can expect for a sample size $N$. 
Here, we use them to define an effective sample size: to each value of the  reconstruction error $\vert\vert \bm{C}_{max} - \bm{C}\vert\vert_2$ corresponds an \emph{i.i.d.} sample of size $N$ for which the same $l2$-error is obtained when using the empirical estimation of the covariance matrix.
This allows us to formulate the gains obtained by our method in terms of an equivalent increase in \emph{i.i.d.} sample size.

We have reformulated the corresponding paragraph to clarify this idea, and added more explanations to the caption of figure S5.}

\refc{Line 336: It is written that the sampling is repeated 100 times, it is not clear if also the inference is repeated 100 times or if this is done to increase $N$, the sample size. In relation to this, it would be important to write somewhere what $N$ is considered in the computational tests and hence what is the computational complexity given this $N$ (and $L=4,10$).}
\responsedraft{What does he mean by computational complexity given values of $N$ and $L$? 
I've only ever seen computational complexity expressed as an asymptotic quantity.}
\response{The inference is indeed repeated a hundred times. 
We do this to average our results accross different realizations of the OU process. 
We now clearly state $N$ corresponds to the number of leaves of the tree, in our case $2^9=512$.}

\refc{Line 370: Please add some references for the problem of contacts in proteins.}
\refc{Line 421: Please add some references for the evidence that protein landscapes are well described by pairwise models.}
\responsedraft{TODO}

\refc{Regarding the results, do the authors believe that inferring $\gamma$ could be interesting in terms of biological applications? Could it inform about typical evolutionary timescales?}
\response{The inference of $\gamma$ indeed provides information about typicaly times of the underlying evolutionary process, provided that the eigenvalues of the covariance matrix $\bm{C}$ are also known (Eq.~(7) and comments below it). 
If the evolution of, say, protein sequences is reasonably well described by the OU process, we could then be able to make statements about the expected change in an amino acid sequence as a function of time or number of generations.

A current limitation of that approach is that $\gamma$ will be inferred using distances separating leaves on a tree. 
In typical tree-inference software from biological sequences, the branch length is related to the expected number of mutations on that branch, but cannot be directly related to physical time or number of generations. 
}
\responsedraft{Not sure if this is a good answer. Branch lengths on trees can be related to real time using a molecular clock hypothesis, but it's unclear to me if that's consistent with the OU process itself?}

\refc{Fig. 3: I do not understand what the legend symbol for C_iid refers to (left panel). Is this simply the dotted horizontal line at the bottom of the plot? If yes I do not see why the legend has an additional central dot. In the caption, I do not understand ‘The inset in both panels’, isn’t the inset only in the right panel? Same comment applied to caption of Fig. S9.}
\responsedraft{\bm{Edwin}: Can you change the legend to reflect what's on the figure? Legend for Ciid should be a dashed line only. }
\response{The reviewer is right: the legend for figures 3 and 9 is incorrect -- it should be a dashed line only for $\bm{C}_{iid}$ and $\bm{J}_{iid}$ -- and the inset figure is only present in the right panel. 
We have corrected these errors. }

\refc{
	\pierre{Below: issues that I have addressed by changing the MS, but that perhaps do not require one response each.}
	\begin{itemize}
		\item Line 191: I would state explicitly why the inference becomes impossible
		\item Line 240: should ``we construct the direct product of vectors $s_{a}$ and $u_{ka}$, defining vectors $S_{ka}$ of dimensions $L \times N$’ be ‘we construct the direct product of vectors $u_{ka}$ and $s_{a}$, defining vectors $S_{ka}$ of dimensions $N \times L$''?
		I would also state explicitly that the superindex refers to the components of $u$.
		\item Line 278: double brackets around 17.
		\item Is there a minus in front of r.h.s of (19)?
		\item Line 332: I would add a comment on the choice of the range for gamma (why not even smaller or bigger) - this is probably due to the fact that all the variation concentrates in the range considered but I would state it.
		\item Line 346: As far as I see the l2-error is not defined anywhere - I think it would be important to have it defined somewhere (could be also in a caption or in the appendix).
		\item Fig 4: I think the caption should state clearly what the ‘Perfect’ case refers to. Same comment applies to caption of Fig. S10.
		\item I think the author of Ref. [15] was repeated twice by mistake.
	\end{itemize}
}

\end{document}

### 2021-04-26 main.tex
answer comments of R3 on main text

@@
     \item {\it Phylogenetic information:} the distances between sequences carry information about the evolutionary time since their common ancestor.
-Using phylogeny inference methods \cite{felsenstein_phylogenies_1988,felsenstein2004inferring} we may reconstruct the evolutionary history of our dataset, represented by a phylogenetic tree.
+Using phylogeny inference methods \cite{felsenstein_phylogenies_1988,felsenstein_joseph_inferring_2003} we may reconstruct the evolutionary history of our dataset, represented by a phylogenetic tree.
     \item {\it Co-evolutionary information:} positions in a sequence typically do not evolve independently, but rather in a correlated way.
@@
 The manuscript is divided as follows: we first review in section \ref{sec:ornstein_uhlenbeck_dynamics} the main characteristics of the multivariate OU process. 
-We then describe the setting of the inference problem that we want to solve in section \ref{sub:statement_of_the_problem}, propose a solution in sections \ref{sub:Calculation_of_the_likelihood} and \ref{sub:maximization_of_the_likelihood}.
+We then describe the setting of the inference problem that we want to solve in section \ref{sec:statement_of_the_problem} and propose a solution in section \ref{sec:methods}.
 Finally, we present results obtained on simulated data in section \ref{sec:Results}, with the context of pairwise models of protein sequences in mind.
@@
 
-Eq.~\eqref{eq:pairwisecov} allows us to readily distinguish two regimes. 
+Eq.~\eqref{eq:pairwisecov} allows us to identify typical timescales associated to the OU process. 
 Let us call $\rho_a$ the eigenvalues of $\bm{J}$. 
 The eigenvalues of $\Lam^{\Delta t}\bm{C}$ are then equal to $\rho_a^{-1}e^{-\gamma\rho_a\Delta t}$. 
-Since all $\rho_a$ are positive, the eigenvalues of $\Lam^{\Delta t}\bm{C}$ vanish exponentially over time.
+Since all $\rho_a$ are positive, the eigenvalues of $\Lam^{\Delta t}\bm{C}$ vanish exponentially over time over timescales of the form $\rho_a\Delta t$. 
 The slowest timescale of exponential decay is set by $\tau_c^{-1} = \gamma\rho_{min}$, with $\rho_{min}$ being the smallest eigenvalue of $\bm{J}$. 
-Thus, for $\Delta t / \tau_c \gg 1$, $\vx_1$ and $\vx_2$ are uncorrelated. 
-If this is verified for all pairs of observations $\vx_i$ and $\vx_j$, the regime is that of \emph{uncorrelated} data -- the inference of $\bm{J}$ can simply be performed by inverting the empirical covariance matrix extracted from the data.
-Inversely, for $\Delta t / \tau_c \ll 1$, $\vx_1$ and $\vx_2$ are highly correlated, defining a \emph{strongly correlated} regime. 
+Thus, for $\Delta t / \tau_c \gg 1$, $\vx_1$ and $\vx_2$ are uncorrelated, whereas they are highly correlated for $\Delta t / \tau_c \ll 1$, $\vx_1$ and $\vx_2$.
 It should be noted that for $\Delta t = 0$, the joint correlation matrix of $\vx_1$ and $\vx_2$ becomes non invertible, and Eq.~\eqref{eq:pairwisecov} becomes irrelevant.
@@
 
-\section{Methods}
-\label{sec:methods}
-
-\subsection{Statement of the problem}
-\label{sub:statement_of_the_problem}
+\section{Statement of the problem}
+\label{sec:statement_of_the_problem}
 
@@
 This problem shows two notable extreme cases: 
-The first one is the case where the typical branch length of the tree is short compared to the timescales of the OU process.
-As a consequence, leaf configurations are close to identical to the root, \ie~$\vx_i\simeq\vx_r$, and the inference of $\bm C$ becomes impossible. 
+The first one is the case where the typical branch length of the tree is short compared to the timescales introduced in section \ref{sec:ornstein_uhlenbeck_dynamics}
+As a consequence, leaf configurations are close to identical to the root, \ie~$\vx_i\simeq\vx_r$, and the inference of $\bm C$ becomes impossible due to lack of independent samples from $P_{eq}$. 
 The second one is the opposite case where the typical branch length of the tree is long compared to the longest timescale of the OU process $\tau_c$.
@@
 We detail the computation of this probability in the following section.
+
+\section{Methods}
+\label{sec:methods}
 
@@
 
-To obtain the eigenmodes of the joint covariance matrix $\mathbb{G}$ as a function of the $\vsa$ and $\vuka$, we construct the direct product of vectors $\vsa$ and $\vuka$, defining vectors $\vec{S}_{ka}$ of dimension $L\times N$:
+To obtain the eigenmodes of the joint covariance matrix $\mathbb{G}$ as a function of the $\vsa$ and $\vuka$, we construct the direct product of vectors $\vuka$ and $\vsa$, defining vectors $\vec{S}_{ka}$ of dimension $L\cdot N$:
 \begin{equation}
@@
 		\vec{S}_{ka} &= \vuka \otimes \vsa \\
-		&= [u_{ka}^1\cdot\vsa, \ldots, u_{ka}^N\cdot\vsa].
-		% &= [u_{ka}(1)s_a(1), \ldots, u_{ka}(1)s_a(L); \ldots; u_{ka}(N)s_a(1), \ldots, u_{ka}(N)s_a(L)]
+		&= [u_{ka}^1\cdot\vsa, \ldots, u_{ka}^N\cdot\vsa],
 	\end{split}
@@
 \end{equation}
+where the superindex $i\in\{1\ldots N\}$ refers to components of $\vuka$.
 The $i$th block vector of $\vec{S}_{ka}$ will thus be written as $\vec{S}_{ka}^i = u_{ka}^i\cdot\vsa$. 
@@
 We also need to find the optimal time scale $\gamma$. 
-In Eq.~(\eqref{eq:likelihood_3}), the likelihood is expressed as a function of $\gamma$ and $\{\rho_a, \vsa\}$, \ie~the eigenvalues and eigenvectors of $\iC$, either directly or through the quantities $\{\lambda_{ka}, \vuka \}$.
+In Eq.~\eqref{eq:likelihood_3}, the likelihood is expressed as a function of $\gamma$ and $\{\rho_a, \vsa\}$, \ie~the eigenvalues and eigenvectors of $\iC$, either directly or through the quantities $\{\lambda_{ka}, \vuka \}$.
 We know attempt to maximize the likelihood with respect to the eigenmodes $\{\rho_a, \vsa\}$ and to the time scale $\gamma$.
@@
 \begin{equation}
-	\frac{\partial \mathcal{L}}{\partial\theta_{\alpha\beta}} = \sum_{k=1}^N \lambda_k^{-1} \left(\sum_{i=1}^N u_k^i\vx_i\cdot\vsa\right) \left(\sum_{i=1}^N u_k^i \vx_i\cdot\frac{\partial \vsa}{\partial\theta_{\alpha\beta}}\right),
+	\frac{\partial \mathcal{L}}{\partial\theta_{\alpha\beta}} = -\sum_{k=1}^N \lambda_k^{-1} \left(\sum_{i=1}^N u_k^i\vx_i\cdot\vsa\right) \left(\sum_{i=1}^N u_k^i \vx_i\cdot\frac{\partial \vsa}{\partial\theta_{\alpha\beta}}\right),
 	\label{eq:gradlikelihood_2}
@@
 
-In order to evaluate our inference procedure, we generate artificial data corresponding to the process described in section \ref{sub:statement_of_the_problem}. 
-We first build a  balanced binary tree $\mathcal{T}$ with $2^9=512$ leaves. 
+In order to evaluate the accuracy of our inference procedure, we want to estimate how close the covariance matrix reconstructed by likelihood maximization $\bm{C}_{max}$ is to the real covariance matrix $\bm{C}$.
+To establish  a reference point for inference quality, we have to compare $\bm{C}$ to the empirical covariance matrix $\bm{C}_{emp}$ computed as if observations were independent, \emph{i.e.} $\bm{C}_{emp}(a,b) = N^{-1}\sum_{i=1}^N x_i^a x_i^b$.
+Note that it is also possible to estimate the accuracy of the method by comparing the coupling matrices obtained by inverting each covariance matrix, namely $\bm{J}$, $\bm{J}_{max}$ and $\bm{J}_{emp}$.
+Here, we described results obtained on artificial data.
+
+Data is generated according to the process described in section \ref{sec:statement_of_the_problem}. 
+We first build a  balanced binary tree $\mathcal{T}$ with $N=2^9=512$ leaves. 
 The length of each branch of $\mathcal{T}$ is chosen from a uniform distribution in the interval $[0,1]$. 
@@
 For $\gamma \gg \gamma_d$, leaf configurations are on average well decorrelated, whereas for $\gamma \ll \gamma_d$ all leaves will be strongly correlated. 
-By simulating data using different $\gamma$ in the range $[10^{-2},2]\cdot\gamma_d$, we investigate all relevant temporal regimes.  
+By simulating data using a range of values of $\gamma$, we can investigate these different regimes. 
+Empirically, we find that the range $\gamma\in[10^{-2},2]\cdot\gamma_d$ covers all relevant temporal regimes.  
 For each value of $\gamma$, we then sample  configurations of leaves of $\mathcal{T}$ using the process described in section \ref{sub:generating_artificial_data} of the supplementary material. 
@@
 For each repetition of the sampling process, we perform our maximum likelihood procedure and obtain an inferred covariance matrix $\bm{C}_{max}$. 
-As a means of comparison, we also compute the empirical covariance matrix $\bm{C}_{emp}$ as if leaf configurations were independent. 
+We also compute the empirical covariance matrix $\bm{C}_{emp}$ as if leaf configurations were independent. 
 Fig.~\ref{fig:pears_L4} shows the Pearson correlation between the real covariance matrix $\bm{C}=\bm{J}^{-1}$ and the empirical or inferred ones in the $L=4$ case (similar figures for $L=10$ are in Appendix \ref{sub:supplementary_figures}).
@@
 However, there exists an intermediate regime where $\bm{C}_{max}$ is much closer to the actual correlation than  $\bm{C}_{emp}$. 
-In Fig.~\ref{fig:error_1_L4}, we plot the relative $l2$-error between either covariance matrices in the left panel or coupling matrices in the right panel. 
+In Fig.~\ref{fig:error_1_L4}, we plot the relative $l2$-error between either covariance matrices in the left panel or coupling matrices in the right panel, where the $l2$-error is defined as the $l2$-norm of the difference between two matrices, \emph{e.g.} $\vert\vert\bm{C}_{max} - \bm{C}\vert\vert_2$.
 In both cases, our maximum-likelihood method results in a consistent improvement over the empirical estimator. 
@@
 Logarithmic scale is chosen for the $y$-axis because of large values of the error at low $\gamma$.
-The inset in both panels show the ratio between the two errors.}
+The inset shows the ratio between the two errors.}
 	\label{fig:error_1_L4}
@@
 Intuitively, the use of correlated samples reduces the information contained in the data, as compared to an equally large dataset of \emph{i.i.d.} configurations. 
-It is therefore interesting to compare the accuracy of our inferences with the accuracy obtained on smaller but \emph{i.i.d.} samples.
-To do so, we report in Fig.~\sref{fig:error_vs_nseqs} the $l2$-error between true and empirical covariances computed from a \emph{i.i.d.} samples of variable sizes $N$.
+It is therefore interesting to compare the accuracy of our inferences with the accuracy obtained by computing the empirical covariance matrix using \emph{i.i.d.} samples. 
+This can serve as a baseline for the minimal error that can be achieved for a given sample size $N$, but also to estimate the \emph{i.i.d.} sample size that the difference between $\bm{C}_{max}$ and $\bm{C}$ corresponds to.
+In other words, for each value of $\vert\vert \bm{C}_{max} - \bm{C}\vert\vert_2$ corresponds an \emph{i.i.d.} sample size $N$ for which the same $l2$-error is obtained.
+
+We report in Fig.~\sref{fig:error_vs_nseqs} the $l2$-error between true and empirical covariances computed from a \emph{i.i.d.} samples of variable sizes $N$. 
 As expected, the error increases with decreasing values of $N$. 
@@
 Predictions are made by taking the largest elements (in absolute terms) of the inferred coupling matrix.
-The PPV is the fraction of correctly predicted contacts for a given number of predictions.}
+The PPV is the fraction of correctly predicted contacts for a given number of predictions.
+The perfect prediction is to the one that would be made using the real coupling matrix $\bm{J}$.}
 	\label{fig:PPV_L4}
@@
 \clearpage
+\bibliographystyle{unsrt}
 \bibliography{bib_phylo}
-\bibliographystyle{unsrt}
 
@@
 	\caption{Relative $l2$-error between the empirical covariance matrix calculated from an \emph{i.i.d.} sample and the true covariance matrix, for system sizes $L=4$ and $L=10$.
+An \emph{i.i.d.} sample size $N$ can be attributed to each value of the $l2$-error.
+By comparing values of the $l2$-error with results in figures \ref{fig:error_1_L4} and \ref{fig:error_1_L10}, we can formulate the gain in accuracy in terms of a gain in effective sample size.
 The dashed vertical line corresponds to the number of leaves of the tree used in the simulations.}
@@
 Logarithmic scale is chosen for the y-axis because of large values of the error at low $\gamma$.
-The inset in both panels show the ratio between the two errors.
+The inset shows the ratio between the two errors.
 For system size $L=10$.}
@@
 Predictions are made by taking the largest elements (in absolute terms) of the inferred coupling matrix.
-The PPV is the fraction of correctly predicted contacts for a given number of predictions.}
+The PPV is the fraction of correctly predicted contacts for a given number of predictions.
+The perfect prediction is to the one that would be made using the real coupling matrix $\bm{J}$.}
 	\label{fig:PPV_L10}
